---
title: "Diploid Cell Type Annotation"
output: html_notebook
date: "2026-05-01"
description: >
  Assigns cell type labels to the integrated diploid human/chimp Seurat
  object (input from Kenneth's SeuratData pipeline) using CelliD, remaps
  fine-grained reference categories to the 20 broad cell types used
  throughout the DD/DE analysis, and produces per-cell-type Seurat/AnnData
  exports plus reference UMAP panels (also used to project the
  allotetraploid data into the same coordinate space).
---

## Overview

Assigns cell type labels to all cells in the integrated diploid human/chimp
Seurat object using **CelliD** (MCA embedding + hypergeometric enrichment test).

Reference gene signatures are drawn from three sources:
- **FCA** (Fly Cell Atlas / Human Cell Atlas via Wenhe): broad cell type markers
- **Linnarsson brain atlas**: neuronal sub-types
- **CM dataset (Gilad lab)**: cardiomyocyte / mesoderm lineages

Because CelliD is memory-intensive, the diploid object is split into 5 chunks
and annotated in parallel. Labels are then remapped from fine-grained reference
categories to the 20 broad cell types used in the DD/DE analysis.

**Outputs:**
- `diploid_cell_annotations.csv` — per-cell barcode + celltype label
- `diploid_metadata.csv` — full Seurat metadata joined with cell type labels
- Per-cell-type Seurat objects and AnnData exports for downstream analyses

---

## Load Packages

`CelliD` is a Bioconductor package (`BiocManager::install("CelliD")`), not
CRAN; everything else here is CRAN.

```{r load-packages, warning=FALSE, message=FALSE}
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(dplyr)
library(CelliD)
library(stringr)
```

---

## Step 1: Split Integrated Object into Chunks

The integrated diploid object is large (~hundreds of thousands of cells).
Splitting into 5 equal chunks allows CelliD to run per chunk without running
out of memory.

```{r split-into-chunks}
obj <- readRDS("/project/gilad/kenneth/scale_project/SeuratData/combined_1kFeatures_integrated.rds")

all_cells <- colnames(obj)
n_cells   <- length(all_cells)

# Assign cells to 5 groups deterministically then split
set.seed(1)
group_id    <- rep(1:5, length.out = n_cells)
cell_groups <- split(all_cells, group_id)

obj_list <- lapply(cell_groups, function(cells_use) subset(obj, cells = cells_use))
sapply(obj_list, ncol)

outdir <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

for (i in seq_along(obj_list)) {
  saveRDS(obj_list[[i]],
          file = paste0(outdir, "combined_1kFeatures_integrated_chunk", i, ".rds"))
}
```

---

## Step 2: Load Cell Type Reference Signatures

Three signature sets are combined into a single list for CelliD:
- **FCA** (Cao et al.): broad cell type markers, with Linnarsson brain subtypes
  substituted for the brain cell types
- **Linnarsson brain atlas**: more granular neuronal categories
- **CM / mesoderm signatures** (Gilad lab, Reem): cardiac lineage markers

Cell cycle genes and non-protein-coding genes are excluded from all signatures.

```{r load-reference-signatures}
# Brain and eye cell types present in FCA that we replace with other atlases
brain_fca <- c("Purkinje neurons", "Inhibitory neurons", "Astrocytes",
               "Granule neurons", "Oligodendrocytes", "Inhibitory interneurons",
               "Microglia", "Limbic system neurons", "Excitatory neurons",
               "Unipolar brush cells")
eye_fca   <- c("Ganglion cells", "Amacrine cells", "Retinal progenitors and Muller glia",
               "Photoreceptor cells", "Bipolar cells", "Horizontal cells",
               "Retinal pigment cells", "Lens fibre cells")

# FCA signatures (Cao et al., Human Cell Atlas)
fca_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/train_test/Cao_FCA_1Kcell_train_sce_250_signature_common_20iterate_cut500reads_noCTRiboMT.rds")

# Add iPSC markers manually
fca_signatures[["iPSCs"]] <- c("SOX2", "POU5F1", "NANOG", "NODAL", "ZFP42",
                                "DNMT3B", "UTF1", "LCK", "DPPA2", "IDO1",
                                "DPPA5", "LEFTY1", "ESRG", "CNMD", "SFRP2",
                                "THY1", "USP44", "VSNL1", "SPP1", "KLF4",
                                "GDF3", "L1TD1", "VRTN", "ZSCAN10")

# Linnarsson brain atlas — replaces FCA brain entries
linn_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/human_brain/brain_signature_reiterated_1000_Celltype_TopLevelCluster_noRiboCCMT.rds")

# Cardiac / mesoderm markers from Gilad lab
CM_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/CM_Reem/CM_signature_reiterated_1000_noRiboCCMT.rds")
CM_signatures <- CM_signatures[2:4]
names(CM_signatures) <- c("Early mesoderm cells", "Cardiac mesoderm cells", "Cardiac progenitor cells")

# Combine: FCA (minus brain) + Linnarsson brain + CM
fca_signatures <- c(
  fca_signatures[!names(fca_signatures) %in% brain_fca],
  linn_signatures,
  CM_signatures
)

# Remove cell cycle genes and keep only protein-coding genes
fca_signatures <- lapply(fca_signatures, function(genes) {
  genes[!genes %in% unlist(cc.genes.updated.2019) & genes %in% HgProteinCodingGenes]
})
```

---

## Step 3: Run CelliD Per Chunk

For each chunk:
1. Filter to protein-coding genes expressed in >5 cells
2. Compute highly variable genes (HVG), remove MT / ribosomal / cell cycle genes
3. Run MCA (Multiple Correspondence Analysis) embedding (50 dimensions)
4. Score each cell against all reference signatures via hypergeometric test
5. Assign the highest-scoring label (if -log10 p-value ≥ 2, i.e. p ≤ 0.01)

```{r run-cellid-per-chunk}
data_inputs <- paste0("combined_1kFeatures_integrated_chunk", 1:5, ".rds")
outdir      <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/"
all_cellid_orig <- c()

for (di in data_inputs) {
  obj <- readRDS(paste0(outdir, di))

  DefaultAssay(obj) <- "RNA"
  obj[["RNA"]] <- JoinLayers(obj[["RNA"]])

  # Keep protein-coding genes only
  counts_mat <- LayerData(obj, assay = "RNA", layer = "counts")
  genes_use  <- intersect(rownames(counts_mat), HgProteinCodingGenes)
  obj        <- subset(obj, features = genes_use)

  # Keep genes expressed in >5 cells
  counts_mat <- LayerData(obj, assay = "RNA", layer = "counts")
  gene_keep  <- rownames(counts_mat)[Matrix::rowSums(counts_mat > 0) > 5]
  obj        <- subset(obj, features = gene_keep)

  # HVG — exclude MT, ribosomal, and cell cycle genes
  hvg <- VariableFeatures(FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000))
  hvg <- hvg[!grepl("^RPL|^RPS|^MT-", hvg) & !hvg %in% unlist(cc.genes.updated.2019)]

  # Feature set = signature genes ∪ HVG (intersected with genes in object)
  obj_hvg <- c(unlist(fca_signatures), hvg) %>% unique() %>% intersect(rownames(obj))
  obj     <- ScaleData(obj, features = obj_hvg)

  print(paste("Running MCA on", di))
  obj <- RunMCA(obj, nmcs = 50, assay = "RNA", slot = "data", features = obj_hvg)

  # Hypergeometric enrichment test: returns -log10(corrected p-value) per cell x cell type
  cellid_enrichments <- RunCellHGT(obj, pathways = fca_signatures, dims = 1:50, n.features = 300)

  # Assign label with highest enrichment; mark as "unassigned" if max < 2 (-log10 p > 0.01)
  cellid_labels <- rownames(cellid_enrichments)[apply(cellid_enrichments, 2, which.max)]
  cellid_labels <- ifelse(apply(cellid_enrichments, 2, max) >= 2, cellid_labels, "unassigned")

  cellid_orig <- as_tibble(cellid_labels, rownames = "cell_barcode") %>%
    dplyr::select(cell_barcode, celltype = value)

  # Strip trailing cluster numbers appended by Linnarsson atlas (e.g. "Radial glia 3" → "Radial glia")
  cellid_orig$celltype <- sub("\\s*\\d+$", "", cellid_orig$celltype)

  all_cellid_orig <- rbind(all_cellid_orig, cellid_orig)
}
```

---

## Step 4: Remap Fine-Grained Labels to Broad Cell Types

Reference atlases contain more granular categories than we use. Remap to the
20 broad cell types used across all downstream DD/DE analyses.

```{r remap-celltype-labels}
all_cellid_orig_merged <- all_cellid_orig %>%
  mutate(celltype = replace(celltype, celltype == "Visceral neurons",             "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Chromaffin cells",             "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Sympathoblasts",               "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Schwann cells",                "ENS glia")) %>%
  mutate(celltype = replace(celltype, celltype == "Ureteric bud cells",           "Metanephric cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Bipolar cells",                "Photoreceptor cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Thymocytes",                   "Lymphoid cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Mesothelial cells",            "Epicardial fat cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Amacrine cells",               "Horizontal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Ganglion cells",               "Horizontal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Lymphatic endothelial cells",  "Vascular endothelial cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Mesangial cells",              "Stromal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Fibroblast",                   "Stromal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Parietal and chief cells",     "Goblet cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Cardiac progenitor cells",     "Cardiomyocytes")) %>%
  mutate(celltype = replace(celltype, celltype == "Horizontal cells",             "Retinal neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Photoreceptor cells",          "Retinal neurons"))

outpath <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
write.csv(all_cellid_orig_merged,
          paste0(outpath, "diploid_cell_annotations.csv"),
          row.names = FALSE)
```

---

## Step 5: Save Full Metadata

Join cell type labels onto full Seurat metadata and save.

```{r save-full-metadata}
seurat_data <- readRDS('/project/gilad/kenneth/scale_project/SeuratData/combined_1kFeatures_integrated.rds')
path        <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
cell_annotations_diploid <- read.csv(paste0(path, 'diploid_cell_annotations.csv'))

metadata <- seurat_data@meta.data %>%
  tibble::rownames_to_column("cell_barcode") %>%
  dplyr::left_join(cell_annotations_diploid, by = "cell_barcode")

write.csv(metadata, paste0(path, "diploid_metadata.csv"), row.names = FALSE)
```

---

## Step 6: QC — Cell Counts by Cell Type and Species

```{r load-annotations-for-qc}
path <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
cell_annotations_diploid <- read.csv(paste0(path, 'diploid_cell_annotations.csv'))
metadata                 <- read.csv(paste0(path, "diploid_metadata.csv"))
```

All cell types — total counts by species.

```{r plot-celltype-counts-by-species}
plot_df <- cell_annotations_diploid %>%
  mutate(
    species = case_when(
      str_starts(cell_barcode, "chimp") ~ "Chimpanzee",
      str_starts(cell_barcode, "human") ~ "Human",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(species)) %>%
  count(celltype, species, name = "cell_count") %>%
  group_by(celltype) %>%
  mutate(total_cells = sum(cell_count)) %>%
  ungroup() %>%
  mutate(
    celltype = reorder(celltype, -total_cells),
    species  = factor(species, levels = c("Human", "Chimpanzee"))
  )

ggplot(plot_df, aes(x = celltype, y = cell_count, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = cell_count), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Human" = "#D55E00", "Chimpanzee" = "#0072B2")) +
  labs(x = "", y = "Number of Cells", fill = "") +
  theme_classic() +
  theme(
    axis.text.x       = element_text(angle = 45, hjust = 1),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position   = c(0.9, 0.9),
    legend.background = element_blank(),
    legend.key        = element_blank()
  ) +
  expand_limits(y = max(plot_df$cell_count) * 1.1)
```

Cell types with ≥ 1,000 cells in both human and chimp (used in DD/DE analysis).

```{r plot-celltype-counts-filtered}
plot_df <- cell_annotations_diploid %>%
  dplyr::mutate(
    species = dplyr::case_when(
      stringr::str_starts(cell_barcode, "chimp") ~ "Chimpanzee",
      stringr::str_starts(cell_barcode, "human") ~ "Human",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(species), celltype != "unassigned") %>%
  dplyr::count(celltype, species, name = "cell_count") %>%
  dplyr::group_by(celltype) %>%
  dplyr::filter(all(cell_count >= 1000)) %>%
  dplyr::mutate(min_cells = min(cell_count)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    celltype = reorder(celltype, -min_cells),
    species  = factor(species, levels = c("Human", "Chimpanzee"))
  )

ggplot(plot_df, aes(x = celltype, y = cell_count, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = cell_count), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Human" = "#D55E00", "Chimpanzee" = "#0072B2")) +
  labs(x = "", y = "Number of Cells", fill = "") +
  theme_classic() +
  theme(
    axis.text.x       = element_text(angle = 45, hjust = 1),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position   = c(0.9, 0.9),
    legend.background = element_blank(),
    legend.key        = element_blank()
  ) +
  expand_limits(y = max(plot_df$cell_count) * 1.1)
```

Cell type proportions by individual donor (human and chimp separately).

```{r celltype-proportions-by-donor}
keep_celltypes <- metadata %>%
  filter(!is.na(species), !is.na(celltype), !is.na(donor_id),
         tolower(celltype) != "unassigned", species %in% c("human", "chimp")) %>%
  count(celltype, species) %>%
  tidyr::complete(celltype, species, fill = list(n = 0)) %>%
  group_by(celltype) %>%
  filter(all(n >= 1000)) %>%
  pull(celltype) %>%
  unique()

shared_celltype_order <- metadata %>%
  dplyr::filter(!is.na(species), !is.na(celltype), !is.na(donor_id),
                tolower(celltype) != "unassigned", species %in% c("human", "chimp"),
                celltype %in% keep_celltypes) %>%
  dplyr::count(celltype, name = "total_cells_all_species") %>%
  dplyr::arrange(dplyr::desc(total_cells_all_species)) %>%
  dplyr::pull(celltype)

#' Plot per-donor cell type composition for one species.
#'
#' Horizontal stacked bar chart showing, for a given species, what fraction
#' of each cell type's cells come from each donor -- used as a QC check that
#' no cell type is dominated by a single individual before it's used
#' downstream in DD/DE analysis.
#'
#' @param metadata_df Data frame (or tibble) of per-cell metadata with
#'   columns `species`, `celltype`, `donor_id`.
#' @param species_name Value of `species` to filter to (e.g. "human" or
#'   "chimp").
#' @param title Plot title.
#' @param celltype_order Character vector giving the y-axis cell type order
#'   (top to bottom); typically `shared_celltype_order` computed above.
#' @return A ggplot object.
make_species_prop_plot <- function(metadata_df, species_name, title, celltype_order) {
  plot_df <- metadata_df %>%
    dplyr::filter(!is.na(species), !is.na(celltype), !is.na(donor_id),
                  tolower(celltype) != "unassigned",
                  species == species_name, celltype %in% celltype_order) %>%
    dplyr::count(celltype, donor_id, name = "cell_count") %>%
    dplyr::group_by(celltype) %>%
    dplyr::mutate(total_cells = sum(cell_count), proportion = cell_count / sum(cell_count)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(celltype = factor(celltype, levels = rev(celltype_order)))

  ggplot(plot_df, aes(x = proportion, y = celltype, fill = donor_id)) +
    geom_col(width = 0.8) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
    labs(title = title, x = "Proportion of Cells", y = "", fill = "Individual") +
    theme_classic() +
    theme(
      panel.border  = element_rect(color = "black", fill = NA, linewidth = 0.8),
      axis.text.y   = element_text(size = 10),
      legend.position = "right"
    )
}

p_human <- make_species_prop_plot(metadata, "human", "Human",      shared_celltype_order)
p_chimp <- make_species_prop_plot(metadata, "chimp", "Chimpanzee", shared_celltype_order)

p_human
p_chimp

results_path <- '/project/gilad/awchen55/mechanism/results/cell_annotations/'
ggsave(paste0(results_path, "human_diploid_celltype_prop.png"), plot = p_human, width = 7, height = 5, dpi = 300)
ggsave(paste0(results_path, "chimp_diploid_celltype_prop.png"), plot = p_chimp, width = 7, height = 5, dpi = 300)
```

---

## Step 7: Prepare Seurat Objects with Annotations

Add cell type labels to the full Seurat object and save annotated versions.

```{r annotate-full-seurat-object}
seurat_data <- readRDS('/project/gilad/kenneth/scale_project/SeuratData/combined_1kFeatures_integrated.rds')

seurat_data@meta.data <- seurat_data@meta.data %>%
  tibble::rownames_to_column("cell_barcode") %>%
  dplyr::left_join(cell_annotations_diploid, by = "cell_barcode") %>%
  tibble::column_to_rownames("cell_barcode")

outdir <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/"
saveRDS(seurat_data, file = paste0(outdir, "combined_1kFeatures_integrated_annotated.rds"))
```

Add annotations to each chunk as well.

```{r annotate-chunks}
outdir <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/"
path   <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
cell_annotations_diploid <- read.csv(paste0(path, 'diploid_cell_annotations.csv'))

for (i in 1:5) {
  seurat_sub <- readRDS(paste0(outdir, "combined_1kFeatures_integrated_chunk", i, ".rds"))

  seurat_sub@meta.data <- seurat_sub@meta.data %>%
    tibble::rownames_to_column("cell_barcode") %>%
    dplyr::left_join(cell_annotations_diploid, by = "cell_barcode") %>%
    tibble::column_to_rownames("cell_barcode")

  saveRDS(seurat_sub, file = paste0(outdir, "combined_1kFeatures_integrated_annotated_chunk", i, ".rds"))
}
```

---

## Step 8: Split by Cell Type

Split the annotated object into one Seurat object per cell type. Only cell types
with ≥ 1,000 cells in both human and chimp are kept.

```{r example-split-by-celltype, eval=FALSE}
outdir     <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/"
seurat_data <- readRDS(paste0(outdir, "combined_1kFeatures_integrated_annotated.rds"))

seurat_list <- SplitObject(seurat_data, split.by = "celltype")

ct_outdir <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/celltype/"
for (ct in names(seurat_list)) {
  saveRDS(seurat_list[[ct]],
          file = file.path(ct_outdir, paste0("combined_1kFeatures_integrated_", ct, ".rds")))
}
```

Determine which cell types meet the ≥ 1,000 cell threshold in both species.

```{r example-determine-celltypes-to-keep, eval=FALSE}
celltypes_keep <- cell_annotations_diploid %>%
  mutate(
    species = case_when(
      str_starts(cell_barcode, "chimp") ~ "Chimpanzee",
      str_starts(cell_barcode, "human") ~ "Human",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(species), celltype != "unassigned") %>%
  count(celltype, species, name = "cell_count") %>%
  group_by(celltype) %>%
  filter(all(cell_count >= 1000)) %>%
  ungroup() %>%
  distinct(celltype) %>%
  pull(celltype)
```

---

## Step 9: Export to AnnData Format

Export each cell type's counts, metadata, and gene/cell lists as flat files
for conversion to AnnData in Python.

```{r export-to-anndata}
library(Matrix)

outdir      <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/celltype/"
export_base <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/anndata_exports/"

celltypes_keep_underscore <- gsub(" ", "_", celltypes_keep)

for (ct in celltypes_keep_underscore) {
  message("Exporting ", ct)

  seurat_sub <- readRDS(paste0(outdir, "combined_1kFeatures_integrated_", ct, ".rds"))
  DefaultAssay(seurat_sub) <- "RNA"
  seurat_sub <- JoinLayers(seurat_sub, assay = "RNA")

  chunk_dir <- paste0(export_base, ct, "/")
  dir.create(chunk_dir, showWarnings = FALSE, recursive = TRUE)

  # Raw counts
  counts_mat <- as(seurat_sub[["RNA"]]$counts, "dgCMatrix")
  Matrix::writeMM(counts_mat, file = paste0(chunk_dir, "counts.mtx"))

  # Normalised data (if present)
  data_mat <- tryCatch(seurat_sub[["RNA"]]$data, error = function(e) NULL)
  if (!is.null(data_mat)) {
    Matrix::writeMM(as(data_mat, "dgCMatrix"), file = paste0(chunk_dir, "data.mtx"))
  }

  # Cell metadata
  meta <- seurat_sub@meta.data
  meta$cell_barcode <- rownames(meta)
  write.csv(meta, file = paste0(chunk_dir, "metadata.csv"), row.names = FALSE, quote = TRUE)

  # Gene and cell name lists
  write.csv(data.frame(gene = rownames(counts_mat)),
            file = paste0(chunk_dir, "genes.csv"), row.names = FALSE, quote = FALSE)
  write.csv(data.frame(cell_barcode = colnames(counts_mat)),
            file = paste0(chunk_dir, "cells.csv"), row.names = FALSE, quote = FALSE)

  # Dimensionality reductions if present
  for (red in c("pca", "umap", "tsne")) {
    if (red %in% names(seurat_sub@reductions)) {
      red_df <- as.data.frame(Embeddings(seurat_sub, reduction = red))
      red_df$cell_barcode <- rownames(red_df)
      write.csv(red_df, file = paste0(chunk_dir, red, ".csv"), row.names = FALSE, quote = TRUE)
    }
  }

  rm(seurat_sub, counts_mat, data_mat, meta)
  gc()
}
```

---

## Step 10: Verify AnnData Exports

Check that cell counts in exported `.mtx` files match annotation table.

```{r verify-anndata-exports}
library(Matrix)

base_dir <- "/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/anndata_exports/"

expected_counts <- as.data.frame(table(cell_annotations_diploid$celltype)) %>%
  dplyr::rename(celltype = Var1, expected_ncells = Freq) %>%
  dplyr::mutate(expected_ncells = as.integer(expected_ncells))

celltype_dirs <- list.dirs(base_dir, full.names = TRUE, recursive = FALSE)

results_df <- bind_rows(lapply(celltype_dirs, function(ct_dir) {
  celltype_name <- basename(ct_dir)
  mtx_file      <- file.path(ct_dir, "counts.mtx")

  if (!file.exists(mtx_file)) {
    return(data.frame(celltype = celltype_name, n_rows = NA_integer_, n_cols = NA_integer_,
                      expected_ncells = expected_counts$expected_ncells[match(celltype_name, expected_counts$celltype)],
                      match = NA, stringsAsFactors = FALSE))
  }

  mat      <- readMM(mtx_file)
  expected <- expected_counts$expected_ncells[match(celltype_name, expected_counts$celltype)]

  data.frame(celltype = celltype_name, n_rows = nrow(mat), n_cols = ncol(mat),
             expected_ncells = expected, match = !is.na(expected) & (ncol(mat) == expected),
             stringsAsFactors = FALSE)
}))

results_df

# Cell types where export count does not match annotation count
results_df %>% filter(is.na(match) | match == FALSE) %>% pull(celltype)
```

---

## Step 11: UMAPs by Cell Type (Filtered Data)

Four UMAPs, all restricted to the same 20 broad cell types used throughout
the DD/DE analysis (`celltypes_keep` from Step 8, i.e. >= 1,000 cells in
both human and chimp diploid), all sharing one color-per-celltype palette so
the same cell type is the same color across panels. `celltype` is coerced to
a factor with levels locked to exactly these 20 (via `restrict_celltypes()`
below) so no stray/off-target category (e.g. `unassigned`, rare cell types
outside the analysis set) can slip into a legend or label:

1. Diploid, human + chimp combined -- **labeled directly on the UMAP**
2. Diploid, human only -- no cell type labels or legend at all
3. Diploid, chimp only -- no cell type labels or legend at all
4. Allotetraploid -- **labeled directly on the UMAP**, same style as (1)

```{r umap-setup-celltypes-and-colors}
# Exact 20 cell types used throughout the DD/DE analysis (matches
# celltypes2/celltypes_display in dd_de_interspecies_analysis.Rmd). NOT
# recomputed here via a >=1000-cells-in-both-species threshold on the
# current annotations, since that threshold currently also passes 3 extra
# cell types (Early mesoderm cells, Smooth muscle cells, Stellate cells)
# that aren't part of the actual downstream analysis set.
celltypes_keep <- c(
  "Bronchiolar and alveolar epithelial cells", "Cardiac mesoderm cells",
  "Cardiomyocytes", "Ciliated epithelial cells", "ENS glia",
  "Epicardial fat cells", "Glioblast", "Goblet cells", "Hepatoblasts",
  "Intestinal epithelial cells", "Metanephric cells", "Neuroblast",
  "Neuron", "Neuronal IPC", "Radial glia", "Retinal neurons",
  "Squamous epithelial cells", "Stromal cells",
  "Vascular endothelial cells", "iPSCs"
)

celltype_colors <- setNames(
  scales::hue_pal()(length(celltypes_keep)),
  sort(celltypes_keep)
)

# Locks `celltype` to exactly `celltypes_keep` -- any cell not in this set
# should already have been dropped by subset(), but forcing the factor
# levels too guarantees no stray level can appear in a legend/label even if
# it slips through with 0 cells.
restrict_celltypes <- function(obj) {
  obj$celltype <- factor(as.character(obj$celltype), levels = sort(celltypes_keep))
  obj
}

umap_outdir <- "/project/gilad/awchen55/mechanism/results/dispersion_paper/main/umaps/"

```

### Diploid UMAPs

```{r fit-diploid-reference-umap}
# BPCells is needed to load the object's on-disk "sketch" assay, and, as it
# turns out, some of the RNA assay's own layers too.
.libPaths(c("/project/gilad/awchen55/Rlibs/R4-4-1", .libPaths()))

seurat_data <- readRDS("/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/combined_1kFeatures_integrated_annotated.rds")

DefaultAssay(seurat_data) <- "RNA"
seurat_data[["sketch"]] <- NULL
seurat_data <- JoinLayers(seurat_data)

seurat_diploid_filt <- subset(seurat_data, subset = celltype %in% celltypes_keep) %>%
  restrict_celltypes()
rm(seurat_data)
gc()

# Seurat v5 sketch-based objects carry a sketch-subsample-only PCA ("pca",
# fit directly -- has real feature loadings) alongside full-dataset
# extensions. FindTransferAnchors (used below for the allotetraploid
# projection) needs reference.reduction to BOTH have real loadings AND
# cover every cell in the reference. "pca"/"pca.full" don't satisfy both at
# once -- but "integrated.rpca.full" does: it's the full-dataset extension
# of this object's own RPCA cross-species integration (see
# RunUMAP.RNA.integrated.rpca.full in seurat_data@commands), already has
# real loadings (2000 features) AND covers all cells, and is exactly what
# the object's original "umap.full" panel was built on -- i.e. it's already
# batch-corrected across human/chimp. (We first tried fitting a totally
# fresh, uncorrected PCA here instead of reusing this -- that reintroduced
# a human-vs-chimp batch effect into cell type cluster proportions that the
# original integration had removed.) So we just refit UMAP with
# return.model = TRUE directly on this existing reduction (the original
# "umap.full" wasn't saved with a model, so it can't be projected onto
# as-is) -- no fresh PCA needed. The human-only and chimp-only panels below
# then subset cells out of THIS single fit (never independently refit), and
# the tetraploid reference-mapping
# (https://github.com/satijalab/seurat/issues/3864) targets this same fit
# too -- so all four panels share one consistent, properly-integrated
# coordinate space.
ref_pca_name <- "integrated.rpca.full"
seurat_diploid_filt <- RunUMAP(
  seurat_diploid_filt, dims = 1:50, reduction = ref_pca_name,
  reduction.name = "refumap", reduction.key = "refUMAP_",
  return.model = TRUE
)
```

```{r plot-diploid-combined-umap, fig.width=9, fig.height=7}
p_diploid_combined <- DimPlot(
  seurat_diploid_filt, reduction = "refumap", group.by = "celltype",
  cols = celltype_colors, label = FALSE
) +
  labs(title = "", x = "UMAP 1", y = "UMAP 2") +
  NoLegend()

p_diploid_combined_labeled <- LabelClusters(
  p_diploid_combined, id = "celltype",
  repel = TRUE, fontface = "bold", size = 3.5
)
p_diploid_combined_labeled

ggsave(
  file.path(umap_outdir, "diploid_combined_umap.pdf"),
  plot = p_diploid_combined_labeled, width = 9, height = 7
)
```

```{r plot-diploid-human-umap, fig.width=9, fig.height=7}
# Subset out of the single combined fit above -- NOT an independent refit --
# so this panel's coordinates are a strict subset of the combined UMAP
# (guarantees the same cell type lands in the same place in every panel).
seurat_diploid_human <- subset(seurat_diploid_filt, subset = species == "human") %>%
  restrict_celltypes()

# No cell type labels or legend at all for the human/chimp panels --
# identification relies on the combined, labeled panel above.
p_diploid_human <- DimPlot(seurat_diploid_human, reduction = "refumap", group.by = "celltype", cols = celltype_colors) +
  labs(title = "Diploid (human)", x = "UMAP 1", y = "UMAP 2") +
  NoLegend()
p_diploid_human

ggsave(
  file.path(umap_outdir, "diploid_human_umap.pdf"),
  plot = p_diploid_human, width = 9, height = 7
)
```

```{r plot-diploid-chimp-umap, fig.width=9, fig.height=7}
seurat_diploid_chimp <- subset(seurat_diploid_filt, subset = species == "chimp") %>%
  restrict_celltypes()

p_diploid_chimp <- DimPlot(seurat_diploid_chimp, reduction = "refumap", group.by = "celltype", cols = celltype_colors) +
  labs(title = "Diploid (chimp)", x = "UMAP 1", y = "UMAP 2") +
  NoLegend()
p_diploid_chimp

ggsave(
  file.path(umap_outdir, "diploid_chimp_umap.pdf"),
  plot = p_diploid_chimp, width = 9, height = 7
)
```

### Allotetraploid UMAP

Rather than fitting an independent UMAP for the allotetraploid data (which
would land in an arbitrary, incomparable coordinate space), we **project**
it into the same combined-diploid reference UMAP (`seurat_diploid_filt`,
`refumap`) that the combined/human/chimp panels above all share, via
Seurat's anchor-based reference mapping (`FindTransferAnchors` + `MapQuery`;
see https://github.com/satijalab/seurat/issues/3864). Cell type coloring
still uses the allotetraploid's own CelliD calls (Step 3) -- the projection
only supplies the x/y placement, not the label.

Each hybrid cell contributes two rows here -- one human-allele and one
chimp-allele pseudo-observation (`species` column) -- rather than two
distinct organisms as in the diploid data (same convention used for cell
counts elsewhere in this document, e.g. Step 6). We show only one species
(**human**) per physical cell so each cell appears once in the UMAP.

```{r load-tetraploid-for-projection}
tetraploid_obj <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_combined_variablfeatures_filtered.rds")

DefaultAssay(tetraploid_obj) <- "RNA"
tetraploid_filt <- subset(tetraploid_obj, subset = celltype %in% celltypes_keep & species == "human") %>%
  restrict_celltypes()
rm(tetraploid_obj)
gc()

tetraploid_filt <- NormalizeData(tetraploid_filt)
```

```{r map-tetraploid-onto-diploid-umap}
# FindTransferAnchors parallelizes internally via `future`; with the
# reference now the full combined (human + chimp) object rather than just
# human, the per-worker payload exceeds the default 500 MiB `future`
# transfer limit. Raise it rather than letting the call fail.
options(future.globals.maxSize = 8000 * 1024^2)

anchors <- FindTransferAnchors(
  reference = seurat_diploid_filt,
  query = tetraploid_filt,
  dims = 1:50,
  reference.reduction = ref_pca_name
)

tetraploid_mapped <- MapQuery(
  anchorset = anchors,
  reference = seurat_diploid_filt,
  query = tetraploid_filt,
  refdata = list(predicted.celltype = "celltype"),
  reference.reduction = ref_pca_name,
  reduction.model = "refumap"
)

# MapQuery's own metadata handling can drop factor levels -- reapply so the
# palette/legend stay locked to exactly celltypes_keep.
tetraploid_mapped <- restrict_celltypes(tetraploid_mapped)
```

```{r plot-allotetraploid-umap, fig.width=9, fig.height=7}
p_tetraploid <- DimPlot(
  tetraploid_mapped, reduction = "ref.umap", group.by = "celltype",
  cols = celltype_colors, label = FALSE
) +
  labs(title = "", x = "UMAP 1", y = "UMAP 2") +
  NoLegend()

p_tetraploid_labeled <- LabelClusters(
  p_tetraploid, id = "celltype",
  repel = TRUE, fontface = "bold", size = 3.5
)
p_tetraploid_labeled

ggsave(
  file.path(umap_outdir, "allotetraploid_umap.pdf"),
  plot = p_tetraploid_labeled, width = 9, height = 7
)
```
