---
title: "Allotetraploid Object Creation and Cell Annotation"
output: html_notebook
date: "2026-05-01"
description: >
  Builds the allotetraploid Seurat object from per-lane STARsolo outputs
  (humanized/human/chimp allele-specific alignments from Kenneth's
  pipeline), annotates cell types with CelliD on the humanized allele, then
  merges the human and chimp allele counts into one combined,
  shared-barcode-filtered object used throughout the DD/DE and cis/trans
  mechanism analyses.
---

## Overview

Creates and annotates Seurat objects for the **allotetraploid** cell lines.
The allotetraploid system allows cis vs. trans regulatory dissection: each cell
contains both human and chimp genomes, and reads are separated by allele of
origin during alignment.

Three alignment outputs are used per lane:
- **humanized** — reads assigned to the human allele of a chimped genome (used for CelliD annotation)
- **human** — reads from the human genome
- **chimp** — reads from the chimp genome

**Workflow:**
1. Load humanized STARsolo counts (8 lanes), create BPCells-backed Seurat objects
2. QC plots; normalize and find variable features
3. Run CelliD (MCA + hypergeometric test) for cell type annotation on humanized data
4. Load human and chimp allele objects separately
5. Find cells present in all three objects (humanized, human, chimp)
6. Merge human + chimp into a combined object; filter to shared-barcode cells
7. Save annotated combined object and export per-cell-type AnnData files

**Outputs:**
- `tetraploid_cell_annotations.csv` — per-cell barcode + cell type label
- `tetraploid_metadata.csv` — full metadata for filtered combined object
- `tetraploid_combined_variablfeatures_filtered.rds` — final analysis-ready object
- Per-cell-type AnnData exports under `data/tetraploid/anndata_exports/`

---

## Load Packages

`BPCells` and `CelliD` are Bioconductor/GitHub packages
(`BiocManager::install("BPCells")`, `BiocManager::install("CelliD")`), not
CRAN; everything else here is CRAN.

```{r load-packages, warning=FALSE, message=FALSE}
require(Seurat)
require(dplyr)
require(parallel)
require(BPCells)
library(ggplot2)
library(CelliD)
library(stringr)
```

---

## Helper Function: Create Seurat Object from STARsolo Output

Reads count matrices from a list of STARsolo raw output directories, filters to
cells with > 1,000 features, and merges all lanes into one object with
BPCells-backed count storage.

```{r create-object-helper}
#' Build one Seurat object from a set of STARsolo raw-output directories.
#'
#' Reads each directory's counts via `ReadMtx`, filters to cells with
#' > 1,000 detected features, and merges all directories (lanes) into a
#' single object. Used identically for the humanized, human, and chimp
#' allele alignments -- only the input directory list and `cell_id` prefix
#' differ between calls.
#'
#' @param counts Character vector of STARsolo raw-output directory paths
#'   (each containing `matrix.mtx`, `barcodes.tsv`, `features.tsv`), one per
#'   lane to merge.
#' @param cell_id String prepended to every cell barcode in this call (with
#'   a space, per `paste()`'s default separator) before merging -- callers
#'   in this document pass `""`, i.e. no prefix, and add a species prefix
#'   separately later via `RenameCells()`.
#' @return A merged Seurat object (layers joined) with counts in the `RNA`
#'   assay; a single directory's object is returned as-is if only one lane
#'   is provided.
create_object <- function(counts, cell_id) {
  obj_list <- list()
  for (i in seq_along(counts)) {
    obj <- ReadMtx(
      mtx      = paste0(counts[i], "/matrix.mtx"),
      cells    = paste0(counts[i], "/barcodes.tsv"),
      features = paste0(counts[i], "/features.tsv")
    )
    colnames(obj) <- paste(cell_id, colnames(obj))
    obj <- CreateSeuratObject(obj)
    obj <- subset(obj, subset = nFeature_RNA > 1000)
    obj_list[[i]] <- obj
  }
  if (length(obj_list) > 1) {
    obj <- merge(obj_list[[1]], obj_list[2:length(obj_list)])
    obj <- JoinLayers(obj)
    return(obj)
  } else {
    return(obj_list[[1]])
  }
}
```

---

## Step 1: Load Humanized Allotetraploid Data (8 Lanes)

STARsolo outputs are in `.humanized/Solo.out/GeneFull_Ex50pAS/raw`. Counts are
stored on disk using BPCells to keep memory usage manageable.

```{r load-humanized-data-8-lanes}
bpcells_base <- '/project/gilad/awchen55/mechanism/data/tetraploid/'

obj_list <- mclapply(1:8, function(i) {
  lane     <- paste0(i)
  base_dir <- paste0("/project/gilad/kenneth/scale_project/tetraploid/alignment/",
                     "Sample-Tetraploid.QSR-", lane, "/Sample-Tetraploid.QSR-", lane,
                     ".star.align/split_bams/")

  dirs      <- list.dirs(base_dir)
  this.dirs <- dirs[grep(".humanized/Solo.out/GeneFull_Ex50pAS/raw", dirs, fixed = TRUE)]

  obj <- create_object(this.dirs, "")

  # Write counts to BPCells on-disk format and reopen
  this.bpcells <- paste0(bpcells_base, "tetraploid_", lane)
  write_matrix_dir(mat = convert_matrix_type(obj[["RNA"]]$counts, "uint32_t"), dir = this.bpcells)
  obj[["RNA"]]$counts <- open_matrix_dir(dir = this.bpcells)

  obj$well <- lane
  obj
}, mc.cores = 8)

# Merge all 8 lanes into one object
obj <- merge(obj_list[[1]], obj_list[2:8])
obj <- JoinLayers(obj)
saveRDS(obj, paste0(bpcells_base, "tetraploid.rds"))
```

---

## Step 2: QC — Humanized Object

```{r qc-humanized-nfeature}
obj_test <- readRDS(paste0("/project/gilad/awchen55/mechanism/data/tetraploid/", "tetraploid.rds"))

VlnPlot(obj_test, "nFeature_RNA", pt.size = 0, group.by = "well", log = TRUE) +
  geom_hline(yintercept = 1500, linetype = 2)
```

```{r qc-humanized-percent-mt-rp-malat1}
obj_test[["percent.mt"]]     <- PercentageFeatureSet(obj_test, pattern = "^MT-")
obj_test[["percent.rp"]]     <- PercentageFeatureSet(obj_test, pattern = "^RPS|^RPL")
obj_test[["percent.malat1"]] <- PercentageFeatureSet(obj_test, pattern = "MALAT1")

VlnPlot(obj_test, "percent.mt",     pt.size = 0, group.by = "well")
VlnPlot(obj_test, "percent.rp",     pt.size = 0, group.by = "well")
VlnPlot(obj_test, "percent.malat1", pt.size = 0, group.by = "well")
```

```{r normalize-humanized-and-save}
# Normalize and find variable features for use in CelliD
obj_test <- NormalizeData(obj_test)
obj_test <- FindVariableFeatures(obj_test, nfeatures = 3000)
saveRDS(obj_test, file = "/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_humanized_variablefeatures.rds")
saveRDS(obj_test, file = "/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_humanized.rds")
```

---

## Step 3: Load Cell Type Reference Signatures

Same three-source signature set used for diploid annotation (FCA + Linnarsson
brain atlas + Gilad lab cardiac signatures). See `cell_annotations_diploid_lines.Rmd`
for full description.

```{r load-reference-signatures}
brain_fca <- c("Purkinje neurons", "Inhibitory neurons", "Astrocytes",
               "Granule neurons", "Oligodendrocytes", "Inhibitory interneurons",
               "Microglia", "Limbic system neurons", "Excitatory neurons",
               "Unipolar brush cells")

fca_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/train_test/Cao_FCA_1Kcell_train_sce_250_signature_common_20iterate_cut500reads_noCTRiboMT.rds")

fca_signatures[["iPSCs"]] <- c("SOX2", "POU5F1", "NANOG", "NODAL", "ZFP42",
                                "DNMT3B", "UTF1", "LCK", "DPPA2", "IDO1",
                                "DPPA5", "LEFTY1", "ESRG", "CNMD", "SFRP2",
                                "THY1", "USP44", "VSNL1", "SPP1", "KLF4",
                                "GDF3", "L1TD1", "VRTN", "ZSCAN10")

linn_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/human_brain/brain_signature_reiterated_1000_Celltype_TopLevelCluster_noRiboCCMT.rds")

CM_signatures <- readRDS("/project2/gilad/wenhe/EB/combined_3_4/data/celltype_annotate/CM_Reem/CM_signature_reiterated_1000_noRiboCCMT.rds")
CM_signatures <- CM_signatures[2:4]
names(CM_signatures) <- c("Early mesoderm cells", "Cardiac mesoderm cells", "Cardiac progenitor cells")

fca_signatures <- c(fca_signatures[!names(fca_signatures) %in% brain_fca], linn_signatures, CM_signatures)
fca_signatures <- lapply(fca_signatures, function(genes) {
  genes[!genes %in% unlist(cc.genes.updated.2019) & genes %in% HgProteinCodingGenes]
})
```

---

## Step 4: Run CelliD on Humanized Data (2 Chunks)

Split into 2 chunks to manage memory during MCA embedding.

```{r split-humanized-into-chunks}
obj_humanized <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_humanized_variablefeatures.rds")

all_cells  <- colnames(obj_humanized)
set.seed(1)
cell_groups <- split(all_cells, rep(1:2, length.out = length(all_cells)))

obj_list <- lapply(cell_groups, function(cells_use) subset(obj_humanized, cells = cells_use))
sapply(obj_list, ncol)

outdir <- "/project/gilad/awchen55/mechanism/data/tetraploid/"
for (i in seq_along(obj_list)) {
  saveRDS(obj_list[[i]],
          file = paste0(outdir, "tetraploid_humanized_variablefeatures_chunk", i, ".rds"))
}
```

```{r run-cellid-per-chunk}
data_inputs     <- c('tetraploid_humanized_variablefeatures_chunk1.rds',
                     'tetraploid_humanized_variablefeatures_chunk2.rds')
outdir          <- "/project/gilad/awchen55/mechanism/data/tetraploid/"
all_cellid_orig <- c()

for (di in data_inputs) {
  obj <- readRDS(paste0(outdir, di))

  DefaultAssay(obj) <- "RNA"
  obj[["RNA"]] <- JoinLayers(obj[["RNA"]])

  # Filter to protein-coding genes expressed in >5 cells
  counts_mat <- LayerData(obj, assay = "RNA", layer = "counts")
  genes_use  <- intersect(rownames(counts_mat), HgProteinCodingGenes)
  obj        <- subset(obj, features = genes_use)
  counts_mat <- LayerData(obj, assay = "RNA", layer = "counts")
  gene_keep  <- rownames(counts_mat)[Matrix::rowSums(counts_mat > 0) > 5]
  obj        <- subset(obj, features = gene_keep)

  # HVG — exclude MT, ribosomal, and cell cycle genes
  hvg <- VariableFeatures(FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000))
  hvg <- hvg[!grepl("^RPL|^RPS|^MT-", hvg) & !hvg %in% unlist(cc.genes.updated.2019)]

  obj_hvg <- c(unlist(fca_signatures), hvg) %>% unique() %>% intersect(rownames(obj))
  obj     <- ScaleData(obj, features = obj_hvg)

  print(paste("Running MCA on", di))
  obj <- RunMCA(obj, nmcs = 50, assay = "RNA", slot = "data", features = obj_hvg)

  cellid_enrichments <- RunCellHGT(obj, pathways = fca_signatures, dims = 1:50, n.features = 300)

  cellid_labels <- rownames(cellid_enrichments)[apply(cellid_enrichments, 2, which.max)]
  cellid_labels <- ifelse(apply(cellid_enrichments, 2, max) >= 2, cellid_labels, "unassigned")

  cellid_orig <- as_tibble(cellid_labels, rownames = "cell_barcode") %>%
    dplyr::select(cell_barcode, celltype = value)
  cellid_orig$celltype <- sub("\\s*\\d+$", "", cellid_orig$celltype)

  all_cellid_orig <- rbind(all_cellid_orig, cellid_orig)
}
```

---

## Step 5: Remap Labels and Save Tetraploid Annotations

Same remapping logic as diploid — fine-grained reference categories collapsed
to the 20 broad cell types used in the DD/DE analysis.

```{r remap-celltype-labels-and-save}
all_cellid_orig_merged <- all_cellid_orig %>%
  mutate(celltype = replace(celltype, celltype == "Visceral neurons",            "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Chromaffin cells",            "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Sympathoblasts",              "ENS neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Schwann cells",               "ENS glia")) %>%
  mutate(celltype = replace(celltype, celltype == "Ureteric bud cells",          "Metanephric cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Bipolar cells",               "Photoreceptor cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Thymocytes",                  "Lymphoid cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Mesothelial cells",           "Epicardial fat cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Amacrine cells",              "Horizontal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Ganglion cells",              "Horizontal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Lymphatic endothelial cells", "Vascular endothelial cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Mesangial cells",             "Stromal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Fibroblast",                  "Stromal cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Parietal and chief cells",    "Goblet cells")) %>%
  mutate(celltype = replace(celltype, celltype == "Cardiac progenitor cells",    "Cardiomyocytes")) %>%
  mutate(celltype = replace(celltype, celltype == "Horizontal cells",            "Retinal neurons")) %>%
  mutate(celltype = replace(celltype, celltype == "Photoreceptor cells",         "Retinal neurons"))

outpath <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
write.csv(all_cellid_orig_merged, paste0(outpath, "tetraploid_cell_annotations.csv"), row.names = FALSE)
```

```{r plot-celltype-distribution}
cell_annotations_tetraploid <- read.csv(paste0(outpath, "tetraploid_cell_annotations.csv"))

# Distribution of cell types
plot_df <- cell_annotations_tetraploid %>%
  dplyr::count(celltype, name = "n_cells") %>%
  dplyr::arrange(desc(n_cells)) %>%
  dplyr::mutate(celltype = factor(celltype, levels = celltype))

ggplot(plot_df, aes(x = celltype, y = n_cells)) +
  geom_col(fill = "steelblue") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Cell type", y = "Number of cells")

table(cell_annotations_tetraploid$celltype)
```

---

## Step 6: Load Human and Chimp Allele Objects

Each lane was aligned separately to the human and chimp genomes. Load and merge
across 8 lanes using the same BPCells-backed approach.

```{r load-human-allele-data}
bpcells_base <- '/project/gilad/awchen55/mechanism/data/tetraploid/'

# Human allele reads
obj_list <- mclapply(1:8, function(i) {
  lane     <- paste0(i)
  base_dir <- paste0("/project/gilad/kenneth/scale_project/tetraploid/alignment/",
                     "Sample-Tetraploid.QSR-", lane, "/Sample-Tetraploid.QSR-", lane,
                     ".star.align/split_bams/")
  dirs      <- list.dirs(base_dir)
  this.dirs <- dirs[grep(".human/Solo.out/GeneFull_Ex50pAS/raw", dirs, fixed = TRUE)]
  obj       <- create_object(this.dirs, "")
  this.bpcells <- paste0(bpcells_base, "tetraploid_human_", lane)
  write_matrix_dir(mat = convert_matrix_type(obj[["RNA"]]$counts, "uint32_t"), dir = this.bpcells)
  obj[["RNA"]]$counts <- open_matrix_dir(dir = this.bpcells)
  obj$well <- lane; obj$species <- "human"
  obj
}, mc.cores = 8)
obj_human <- merge(obj_list[[1]], obj_list[2:8])
obj_human <- JoinLayers(obj_human)
saveRDS(obj_human, paste0(bpcells_base, "tetraploid_human.rds"))
```

```{r load-chimp-allele-data}
# Chimp allele reads
obj_list <- mclapply(1:8, function(i) {
  lane     <- paste0(i)
  base_dir <- paste0("/project/gilad/kenneth/scale_project/tetraploid/alignment/",
                     "Sample-Tetraploid.QSR-", lane, "/Sample-Tetraploid.QSR-", lane,
                     ".star.align/split_bams/")
  dirs      <- list.dirs(base_dir)
  this.dirs <- dirs[grep(".chimp/Solo.out/GeneFull_Ex50pAS/raw", dirs, fixed = TRUE)]
  obj       <- create_object(this.dirs, "")
  this.bpcells <- paste0(bpcells_base, "tetraploid_chimp_", lane)
  write_matrix_dir(mat = convert_matrix_type(obj[["RNA"]]$counts, "uint32_t"), dir = this.bpcells)
  obj[["RNA"]]$counts <- open_matrix_dir(dir = this.bpcells)
  obj$well <- lane; obj$species <- "chimp"
  obj
}, mc.cores = 8)
obj_chimp <- merge(obj_list[[1]], obj_list[2:8])
obj_chimp <- JoinLayers(obj_chimp)
saveRDS(obj_chimp, paste0(bpcells_base, "tetraploid_chimp.rds"))
```

---

## Step 7: Match Barcodes Across All Three Objects

Cells must have a matching barcode in humanized, human, and chimp objects.
This ensures each cell has expression data from both genome copies.

```{r match-barcodes-across-objects}
obj_humanized <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_humanized.rds")
obj_human     <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_human.rds")
obj_chimp     <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_chimp.rds")

DefaultAssay(obj_humanized) <- "RNA"
DefaultAssay(obj_human)     <- "RNA"
DefaultAssay(obj_chimp)     <- "RNA"

counts_humanized <- GetAssayData(obj_humanized, slot = "counts")
counts_human     <- GetAssayData(obj_human,     slot = "counts")
counts_chimp     <- GetAssayData(obj_chimp,     slot = "counts")

# Strip species prefixes to get raw barcodes for matching
colnames(counts_humanized) <- stringr::str_trim(colnames(counts_humanized))
colnames(counts_human)     <- stringr::str_remove(colnames(counts_human), "^human_\\s*") %>% stringr::str_trim()
colnames(counts_chimp)     <- stringr::str_remove(colnames(counts_chimp), "^chimp_\\s*") %>% stringr::str_trim()

common_barcodes <- Reduce(intersect, list(
  colnames(counts_humanized),
  colnames(counts_human),
  colnames(counts_chimp)
))

cat("Cells with matching barcodes across all three objects:", length(common_barcodes), "\n")

# Check UMI allocation across alleles
df <- data.frame(
  barcode       = common_barcodes,
  humanized_umi = Matrix::colSums(counts_humanized[, common_barcodes]),
  human_umi     = Matrix::colSums(counts_human[,     common_barcodes]),
  chimp_umi     = Matrix::colSums(counts_chimp[,     common_barcodes])
) %>%
  dplyr::mutate(
    percent_human_kept = human_umi / humanized_umi,
    percent_chimp_kept = chimp_umi / humanized_umi
  )

cat("Global % UMIs assigned to human allele:", round(100 * sum(df$human_umi) / sum(df$humanized_umi), 1), "\n")
cat("Global % UMIs assigned to chimp allele:", round(100 * sum(df$chimp_umi) / sum(df$humanized_umi), 1), "\n")
```

---

## Step 8: Add Cell Annotations and Build Combined Object

```{r transfer-annotations-to-allele-objects}
# Transfer tetraploid cell type labels onto human and chimp allele objects
anno              <- cell_annotations_tetraploid
rownames(anno)    <- anno$cell_barcode
anno              <- anno[, "celltype", drop = FALSE]

obj_human <- AddMetaData(obj_human, metadata = anno)
obj_chimp <- AddMetaData(obj_chimp, metadata = anno)

table(obj_human@meta.data$celltype)
table(obj_chimp@meta.data$celltype)
```

```{r merge-human-chimp-objects}
# Add species prefix to cell barcodes and merge into one object
obj_chimp <- RenameCells(obj_chimp, add.cell.id = "chimp")
obj_human <- RenameCells(obj_human, add.cell.id = "human")

obj_tetraploid <- merge(x = obj_chimp, y = obj_human)
obj_tetraploid <- JoinLayers(obj_tetraploid)
```

---

## Step 9: Filter Genes and Cells

Remove genes expressed in only one species (these cannot be used for
cis/trans comparison). Filter to cells where the same barcode exists in
both human and chimp allele data.

```{r filter-species-specific-genes}
obj_tetraploid[["percent.mt"]]     <- PercentageFeatureSet(obj_tetraploid, pattern = "^MT-")
obj_tetraploid[["percent.rp"]]     <- PercentageFeatureSet(obj_tetraploid, pattern = "^RPS|^RPL")
obj_tetraploid[["percent.malat1"]] <- PercentageFeatureSet(obj_tetraploid, pattern = "MALAT1")

# Remove genes with zero expression in either species
agg        <- AggregateExpression(obj_tetraploid, group.by = "species")
rna        <- as.data.frame(agg$RNA)
only_human <- rownames(rna %>% filter(chimp == 0 & human > 0))
only_chimp <- rownames(rna %>% filter(chimp > 0 & human == 0))
neither    <- rownames(rna %>% filter(chimp == 0 & human == 0))
remove     <- unique(c(only_human, only_chimp, neither))

cat("Genes removed (species-specific or absent):", length(remove), "\n")
cat("Genes kept:", nrow(rna) - length(remove), "\n")

obj_tetraploid <- subset(obj_tetraploid, features = setdiff(rownames(rna), remove))
saveRDS(obj_tetraploid, file = "/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_combined.rds")
```

```{r filter-shared-barcode-cells}
# Keep only cells with matching barcodes in both human and chimp
cells   <- colnames(obj_tetraploid)
species <- ifelse(grepl("^chimp_", cells), "chimp", ifelse(grepl("^human_", cells), "human", NA))
barcode <- sub("^(chimp_|human_)", "", cells)

shared_barcodes <- data.frame(cell = cells, species = species, barcode = barcode) %>%
  dplyr::group_by(barcode) %>%
  dplyr::filter(dplyr::n_distinct(species) == 2) %>%
  dplyr::pull(barcode) %>%
  unique()

cells_keep     <- cells[sub("^(chimp_|human_)", "", cells) %in% shared_barcodes]
obj_filtered   <- subset(obj_tetraploid, cells = cells_keep)

# Additional QC: keep cells with > 1,000 features
obj_filtered <- subset(obj_filtered, subset = nFeature_RNA > 1000)

table(obj_filtered@meta.data$species, obj_filtered@meta.data$celltype)
dim(obj_filtered)
```

```{r normalize-and-save-filtered-object}
# Normalize and find variable features on the filtered combined object
obj_filtered <- NormalizeData(obj_filtered)
obj_filtered <- FindVariableFeatures(obj_filtered, nfeatures = 3000)
saveRDS(obj_filtered, file = "/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_combined_variablfeatures_filtered.rds")

# Save full metadata table
metadata_tetraploid <- obj_filtered@meta.data %>%
  tibble::rownames_to_column(var = "cell_barcode") %>%
  dplyr::select(cell_barcode, dplyr::everything())

write.csv(metadata_tetraploid,
          paste0('/project/gilad/awchen55/mechanism/data/cell_annotations/', "tetraploid_metadata.csv"),
          row.names = FALSE)
```

---

## Step 10: Cell Count Summary — Diploid vs Tetraploid

Compare cell counts across species and contexts. Filters applied:
- Diploid: ≥ 1,000 cells in both human and chimp
- Tetraploid: ≥ 300 cells

```{r plot-diploid-vs-tetraploid-counts}
path <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
cell_annotations_diploid    <- read.csv(paste0(path, 'diploid_cell_annotations.csv'))
metadata_tetraploid         <- read.csv(paste0(path, 'tetraploid_metadata.csv'))

diploid_counts <- cell_annotations_diploid %>%
  dplyr::mutate(
    species = dplyr::case_when(
      stringr::str_starts(cell_barcode, "chimp") ~ "Chimpanzee",
      stringr::str_starts(cell_barcode, "human") ~ "Human",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(species), celltype != "unassigned") %>%
  dplyr::count(celltype, species, name = "cell_count")

# Count each cell only once (human allele copy) to avoid double-counting
tetraploid_counts <- metadata_tetraploid %>%
  dplyr::filter(stringr::str_starts(cell_barcode, "human_"), celltype != "unassigned") %>%
  dplyr::count(celltype, name = "cell_count") %>%
  dplyr::mutate(species = "Tetraploid")

plot_df <- dplyr::bind_rows(diploid_counts, tetraploid_counts) %>%
  tidyr::pivot_wider(names_from = species, values_from = cell_count, values_fill = 0) %>%
  dplyr::filter(Human >= 1000, Chimpanzee >= 1000, Tetraploid >= 300) %>%
  tidyr::pivot_longer(cols = c(Human, Chimpanzee, Tetraploid),
                      names_to = "species", values_to = "cell_count") %>%
  dplyr::group_by(celltype) %>%
  dplyr::mutate(total_cells = sum(cell_count)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    celltype = reorder(celltype, -total_cells),
    species  = factor(species, levels = c("Human", "Chimpanzee", "Tetraploid"))
  )

ggplot(plot_df, aes(x = celltype, y = cell_count, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = cell_count), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Human" = "#D55E00", "Chimpanzee" = "#0072B2", "Tetraploid" = "#009E73")) +
  labs(x = "", y = "Number of Cells", fill = "") +
  theme_classic() +
  theme(
    axis.text.x       = element_text(angle = 45, hjust = 1),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position   = c(0.88, 0.9),
    legend.background = element_blank(),
    legend.key        = element_blank()
  ) +
  expand_limits(y = max(plot_df$cell_count) * 1.1)
```

---

## Step 11: Split by Cell Type and Export to AnnData

```{r split-by-celltype}
seurat_data <- readRDS("/project/gilad/awchen55/mechanism/data/tetraploid/tetraploid_combined_variablfeatures_filtered.rds")

seurat_list <- SplitObject(seurat_data, split.by = "celltype")

ct_outdir <- "/project/gilad/awchen55/mechanism/data/tetraploid/celltype/"
for (ct in names(seurat_list)) {
  saveRDS(seurat_list[[ct]],
          file = file.path(ct_outdir,
                           paste0("tetraploid_combined_variablfeatures_filtered_", ct, ".rds")))
}
```

```{r export-to-anndata}
# Cell types meeting minimum thresholds
celltypes_keep <- dplyr::bind_rows(diploid_counts, tetraploid_counts) %>%
  tidyr::pivot_wider(names_from = species, values_from = cell_count, values_fill = 0) %>%
  dplyr::filter(Human >= 1000, Chimpanzee >= 1000, Tetraploid >= 300) %>%
  dplyr::pull(celltype)

ct_outdir   <- "/project/gilad/awchen55/mechanism/data/tetraploid/celltype/"
export_base <- "/project/gilad/awchen55/mechanism/data/tetraploid/anndata_exports/"

for (ct in celltypes_keep) {
  ct2 <- gsub(" ", "_", ct)
  message("Exporting ", ct)

  seurat_sub <- readRDS(paste0(ct_outdir, "tetraploid_combined_variablfeatures_filtered_", ct, ".rds"))
  DefaultAssay(seurat_sub) <- "RNA"
  seurat_sub <- JoinLayers(seurat_sub, assay = "RNA")

  chunk_dir <- paste0(export_base, ct2, "/")
  dir.create(chunk_dir, showWarnings = FALSE, recursive = TRUE)

  counts_mat <- as(seurat_sub[["RNA"]]$counts, "dgCMatrix")
  Matrix::writeMM(counts_mat, file = paste0(chunk_dir, "counts.mtx"))

  data_mat <- tryCatch(seurat_sub[["RNA"]]$data, error = function(e) NULL)
  if (!is.null(data_mat)) Matrix::writeMM(as(data_mat, "dgCMatrix"), file = paste0(chunk_dir, "data.mtx"))

  meta <- seurat_sub@meta.data
  meta$cell_barcode <- rownames(meta)
  write.csv(meta, file = paste0(chunk_dir, "metadata.csv"), row.names = FALSE, quote = TRUE)

  write.csv(data.frame(gene = rownames(counts_mat)),
            file = paste0(chunk_dir, "genes.csv"), row.names = FALSE, quote = FALSE)
  write.csv(data.frame(cell_barcode = colnames(counts_mat)),
            file = paste0(chunk_dir, "cells.csv"), row.names = FALSE, quote = FALSE)

  for (red in c("pca", "umap", "tsne")) {
    if (red %in% names(seurat_sub@reductions)) {
      red_df <- as.data.frame(Embeddings(seurat_sub, reduction = red))
      red_df$cell_barcode <- rownames(red_df)
      write.csv(red_df, file = paste0(chunk_dir, red, ".csv"), row.names = FALSE, quote = TRUE)
    }
  }

  rm(seurat_sub, counts_mat, data_mat, meta)
  gc()
}
```
