---
title: "Dispersion Estimate Diagnostics"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
date: "2026-07-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(purrr)

MEMENTO_DIR <- "/project/gilad/brendan/dispersion/pilot/cHDC_data/cellranger_cluster-mode_trial/analysis/datasets/memento"

CELL_TYPES <- c("Cardiac_fibroblasts", "Activated_fibroblasts", "Cardiomyocytes",
                "Epithelial_cells", "Mesenchymal_cells", "Foregut_endodermal_cells")
INDIVIDUALS <- c("NA19099", "NA19093", "NA18522")

theme_diag <- theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        strip.background = element_blank())

# Replace underscores with spaces when displaying cell type names in titles/labels
ct_label <- function(x) gsub("_", " ", x)

results_path <- '/project/gilad/awchen55/differentialDispersion/analysis/dispersion_diagnostics/figure/'
```

## Load Data

```{r load}
res_var_wide <- read.csv(
  file.path(MEMENTO_DIR, "new_res_var_by_cell_type_ind_1000_cts_new_predicted_id.csv"),
  row.names = 1, check.names = FALSE
)
mean_wide <- read.csv(
  file.path(MEMENTO_DIR, "new_mean_expression_by_cell_type_ind_1000_cts_new_predicted_id.csv"),
  row.names = 1, check.names = FALSE
)

cat("Residual variance matrix:", nrow(res_var_wide), "genes ×", ncol(res_var_wide), "columns\n")
cat("Mean expression matrix:  ", nrow(mean_wide),    "genes ×", ncol(mean_wide),    "columns\n")
cat("Columns:", paste(colnames(res_var_wide), collapse = ", "), "\n")
```

```{r tidy}
# Pivot to long format: gene, individual, celltype, res_var, mean
rv_long <- res_var_wide |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "res_var") |>
  mutate(individual = sub("_.*", "", sample),
         celltype   = sub("^[^_]+_", "", sample))

mn_long <- mean_wide |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "mean_expr") |>
  mutate(individual = sub("_.*", "", sample),
         celltype   = sub("^[^_]+_", "", sample))

df <- inner_join(rv_long, mn_long, by = c("gene", "sample", "individual", "celltype")) |>
  filter(!is.na(res_var), !is.na(mean_expr))

cat("Long-format rows:", nrow(df), "\n")
```

Shared filters and a per-cell-type-averaged view of `df`, used by several
sections below (Section 5 introduces the expression cutoff itself; Section
3 applies it together with the mito/ribo gene filter).

```{r shared_filters}
CUTOFF <- -6

# Mitochondrial (MT-), mitochondrial ribosomal (MRPL/MRPS), and ribosomal
# protein (RPL/RPS) genes.
is_mito_ribo <- function(g) grepl("^MT-|^MRPL|^MRPS|^RPL|^RPS", g)

# One row per gene x cell type (mean expression averaged across individuals),
# for Section 5's per-cell-type cutoff table.
df_ct_avg <- df |>
  group_by(gene, celltype) |>
  summarise(mean_expr = mean(mean_expr, na.rm = TRUE), .groups = "drop")
```

---


# 1. Total Variability Decomposition

What fraction of a gene's total (raw 2nd-moment) variability is captured by
memento's mean-corrected dispersion estimate? Three quantities, all computed
per gene x cell type x ploidy x species x donor (diploid) or x species
(tetraploid, no donor level) by `memento_variance_decomposition.py`
(mechanism project):

## How memento estimates mean and dispersion

Memento estimates each gene's mean and variance with a hypergeometric
method-of-moments estimator, modeling observed UMI counts as a
hypergeometric sample of a cell's true transcript pool (Kim et al. 2024,
PMC11556465, Equation 8):

$$\hat{\mu}_g = \frac{1}{n_{cells}} \sum_c \frac{Y_{cg}}{N_c \, q}$$

$$\hat{\sigma}_g^2 = \frac{1}{n_{cells}} \sum_c \frac{Y_{cg}^2 - Y_{cg}(1 - q)}{N_c^2 \, q^2} - \hat{\mu}_g^2$$

where, for gene $g$:

- $Y_{cg}$ -- the observed (sequenced) UMI count for gene $g$ in cell $c$.
- $N_c$ -- the total UMI count (total transcript count) of cell $c$.
- $q$ -- the capture efficiency: the assumed probability that a true
  transcript molecule in a cell is captured and sequenced (`memento_q` in
  this project's pipeline config; the same value is used for every cell in
  a given memento run).
- $n_{cells}$ -- the number of cells in the group being estimated (e.g.
  one species x donor x cell-type sample).
- $\hat{\mu}_g$ -- the estimated mean (true relative expression) of gene
  $g$.
- $\hat{\sigma}_g^2$ -- the estimated variance of gene $g$.

The $Y_{cg}(1 - q)$ term corrects the raw second moment for the extra
sampling variance introduced purely by imperfect capture -- without it,
$\hat{\sigma}_g^2$ would be a biased, capture-noise-inflated estimate of
the gene's true variance, and cells with lower $q$ would look spuriously
more variable for purely technical reasons.

Memento's actual implementation (`_hyper_1d_relative` in `estimator.py`,
the function this project's pipeline calls) computes an algebraically
equivalent quantity in *relative* units: it doesn't divide the mean by $q$,
and its raw second moment isn't divided by $q^2$ -- i.e. it computes
$q \, \hat{\mu}_g$ and $q^2 \, \hat{\sigma}_g^2$ from the equations above.
Since the same $q$ is used for every gene within one memento run, this is a
constant rescaling (an additive shift in log space) that leaves the
mean-variance trend fit's residuals -- and therefore `log.dispersion`,
defined below -- exactly unchanged.

This per-gene `(mean, variance)` pair -- one point per gene -- is what
feeds the mean-variance trend fit and dispersion definition below.

Final column set in `all_celltypes_variance_genes.csv` (`gene`, `celltype`,
`ploidy`, `species`, `donor` identify the row; the rest are described
below, in file order):

- **`log.mean`** -- $\log(\hat{\mu}_g)$, memento's own log(mean) estimate
  (the mean value `get_1d_moments()` normally returns), unchanged from the
  rest of this project's convention.
- **`log.dispersion`** -- memento's own mean-corrected dispersion (residual
  variance), matching this project's standard convention. Writing
  $V = \hat{\sigma}_g^2$ for the gene's raw (total) variance and $E$ for the
  mean-variance trend's fitted value at that gene's mean (defined in
  `log.mean.component.of.variance` below), dispersion in log space is
  simply the log of their ratio:

  $$\log(\text{dispersion}) = \log(V) - \log(E) = \log(V / E)$$

  Across all genes in a sample, memento fits
  $\log(V) \sim \text{poly}(\log(\hat{\mu}_g), \text{degree}=2)$
  (`_fit_mv_regressor`) to obtain $\log(E)$ at each gene's mean, then computes
  `log.dispersion` $= \log(V) - \log(E)$ and `exp(log.dispersion)` $= V/E$
  (`_residual_variance`) -- i.e. dispersion is a RATIO relative to that
  fitted trend, not a variance component in the same additive units as the
  gene's raw variance. `log.dispersion = 0` means "exactly on-trend" for a
  gene at that mean (the linear ratio would be 1); call this value `r` below.
- **`log.mean.component.of.variance`** -- $\log(E)$, the mean-variance
  trend's fitted value at gene $g$'s mean expression ($E =
  \exp(\text{fitted\_trend}(\log(\hat{\mu}_g)))$, equivalently the gene's
  raw variance divided by `exp(log.dispersion)`, i.e. $E = V / \exp(r)$).
  This is "what variance this gene would have if it behaved exactly like a
  typical gene at this expression level" -- the mean-driven baseline that
  the gene's actual variance gets compared against to define dispersion.
- **`log.total.variability`** -- $\log(V)$, where $V = \hat{\sigma}_g^2$ is
  memento's raw 2nd moment (`Var[X]` of its internal
  capture-efficiency-corrected relative-expression estimate) -- computed
  inside `memento.compute_1d_moments()` but discarded by its normal API
  (`get_1d_moments()`, which only returns log(mean) and log(residual
  variance)) -- pulled directly out of
  `adata.uns['memento']['1d_moments'][group][1]` before it's dropped (see
  `memento_variance_decomposition.py`'s docstring for the full derivation).
  Satisfies the exact identity $\log(V) = \log(E) + r$, i.e.
  `log.total.variability = log.mean.component.of.variance + log.dispersion`
  (verified to floating-point precision, ~1e-15). Note that being on the
  same *scale* (all natural log)
  doesn't put them in the same *range*: `log.total.variability` and
  `log.mean.component.of.variance` both run roughly -21 to -11 here (since the
  underlying variances are tiny on memento's relative-abundance scale),
  while `log.dispersion` runs roughly -2 to +4 (since it's a ratio centered
  at 0, not a raw variance) -- that gap is a real feature of the data (the
  large negative offset shared by `V` and `E`), not an artifact of logging
  them.
- **`percent.variability.that.is.dispersion`** -- a point-wise metric,
  bounded 0-1 (a fraction, not a `*100` percentage), for how much of an
  individual gene's variance is accounted for by dispersion rather than by
  the mean-variance trend. Taking the absolute value of the log residual
  `r = log.dispersion` before exponentiating gives a formula that is
  bounded on *both* sides by construction:

  `percent.variability.that.is.dispersion = 1 - exp(-|r|)`

  This handles both directions of the residual symmetrically:
    - **`r > 0` (overdispersion):** simplifies to `(V - E) / V` -- the
      exact fraction of the gene's own *observed* variance that is
      "excess" dispersion above the mean trend.
    - **`r < 0` (underdispersion):** simplifies to `(E - V) / E` -- the
      fraction of the *expected* variance that the gene failed to reach.
    - **`r = 0`:** the gene sits exactly on the mean-variance trend -- 0
      ("perfect fit," no dispersion contribution either way).

  Example: `r = 0.223` (mild overdispersion) gives `1 - e^{-0.223} ~= 0.20`
  ("the dispersion excess makes up 20% of the observed variance"); `r =
  -0.223` (an equally-sized deviation in the other direction) gives the
  *same* 0.20 ("the observed variance is 20% deflated relative to the
  trend line") -- the magnitude of the deviation is treated identically
  regardless of direction, even though `r > 0` and `r < 0` are dividing by
  different denominators (`V` vs. `E`) under the hood.
- **`percent.variability.that.is.mean`** -- the direct complement,
  `exp(-|r|) = 1 - percent.variability.that.is.dispersion`, using the same
  `r = log.dispersion` value. A gene sitting exactly on-trend (`r = 0`)
  reads 1 here (all of its variance is consistent with what mean alone
  predicts); as dispersion becomes extreme in either direction, this falls
  toward 0. The two `percent.variability.*` columns always sum to exactly
  1 by construction (verified to floating-point precision, ~1e-14).

**Not accounted for here:** the production DD test
(`diploid_dd_limma_ct.R`) runs `limma::normalizeQuantiles()` on the
per-celltype dispersion matrix (across donor columns) before testing for a
species effect -- i.e. the `dd.lfsr`/`dd.logFC` significance calls operate
on quantile-normalized dispersion, not the raw `exp(log.dispersion)` values
used throughout this section. That's a deliberate choice: this section
characterises memento's dispersion *estimator* itself (is it mean-corrected,
how much of a gene's variance does it capture), which is a different
question from what the downstream cross-donor significance test sees after
its own normalization. Don't use these per-gene percentages to explain a
specific `dd.lfsr` call without re-deriving them on the quantile-normalized
values first.


```{r tv_load}
tv_data_path <- '/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q0.2/variance_decomposition/'

tv_summary <- read.csv(paste0(tv_data_path, "all_celltypes_variance_summary.csv"))
tv_genes   <- read.csv(paste0(tv_data_path, "all_celltypes_variance_genes.csv"))

tv_genes_cm <- tv_genes |>
  dplyr::filter(celltype == "Cardiomyocytes", ploidy == "diploid",
                is.finite(log.dispersion)) |>
  dplyr::mutate(panel = paste0(species, " - ", donor))

tv_summary_cm <- tv_summary |>
  dplyr::filter(celltype == "Cardiomyocytes", ploidy == "diploid") |>
  dplyr::mutate(panel = paste0(species, " - ", donor))
```

```{r}
tv_genes_cm
tv_summary_cm
```

# 2. Mean Dependence

Does the residual variance estimate still correlate with mean expression?
Memento corrects for mean internally, so we expect little residual association.
Each point is one gene × individual × cell type; no expression or gene
filtering applied here.

```{r mean_dependence_raw, fig.width = 12, fig.height = 16}
ggplot(df, aes(x = mean_expr, y = res_var)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.25, size = 0.6, colour = "grey30") +
  facet_grid(celltype ~ individual, labeller = labeller(celltype = ct_label)) +
  labs(x = "Log mean expression", y = "Residual variance") +
  theme_diag
```

---

# 3. LOESS Mean Dependence After Expression Cutoff (log mean ≤ -6)

Repeat the mean-dependence check above, restricted to genes that pass the
`-6` log mean expression cutoff (genes with log mean > -6 excluded) **and**
with mitochondrial/ribosomal genes removed (`MT-`, `MRPL`/`MRPS`, `RPL`/`RPS`),
to check whether residual mean dependence stays flat once high-expression
genes above the cutoff and these housekeeping genes are removed.

Full cell type × individual grid (18 panels), showing every retained gene as a
point (not just the smoothed line/ribbon), so raw scatter and per-panel LOESS
fit can be inspected together. Axes are left free per panel (not forced to the
tiny fitted-value range) since raw per-gene residual variance is far noisier
than the smoothed curve.

```{r loess_filtered_setup}
raw_ind_df_filt <- df |>
  filter(mean_expr <= CUTOFF, !is_mito_ribo(gene))

# Per-panel LOESS fit (with 95% CI), used as the smoothed overlay in the plot
# below. Panels with too few genes to fit reliably (n < 10) are skipped.
loess_ind_df_filt <- raw_ind_df_filt |>
  group_by(celltype, individual) |>
  group_modify(function(sub, ...) {
    if (nrow(sub) < 10) return(tibble())
    fit  <- loess(res_var ~ mean_expr, data = sub, span = 0.5)
    grid <- tibble(mean_expr = seq(min(sub$mean_expr), max(sub$mean_expr), length.out = 100))
    pred <- predict(fit, newdata = grid, se = TRUE)
    grid |> mutate(fitted = pred$fit,
                    lower  = pred$fit - 1.96 * pred$se.fit,
                    upper  = pred$fit + 1.96 * pred$se.fit)
  }) |>
  ungroup()
```

```{r loess_per_individual_points_filtered, fig.width = 12, fig.height = 16}
ggplot(raw_ind_df_filt, aes(x = mean_expr, y = res_var)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.25, size = 0.6, colour = "grey30") +
  geom_ribbon(data = loess_ind_df_filt,
              aes(x = mean_expr, ymin = lower, ymax = upper, y = NULL),
              inherit.aes = FALSE, alpha = 0.3, fill = "#4575b4") +
  geom_line(data = loess_ind_df_filt, aes(x = mean_expr, y = fitted),
            inherit.aes = FALSE, colour = "#4575b4", linewidth = 0.9) +
  facet_grid(celltype ~ individual, labeller = labeller(celltype = ct_label)) +
  labs(x = "Log mean expression", y = "Residual variance",
       subtitle = paste0("Genes with log mean ≤ ", CUTOFF,
                          " and mito/ribo genes removed; each point is one gene; ",
                          "LOESS ± 95% CI per cell type × individual")) +
  theme_diag
```

---

## 3b. Same Diagnostic, Mechanism Project (Human/Chimp Diploid + Allotetraploid)

**Note:** unlike every other section in this document, this subsection uses a
**different dataset** entirely -- the Gilad lab "mechanism" project's
human/chimp diploid + allotetraploid `memento` output (not the cHDC cardiac
pilot data used everywhere else here). It repeats the same mean-dependence
LOESS diagnostic (mito/ribo genes removed, log mean ≤ -6) for:

- **Diploid human** and **diploid chimp** -- full cell type × donor grid
  (7 donors per species, ~20 cell types), analogous to the cardiac
  per-individual grid above. Log mean ≤ -6 (same cutoff as the rest of this
  document), and donor × cell-type groups with fewer than `MIN_CELLS` = 100
  cells are excluded (see below).
- **Tetraploid human** and **tetraploid chimp** -- a single pooled line per
  cell type, restricted to the 20 cell types that overlap with diploid
  (tetraploid has one extra, `Early_mesoderm_cells`, with no diploid
  counterpart, so it's dropped here). The `memento_tetraploid_dd_ct` pipeline
  groups tetraploid cells by allele-of-origin (`species`) × cell type only,
  with no donor/replicate level (there is only one tetraploid line in this
  dataset), so a per-individual breakdown isn't possible here the way it is
  for diploid. Uses the same log mean ≤ -6 cutoff as diploid.

Source data is `mt_rb_mean_filtered_q_thresholds/q0.2/memento/` from the
mechanism project. We use three filters throughout: mitochondrial/ribosomal
gene removal, a log-mean-expression cutoff of -6, and a minimum of 100
cells per donor x cell-type group. All three are already applied upstream
by the pipeline that produced this data; they are reapplied explicitly
below so the filtering this section depends on is visible directly in this
document.

```{r mech_loess_setup}
MECH_DIR <- "/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q0.2/memento"

ct_label_us <- function(x) gsub("_", " ", x)

# Wrapped variant used only for the mean-independence loess grids below --
# long names (e.g. "Bronchiolar and alveolar epithelial cells") otherwise
# overflow the facet panel/strip and get cropped out of the saved plot.
ct_label_us_wrapped <- function(x, wrap_width = 18) {
  stringr::str_wrap(gsub("_", " ", x), width = wrap_width)
}

mech_celltypes_for <- function(dir, prefix) {
  list.files(dir, pattern = paste0("^", prefix, "_.*\\.csv$")) |>
    sub(paste0("^", prefix, "_"), "", x = _) |>
    sub("\\.csv$", "", x = _)
}

mech_read_ct <- function(dir, prefix, ct, ploidy) {
  mean_f <- file.path(dir, paste0(prefix, "_mean_", ct, ".csv"))
  var_f  <- file.path(dir, paste0(prefix, "_var_",  ct, ".csv"))
  if (!file.exists(mean_f) || !file.exists(var_f)) return(NULL)

  mean_long <- read.csv(mean_f, check.names = FALSE) |>
    pivot_longer(-gene, names_to = "sample", values_to = "mean_expr")
  var_long <- read.csv(var_f, check.names = FALSE) |>
    pivot_longer(-gene, names_to = "sample", values_to = "res_var")

  inner_join(mean_long, var_long, by = c("gene", "sample")) |>
    mutate(ploidy   = ploidy,
           celltype = ct,
           species  = sub("^sg\\^([^^]+)\\^.*$", "\\1", sample),
           donor    = sub("^sg\\^[^^]+\\^", "", sample)) |>
    filter(!is.na(mean_expr), !is.na(res_var))
}

mech_dip_cts_all <- mech_celltypes_for(MECH_DIR, "diploid_memento_mean")
mech_tet_cts_all <- mech_celltypes_for(MECH_DIR, "tetraploid_memento_mean")

# Tetraploid has one extra cell type (Early_mesoderm_cells) with no diploid
# counterpart -- restrict to the 20 that overlap so every tetraploid panel
# here has a matching diploid comparison above.
mech_dip_cts <- mech_dip_cts_all
mech_tet_cts <- intersect(mech_tet_cts_all, mech_dip_cts_all)
cat("Tetraploid cell types dropped (no diploid counterpart):",
    paste(setdiff(mech_tet_cts_all, mech_dip_cts_all), collapse = ", "), "\n")

# Reuses CUTOFF (-6) and is_mito_ribo(), defined above -- applied explicitly
# here (in addition to being applied upstream by the pipeline) for clarity.
mech_dip_df <- map_dfr(mech_dip_cts, ~ mech_read_ct(MECH_DIR, "diploid_memento", .x, "Diploid")) |>
  filter(mean_expr <= CUTOFF, !is_mito_ribo(gene))
mech_tet_df <- map_dfr(mech_tet_cts, ~ mech_read_ct(MECH_DIR, "tetraploid_memento", .x, "Tetraploid")) |>
  filter(mean_expr <= CUTOFF, !is_mito_ribo(gene))

# Exclude diploid donor x cell-type groups with fewer than MIN_CELLS cells,
# the third of our three filters (see below for why). Computed directly from
# the raw diploid metadata, matched on (species, donor, celltype); celltype
# names there use spaces where the memento output uses underscores, so
# normalise before joining.
MIN_CELLS <- 100

diploid_meta_for_counts <- read.csv(
  "/project/gilad/awchen55/mechanism/data/cell_annotations/diploid_metadata.csv"
) |>
  mutate(celltype = gsub(" ", "_", celltype))

low_count_combos <- diploid_meta_for_counts |>
  filter(celltype %in% mech_dip_cts) |>
  count(species, donor_id, celltype, name = "n_cells") |>
  filter(n_cells < MIN_CELLS) |>
  rename(donor = donor_id)

cat("Diploid donor x cell-type combos excluded (< ", MIN_CELLS, " cells): ",
    nrow(low_count_combos), "\n", sep = "")

mech_dip_df <- mech_dip_df |>
  anti_join(low_count_combos, by = c("species", "donor", "celltype"))

cat("Diploid rows:", nrow(mech_dip_df), "| cell types:", length(mech_dip_cts), "\n")
cat("Diploid donors -- chimp:", length(unique(mech_dip_df$donor[mech_dip_df$species == "chimp"])),
    "human:", length(unique(mech_dip_df$donor[mech_dip_df$species == "human"])), "\n")
cat("Tetraploid rows:", nrow(mech_tet_df), "| cell types:", length(mech_tet_cts), "\n")
```

```{r mech_loess_diploid_grid, fig.width = 15, fig.height = 32}
MECH_OUTDIR <- "/project/gilad/awchen55/mechanism/results/dispersion_paper/supplement/mean_independence_loess/"

mech_make_grid_plot <- function(sp) {
  sub <- filter(mech_dip_df, species == sp)
  ggplot(sub, aes(x = mean_expr, y = res_var)) +
    geom_point(alpha = 0.15, size = 0.3, colour = "grey30") +
    geom_smooth(method = "loess", span = 0.5, colour = "#d73027", linewidth = 0.5, se = TRUE) +
    facet_grid(celltype ~ donor, labeller = labeller(celltype = ct_label_us_wrapped)) +
    labs(x = "Mean", y = "Dispersion") +
    theme_diag +
    theme(strip.text.y = element_text(angle = 0, size = 13), strip.text.x = element_text(size = 7),
          axis.title = element_text(size = 16))
}

for (sp in c("human", "chimp")) {
  p <- mech_make_grid_plot(sp)
  print(p)
  n_ct <- length(unique(filter(mech_dip_df, species == sp)$celltype))
  n_donor <- length(unique(filter(mech_dip_df, species == sp)$donor))
  # ggsave(file.path(MECH_OUTDIR, paste0(sp, "_diploid_mean_independence_loess.png")),
  #        plot = p, width = max(10, n_donor * 2.2), height = max(10, n_ct * 1.6), limitsize = FALSE, dpi = 150)
}
```

Without the `MIN_CELLS` filter, individual donors with very few cells in a
given cell type (human `NA18913`; chimp `C3651`, `C4955`, `C40280`) show a
fanned, sharply-descending cluster of points in several neural-lineage cell
types (Ciliated epithelial cells, ENS glia, Glioblast, Neuroblast, Neuron,
Retinal neurons) -- the same low-cell-count numerical artifact documented
for chimp donors in `q_threshold_sensitivity.Rmd` Section 8b (memento's
moment-based dispersion estimator collapsing to near-identical,
artificially very-negative values when a donor has too few cells in a cell
type). Every one of those donor × cell-type combos has well under 100 cells
(e.g. chimp `C3651` Glioblast = 4 cells; human `NA18913` Neuron = 29 cells),
confirming the artifact affects human donors as well as chimp. The
100-cell exclusion above removes all of these from the grids.

```{r mech_loess_tetraploid, fig.width = 13, fig.height = 10}
mech_make_tet_plot <- function(sp) {
  sub <- filter(mech_tet_df, species == sp)
  ggplot(sub, aes(x = mean_expr, y = res_var)) +
    geom_point(alpha = 0.15, size = 0.4) +
    geom_smooth(method = "loess", span = 0.5, colour = "red", linewidth = 0.7, se = TRUE) +
    facet_wrap(~ celltype, ncol = 5, labeller = labeller(celltype = ct_label_us_wrapped)) +
    labs(x = "Mean", y = "Dispersion") +
    theme_diag +
    theme(strip.text = element_text(size = 10), axis.title = element_text(size = 16))
}

for (sp in c("human", "chimp")) {
  p <- mech_make_tet_plot(sp)
  print(p)
  # ggsave(file.path(MECH_OUTDIR, paste0(sp, "_allotetraploid_mean_independence_loess.png")),
  #        plot = p, width = 13, height = 10, dpi = 150)
}
```


# 4. Reproducibility Across Individuals

Pairwise Spearman correlation of residual variance between individuals
within each cell type. High concordance supports reliable estimation.

```{r repro_cor}
repro_df <- map_dfr(CELL_TYPES, function(ct) {
  sub_wide <- res_var_wide[, grepl(ct, colnames(res_var_wide)), drop = FALSE]
  if (ncol(sub_wide) < 2) return(NULL)
  inds <- INDIVIDUALS[INDIVIDUALS %in% sub("_.*", "", colnames(sub_wide))]
  pairs <- combn(inds, 2, simplify = FALSE)
  map_dfr(pairs, function(p) {
    col1 <- paste0(p[1], "_", ct)
    col2 <- paste0(p[2], "_", ct)
    if (!all(c(col1, col2) %in% colnames(sub_wide))) return(NULL)
    ok <- !is.na(sub_wide[[col1]]) & !is.na(sub_wide[[col2]])
    data.frame(
      celltype = ct,
      pair     = paste(p, collapse = " vs "),
      spearman = cor(sub_wide[[col1]][ok], sub_wide[[col2]][ok], method = "spearman"),
      n_genes  = sum(ok)
    )
  })
})

repro_df |> arrange(celltype, pair)

pw_indv <- ggplot(repro_df, aes(x = reorder(celltype, spearman), y = spearman, colour = pair)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_discrete(labels = ct_label) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = expression(paste("Spearman ", rho, " (pairwise across individuals)")),
       subtitle = "",
       colour = "Individual pair") +
  theme_diag +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.position = "top")
pw_indv
#ggplot2::ggsave(filename = paste0(results_path, "reproducibility_indv.pdf"), plot = pw_indv,  device = "pdf", width = 7, height = 5, units = "in")
```

```{r repro_scatter, fig.width = 11, fig.height = 9}
# Scatter plots for each cell type: first pair of individuals
plots <- map(CELL_TYPES, function(ct) {
  col1 <- paste0("NA19099_", ct)
  col2 <- paste0("NA19093_", ct)
  if (!all(c(col1, col2) %in% colnames(res_var_wide))) return(NULL)
  df_plot <- data.frame(x = res_var_wide[[col1]], y = res_var_wide[[col2]]) |>
    filter(!is.na(x), !is.na(y))
  rho <- cor(df_plot$x, df_plot$y, method = "spearman") |> round(3)
  ggplot(df_plot, aes(x = x, y = y)) +
    geom_point(alpha = 0.15, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, colour = "#d73027", linetype = "dashed") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
             label = paste0("rho == ", rho), size = 3.5, parse = TRUE) +
    labs(title = gsub("_", " ", ct),
         x = "NA19099 residual variance", y = "NA19093 residual variance") +
    theme_diag
})

gridExtra::grid.arrange(grobs = Filter(Negate(is.null), plots), ncol = 2)
```



# 5. Genes Excluded by Expression Threshold (log mean > -6)

How many genes per cell type are above log mean expression of -6 and would
be removed by applying this filter.

```{r expr_cutoff_table}
# CUTOFF and df_ct_avg were defined earlier, right after the data was loaded.
cutoff_df <- df_ct_avg |>
  group_by(celltype) |>
  summarise(
    n_total    = n(),
    n_excluded = sum(mean_expr > CUTOFF, na.rm = TRUE),
    n_retained = sum(mean_expr <= CUTOFF, na.rm = TRUE),
    pct_excluded = round(100 * n_excluded / n_total, 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_excluded))

cutoff_df
```




## 5b. DD Genes Removed at Stricter Expression Cutoffs (Mechanism Project)

How many of the mechanism project's current DD (differentially dispersed)
genes would be lost if the log-mean-expression cutoff were tightened from
the pipeline's current `-6` to `-6.5` or `-7`? Uses the same mechanism
dataset as Section 3b.

**DD definition:** `diploid_lfsr < 0.05` AND `allotetraploid_lfsr < 0.05`
AND same-sign logFC in both -- confirmed to match the master table's
pre-computed `dd.recapitulated.same.sign.{celltype}` flag exactly (cross-checked
against a manual recomputation for 3 cell types: Cardiomyocytes 83/83,
iPSCs 334/334, Neuron 91/91 genes, zero mismatches), so that flag is used
directly rather than re-derived.

**Removal rule:** matches the pipeline's own per-sample logic
(`memento_diploid.py`: a gene is retained only if **all** of its diploid
species × donor samples have mean expression ≤ the cutoff) -- i.e. a
currently-DD gene is counted as "removed" at a given cutoff if **any** of
its diploid samples exceeds that cutoff. Applied to the current DD gene set
per cell type (from `DD_DE_interspecies_results.csv`), using diploid mean
expression from `mt_rb_mean_filtered_q_thresholds/q0.2/memento/` (same
source as Section 3b). As expected, since `-6` is already the pipeline's
configured cutoff, 0 genes are removed at that value (see table below); the
`-6.5` and `-7` counts show how many additional DD genes a stricter cutoff
would remove on top of that.

```{r dd_cutoff_sweep_setup}
DD_PATH <- "/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q0.2/DD_DE_interspecies_results.csv"
DD_SWEEP_CUTOFFS <- c(-6, -6.5, -7)

dd_master <- read.csv(DD_PATH, check.names = FALSE)

recap_cols <- grep("^dd\\.recapitulated\\.same\\.sign\\.", colnames(dd_master), value = TRUE)
recap_cols <- setdiff(recap_cols, "dd.recapitulated.same.sign.n.celltypes")
dd_sweep_celltypes <- sub("^dd\\.recapitulated\\.same\\.sign\\.", "", recap_cols)

dd_genes_by_ct <- map(dd_sweep_celltypes, function(ct_dot) {
  flag <- dd_master[[paste0("dd.recapitulated.same.sign.", ct_dot)]]
  dd_master$gene[!is.na(flag) & flag == 1]
})
names(dd_genes_by_ct) <- dd_sweep_celltypes

cat("Cell types with DD calls:", length(dd_sweep_celltypes), "\n")
cat("Total unique DD genes across all cell types:", length(unique(unlist(dd_genes_by_ct))), "\n")
```

```{r dd_cutoff_sweep_compute}
# Worst-case (max) log mean expression across all diploid samples per gene,
# per cell type -- mirrors memento_diploid.py's own retention rule.
max_mean_expr_for_ct <- function(ct_dot) {
  ct_us <- gsub("\\.", "_", ct_dot)
  f <- file.path(MECH_DIR, paste0("diploid_memento_mean_", ct_us, ".csv"))
  if (!file.exists(f)) return(NULL)
  wide <- read.csv(f, check.names = FALSE)
  sample_cols <- setdiff(colnames(wide), "gene")
  tibble(gene = wide$gene, max_mean_expr = apply(wide[sample_cols], 1, max, na.rm = TRUE))
}

mean_by_ct <- map(dd_sweep_celltypes, max_mean_expr_for_ct)
names(mean_by_ct) <- dd_sweep_celltypes

dd_cutoff_sweep <- map_dfr(dd_sweep_celltypes, function(ct_dot) {
  dd_genes <- dd_genes_by_ct[[ct_dot]]
  mean_tbl <- mean_by_ct[[ct_dot]]
  if (is.null(mean_tbl) || length(dd_genes) == 0) return(NULL)
  sub <- mean_tbl |> filter(gene %in% dd_genes)
  map_dfr(DD_SWEEP_CUTOFFS, function(co) {
    tibble(celltype = ct_dot, cutoff = co, n_dd_genes = length(dd_genes),
           n_removed = sum(sub$max_mean_expr > co, na.rm = TRUE))
  })
}) |>
  mutate(pct_removed = round(100 * n_removed / n_dd_genes, 1))

dd_cutoff_sweep |> arrange(cutoff, desc(n_removed)) |> as.data.frame()
```

```{r dd_cutoff_sweep_unique_total}
dd_unique_removed <- map_dfr(DD_SWEEP_CUTOFFS, function(co) {
  removed_union <- map(dd_sweep_celltypes, function(ct_dot) {
    dd_genes <- dd_genes_by_ct[[ct_dot]]
    mean_tbl <- mean_by_ct[[ct_dot]]
    if (is.null(mean_tbl) || length(dd_genes) == 0) return(character(0))
    (mean_tbl |> filter(gene %in% dd_genes, max_mean_expr > co))$gene
  }) |> unlist() |> unique()
  tibble(cutoff = co, n_unique_removed = length(removed_union))
})

n_unique_dd_total <- length(unique(unlist(dd_genes_by_ct)))
dd_unique_removed <- dd_unique_removed |>
  mutate(n_unique_dd_total = n_unique_dd_total,
         pct_unique_removed = round(100 * n_unique_removed / n_unique_dd_total, 1))

dd_unique_removed |> as.data.frame()
```

```{r dd_cutoff_sweep_plot, fig.width = 11, fig.height = 6}
ggplot(dd_cutoff_sweep, aes(x = reorder(celltype, -n_removed), y = n_removed, fill = factor(cutoff))) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("-6" = "#4575b4", "-6.5" = "#fdae61", "-7" = "#d73027"),
                     name = "Log mean\ncutoff") +
  scale_x_discrete(labels = ct_label_us) +
  labs(x = NULL, y = "DD genes removed",
       subtitle = "Current DD genes (recapitulated, same-sign) with any diploid sample above cutoff") +
  theme_diag +
  theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "top")
```



# 6. Downsampling Reliability Sweep (Cardiomyocytes)

The sections above look at count-depth dependence using the real
(uncontrolled) cell counts across existing samples, which is confounded: a
sample with fewer cells might also just have genuinely different biology.
Here we run a controlled experiment instead: take Cardiomyocytes (the largest cell type,
30,423 cells pooled across the 3 individuals), downsample each individual's
cells to a series of fixed depths (100-5,000 cells, well below and up to the
main project's ~1,000-cell inclusion threshold), rerun memento with 30
independent replicate subsamples per depth, and compare each replicate's
per-gene dispersion estimate back to that individual's full-data (no
downsampling) reference. This isolates the effect of cell count from
between-sample biological differences.

```{r ds_reliability_load}
DOWNSAMPLE_DIR <- "/project/gilad/awchen55/differentialDispersion/analysis/dispersion_diagnostics/data/downsample_reliability_cardiomyocytes"
DS_DEPTHS      <- c("100", "250", "500", "1000", "2500", "5000")  # finite depths; "full" is the reference
DS_REPLICATES  <- as.character(0:29)

read_ds_dispersion <- function(depth_label, rep_label, dir = DOWNSAMPLE_DIR) {
  f <- file.path(dir, paste0("depth_", depth_label), paste0("dispersion_rep", rep_label, ".csv"))
  if (!file.exists(f)) return(NULL)
  read.csv(f, check.names = FALSE) |>
    pivot_longer(-gene, names_to = "sample", values_to = "dispersion") |>
    mutate(individual = sub("_.*", "", sample),
           depth      = depth_label,
           replicate  = rep_label) |>
    select(gene, individual, depth, replicate, dispersion)
}

ds_full <- read_ds_dispersion("full", "0") |>
  select(gene, individual, dispersion_full = dispersion)

ds_long <- purrr::map_dfr(DS_DEPTHS, function(d) {
  purrr::map_dfr(DS_REPLICATES, function(r) read_ds_dispersion(d, r))
})

ds_joined <- ds_long |>
  inner_join(ds_full, by = c("gene", "individual")) |>
  filter(!is.na(dispersion), !is.na(dispersion_full))

cat("Full-data reference (per individual, no downsampling):", nrow(ds_full), "gene x individual rows\n")
cat("Downsampled rows loaded:", nrow(ds_long), "across", length(DS_DEPTHS), "depths x",
    length(DS_REPLICATES), "replicates\n")
cat("Joined to reference:", nrow(ds_joined), "rows\n")
```

## Reliability vs. depth

```{r ds_reliability_cor, fig.width = 9, fig.height = 5.5}
ds_reliability <- ds_joined |>
  group_by(depth, replicate, individual) |>
  summarise(spearman = cor(dispersion, dispersion_full, method = "spearman", use = "complete.obs"),
            n_genes = n(), .groups = "drop") |>
  mutate(depth_num = as.numeric(depth))

ds_reliability_summary <- ds_reliability |>
  group_by(depth_num, individual) |>
  summarise(mean_spearman = mean(spearman), sd_spearman = sd(spearman), .groups = "drop")

downsamp_plot <- ggplot() +
  geom_point(data = ds_reliability, aes(x = depth_num, y = spearman, colour = individual),
             alpha = 0.35, size = 1.8) +
  geom_ribbon(data = ds_reliability_summary,
              aes(x = depth_num, ymin = mean_spearman - sd_spearman, ymax = mean_spearman + sd_spearman,
                  fill = individual), alpha = 0.15, colour = NA) +
  geom_line(data = ds_reliability_summary, aes(x = depth_num, y = mean_spearman, colour = individual),
            linewidth = 1) +
  ggrepel::geom_text_repel(data = ds_reliability_summary,
                            aes(x = depth_num, y = mean_spearman, colour = individual,
                                label = round(mean_spearman, 2)),
                            size = 3, show.legend = FALSE, seed = 1) +
  scale_x_log10(breaks = as.numeric(DS_DEPTHS)) +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Downsample depth (cells per individual, log scale)",
       y = expression(paste("Spearman ", rho, " vs. full-data dispersion")),
       subtitle = "",
       colour = NULL, fill = NULL) +
  theme_diag +
  theme(legend.position = "top",
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12))
downsamp_plot
# ggplot2::ggsave(filename = paste0(results_path, "downsampling_test.pdf"), plot = downsamp_plot,  device = "pdf", width = 7, height = 5, units = "in")
```





