---
title: "DD Interspecies Analysis"
output: html_document
date: "2026-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Results and Packages

```{r}
library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(Matrix)
library(purrr)
library(ggpubr)
library(stringr)
library(scales)
```

We load DD results.
```{r}
path_data <- '/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q0.2/'
results_path <- '/project/gilad/awchen55/mechanism/results/dispersion_paper/'

dd_results <- read.csv(paste0(path_data,"DD_DE_interspecies_results.csv"), check.names = FALSE)
head(dd_results)

round_up_signif <- function(p, digits = 1) {
  
  if (p < 2e-16) {
    return("2 %*% 10^{-16}")
  }
  
  # round up
  if (p != 0) {
    expo  <- floor(log10(p))
    scale <- 10^(expo - digits + 1)
    p <- ceiling(p / scale) * scale
  }
  
  # force decimal formatting
  if (p >= 1e-4) {
    dec <- min(max(0, -floor(log10(p))), 4)
    return(formatC(p, format = "f", digits = dec))
  }
  
  # scientific notation
  expo <- floor(log10(p))
  base <- round(p / 10^expo, digits)
  
  paste0(base, " %*% 10^{", expo, "}")
}
```

## Summary Tables

### Number of DD Genes

```{r}

# vector of cell types

celltypes2 <- c("Bronchiolar_and_alveolar_epithelial_cells",
    "Cardiac_mesoderm_cells",
    "Cardiomyocytes",
    "Ciliated_epithelial_cells",
    "ENS_glia",
    "Epicardial_fat_cells",
    "Glioblast",
    "Goblet_cells",
    "Hepatoblasts",
    "Intestinal_epithelial_cells",
    "Metanephric_cells",
    "Neuroblast",
    "Neuron",
    "Neuronal_IPC",
    "Radial_glia",
    "Retinal_neurons",
    "Squamous_epithelial_cells",
    "Stromal_cells",
    "Vascular_endothelial_cells",
    "iPSCs")

celltypes <- gsub("_", ".", celltypes2)
celltypes3 <- celltypes
celltypes_display <- gsub("\\.", " ", celltypes)

summary_table <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col  <- paste0("dd.lfsr.diploid.", ct)
  recap_col <- paste0("dd.recapitulated.same.sign.", ct)

  dd_flag <- dd_results[[lfsr_col]] < 0.05

  dd_total <- sum(dd_flag, na.rm = TRUE)
  recap_total <- sum(dd_flag & dd_results[[recap_col]] == 1, na.rm = TRUE)

  tibble::tibble(
    CellType = gsub("\\.", " ", ct),
    `DD Genes` = dd_total,
    Recapitulated = recap_total,
    `% Recapitulated` = recap_total / dd_total
  )
})

summary_table
```


### Plot of DD Genes

Number of DD genes (diploid LFSR < 0.05) per cell type, ordered by
descending total. Each bar is split into genes that are also recapitulated
in the allotetraploid (LFSR < 0.05 in both diploid and allotetraploid, same
effect-size sign; light blue) versus genes that are diploid-DD only (grey).

```{r}

celltypes_space <- c(
    "Radial glia",
  "Cardiomyocytes",
  "Neuroblast",
  "iPSCs",
  "Retinal neurons",
  "Stromal cells",
  "Vascular endothelial cells",
  "Metanephric cells",
  "Squamous epithelial cells",
  "ENS glia",
  "Goblet cells",
  "Bronchiolar and alveolar epithelial cells",
  "Cardiac mesoderm cells",
  "Neuronal IPC",
  "Hepatoblasts",
  "Glioblast",
  "Intestinal epithelial cells",
  "Neuron",
  "Epicardial fat cells",
  "Ciliated epithelial cells"
)

celltypes_dot <- gsub(" ", ".", celltypes_space)

df_totals <- purrr::map_dfr(seq_along(celltypes_space), function(i) {

  ct_label <- celltypes_space[i]
  ct_col   <- celltypes_dot[i]

  diploid_lfsr_col <- paste0("dd.lfsr.diploid.", ct_col)
  tetra_lfsr_col   <- paste0("dd.lfsr.allotetraploid.", ct_col)

  diploid_logfc_col <- paste0("dd.logFC.diploid.", ct_col)
  tetra_logfc_col   <- paste0("dd.logFC.allotetraploid.", ct_col)

  dd_flag <- !is.na(dd_results[[diploid_lfsr_col]]) &
    dd_results[[diploid_lfsr_col]] < 0.05

  recap_flag <- dd_flag &
    !is.na(dd_results[[tetra_lfsr_col]]) &
    dd_results[[tetra_lfsr_col]] < 0.05 &
    !is.na(dd_results[[diploid_logfc_col]]) &
    !is.na(dd_results[[tetra_logfc_col]]) &
    sign(dd_results[[diploid_logfc_col]]) == sign(dd_results[[tetra_logfc_col]])

  n_dd    <- sum(dd_flag, na.rm = TRUE)
  n_recap <- sum(recap_flag, na.rm = TRUE)
  n_not   <- n_dd - n_recap

  tibble::tibble(
    CellType = ct_label,
    Not_Recapitulated = n_not,
    Recapitulated = n_recap
  )
}) %>%
  dplyr::mutate(
    CellType = factor(CellType, levels = celltypes_space)
  ) %>%
  tidyr::pivot_longer(
    cols = c(Not_Recapitulated, Recapitulated),
    names_to = "List",
    values_to = "DD_Genes"
  ) %>%
  dplyr::mutate(
    List = factor(List, levels = c("Not_Recapitulated", "Recapitulated"))
  )

dd_totals_by_ct <- df_totals %>%
  dplyr::group_by(CellType) %>%
  dplyr::summarise(total = sum(DD_Genes), .groups = "drop")

dd_gene_plot <- ggplot2::ggplot(
  df_totals,
  ggplot2::aes(x = CellType, y = DD_Genes, fill = List)
) +
  ggplot2::geom_col() +
  # NOTE: previously hardcoded to limits = c(0, 2400) via scale_y_continuous(),
  # which *drops* data outside the range rather than just zooming the view --
  # this silently truncated the "Not Recapitulated" segment (stacked on top)
  # for any cell type whose total DD gene count exceeded 2400 (Radial glia,
  # ENS glia, Neuronal IPC all do). coord_cartesian() zooms without dropping data.
  ggplot2::coord_cartesian(ylim = c(0, max(dd_totals_by_ct$total, na.rm = TRUE) * 1.05)) +
  ggplot2::scale_y_continuous(
    breaks = seq(0, 3000, by = 500),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    x = "",
    y = "# DD genes (LFSR < 0.05)",
    title = ""
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
  panel.grid = ggplot2::element_blank(),
  axis.line = ggplot2::element_line(color = "black", linewidth = 0.4),
  panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.8),
  axis.title = ggplot2::element_text(size = 14),
  axis.text.x = ggplot2::element_text(size = 12, angle = 45, hjust = 1),
  axis.text.y = ggplot2::element_text(size = 12),
  legend.position = c(0.99, 0.99),        # top-right inside
  legend.justification = c("right", "top"),
  legend.title = ggplot2::element_blank(),
  legend.background = ggplot2::element_rect(fill = NA, color = NA),
  legend.text = ggplot2::element_text(size = 12),
  legend.key.size = grid::unit(0.4, "cm")
) +
  ggplot2::scale_fill_manual(
    values = c(
      "Not_Recapitulated" = "gray87",
      "Recapitulated" = "lightskyblue"
    ),
    labels = c(
      "Not_Recapitulated" = "Not Recapitulated",
      "Recapitulated" = "Recapitulated"
    )
  )

dd_gene_plot
# ggplot2::ggsave(filename = paste0(results_path, "main/recapitulateed_dd_genes.pdf"), plot = dd_gene_plot,  device = "pdf", width = 8, height = 5, units = "in")
```


### Recapitulation rate

Load metadata from allotetraploid
```{r}
path <- '/project/gilad/awchen55/mechanism/data/cell_annotations/'
metadata_tetraploid <- read.csv(paste0(path,'tetraploid_metadata.csv'))
```

For each cell type, the recapitulation rate (fraction of diploid DD genes
that are also DD-significant in the allotetraploid with the same
effect-size sign) plotted alongside the number of allotetraploid cells
available for that cell type (labeled above each bar), to give a sense of
whether cell types with fewer cells also show lower recapitulation rates.

```{r}
recap_plot_df <- metadata_tetraploid %>%
  dplyr::filter(
    stringr::str_starts(cell_barcode, "human_"),
    celltype != "unassigned"
  ) %>%
  dplyr::mutate(
    CellType = gsub("_", " ", celltype)
  ) %>%
  dplyr::filter(CellType %in% celltypes_display) %>%
  dplyr::count(CellType, name = "n_cells") %>%
  dplyr::right_join(summary_table, by = "CellType") %>%
  dplyr::mutate(
    n_cells = tidyr::replace_na(n_cells, 0),
    CellType = factor(CellType, levels = CellType[order(-n_cells)])
  )

recap_plot <- recap_plot_df %>%
  ggplot2::ggplot(
    ggplot2::aes(x = CellType, y = `% Recapitulated`)
  ) +
  ggplot2::geom_col(fill = "#4DBBD5") +
  ggplot2::geom_text(
    ggplot2::aes(label = n_cells),
    vjust = -0.4,
    size = 4
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    expand = c(0, 0.02)
  ) +
  ggplot2::labs(
    x = "",
    y = "Recapitulation rate",
    title = ""
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.4),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.8),
    axis.title = ggplot2::element_text(size = 14),
    axis.text.x = ggplot2::element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 12)
  )

recap_plot
```

Recapitulation rate plotted directly against allotetraploid cell number
(rather than showing cell type on the x-axis as above), to see whether the
two are related in general across cell types. Labeled points are the
top/bottom 3 cell types by recapitulation rate plus the top 3 by cell
count, so both the extremes and the largest cell types are identifiable.

```{r}
# top/bottom 3 by recapitulation rate + top 3 by cell count, for ggrepel
# labeling in the scatter below.
label_df <- recap_plot_df %>%
  dplyr::filter(!is.na(`% Recapitulated`)) %>%
  dplyr::arrange(desc(`% Recapitulated`)) %>%
  dplyr::slice(c(1:3, (n()-2):n())) %>%
  dplyr::bind_rows(
    recap_plot_df %>%
      dplyr::arrange(desc(n_cells)) %>%
      dplyr::slice(1:3)
  ) %>%
  dplyr::distinct()

recap_scatter <- ggplot(recap_plot_df, aes(x = n_cells, y = `% Recapitulated`)) +
  geom_point(size = 2, color = "#4DBBD5") +

  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = CellType),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.3,
    point.padding = 0.2,
    force = 2,
    segment.color = "grey50"
  ) +

  scale_x_continuous(expand = ggplot2::expansion(mult = 0.03)) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = ggplot2::expansion(mult = 0.03)
  ) +

  labs(
    x = "# allotetraploid cells",
    y = "% recapitulated"
  ) +

  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
recap_scatter
#ggplot2::ggsave(filename = paste0(results_path, "supplement/recapitulation_rate/recapitulation_rate_scatter.pdf"), plot = recap_scatter,  device = "pdf", width = 7, height = 5, units = "in")
```

#### Recapitulation rate vs. cell number (downsampling within one cell type)

The scatter above compares % recapitulated across *different* cell types, which
confounds cell number with everything else that differs between cell types
(cell-type-specific biology, gene expression levels, etc). To isolate the
effect of statistical power alone, we downsample **Radial glia** -- the
largest allotetraploid cell type (47,437 hybrid cells) -- to smaller cell
counts (100, 1,000, 2,000, 5,000, 8,000, 10,000, 16,000, 24,000, 32,000,
40,000; 3 random-subsample replicates per size), rerun the production
memento DD test + MASH refit at each size (see
`downsample_recapitulation/submit_downsample_jobs.sh`), and recompute the
recapitulation rate against the same (unchanged) diploid DD gene set used
throughout this notebook. The full-data point reuses the already-computed
production result from `summary_table` above -- no rerun needed.

Replicate spread is noticeably wider at the smallest size (n=100: 0.7-1.1%)
than at n>=1,000 (typically <15% relative spread), consistent with sampling
noise on a small significant-gene count rather than a broken test (no NaNs
or out-of-range values at any size).

```{r}
downsample_dir <- "/project/gilad/awchen55/mechanism/analysis/dd_de_analysis/downsample_recapitulation/downsample_recapitulation_rate_radia_glia/"

downsample_files <- list.files(
  downsample_dir,
  pattern = "^recap_Radial_glia_n[0-9]+_rep[0-9]+\\.csv$",
  full.names = TRUE
)

downsample_df <- purrr::map_dfr(downsample_files, read.csv)

# full-data anchor point: already computed in summary_table, no rerun needed
full_data_point <- summary_table %>%
  dplyr::filter(CellType == "Radial glia") %>%
  dplyr::transmute(
    celltype = "Radial_glia",
    n_cells = max(recap_plot_df$n_cells[recap_plot_df$CellType == "Radial glia"]),
    replicate = 0L,
    n_dd_diploid = `DD Genes`,
    n_recap = Recapitulated,
    pct_recap = `% Recapitulated`
  )

downsample_df <- dplyr::bind_rows(downsample_df, full_data_point) %>%
  dplyr::arrange(n_cells)

# CI on the mean recapitulation rate at each cell count, computed from the
# actual spread across the (now 10) independent random-subsample replicates
# at that size -- this reflects real downsampling variability, unlike a
# binomial CI on a single point's proportion (which would be nonzero even
# for the full, non-downsampled anchor point, where there is no resampling
# and thus no variability to report -- exactly why that point gets no error
# bar below: n_reps = 1 there, so sd/CI are undefined).
downsample_summary <- downsample_df %>%
  dplyr::group_by(n_cells) %>%
  dplyr::summarise(
    mean_pct_recap = mean(pct_recap),
    sd_pct_recap   = sd(pct_recap),
    n_reps         = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    se       = sd_pct_recap / sqrt(n_reps),
    ci_lower = mean_pct_recap - stats::qt(0.975, n_reps - 1) * se,
    ci_upper = mean_pct_recap + stats::qt(0.975, n_reps - 1) * se
  )

downsample_df
```

```{r}
downsample_recap_plot <- ggplot2::ggplot(
  downsample_df,
  ggplot2::aes(x = n_cells, y = pct_recap)
) +
  ggplot2::geom_point(color = "grey60", size = 1, alpha = 0.7) +
  ggplot2::geom_line(
    data = downsample_summary,
    ggplot2::aes(x = n_cells, y = mean_pct_recap),
    color = "#4DBBD5",
    linewidth = 1
  ) +
  ggplot2::geom_point(
    data = downsample_summary,
    ggplot2::aes(x = n_cells, y = mean_pct_recap),
    color = "#4DBBD5",
    size = 2
  ) +
  ggplot2::scale_x_continuous(
    labels = scales::comma,
    expand = ggplot2::expansion(mult = 0.05)
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    expand = ggplot2::expansion(mult = 0.03)
  ) +
  ggplot2::labs(
    x = "# allotetraploid cells (Radial glia, downsampled)",
    y = "% recapitulated"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(color = "black"),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.8),
    axis.title = ggplot2::element_text(size = 14),
    axis.text = ggplot2::element_text(size = 12)
  )

downsample_recap_plot
#ggplot2::ggsave(filename = paste0(results_path, "supplement/recapitulation_rate/downsample_recapitulation_rate_radial_glia.pdf"), plot = downsample_recap_plot, device = "pdf", width = 7, height = 5, units = "in")
```

The 95% CI on the mean recapitulation rate at each downsampled size (computed
from the actual spread across the 10 independent random-subsample replicates,
not shown as error bars above since they are too narrow (~±0.003) to be
visible against the plot's 0-1 y-axis). The full-data anchor point (47,437
cells) has only 1 "replicate" (no resampling was performed on the production
result), so no CI can be computed for it.

```{r}
downsample_summary %>%
  dplyr::transmute(
    `# Cells`       = scales::comma(n_cells),
    `Mean % Recap`  = scales::percent(mean_pct_recap, accuracy = 0.1),
    `SD`            = ifelse(is.na(sd_pct_recap), "--", scales::percent(sd_pct_recap, accuracy = 0.01)),
    `# Replicates`  = n_reps,
    `95% CI`        = ifelse(
      is.na(ci_lower), "--",
      paste0(scales::percent(ci_lower, accuracy = 0.1), " – ", scales::percent(ci_upper, accuracy = 0.1))
    )
  )
```


## Examples of DD Genes

We load the annotated diploid Seurat object (all cell types, all species;
see `cell_annotations_diploid_lines.Rmd`).

The object also carries a "sketch" assay (Seurat v5, BPCells-backed on-disk
matrix), so BPCells must be installed and discoverable on this session's
library path -- see `.libPaths()` below.
```{r}
# BPCells is needed to load the object's on-disk "sketch" assay. Make sure
# it's discoverable regardless of which library path resolves first in this
# session.
.libPaths(c("/project/gilad/awchen55/Rlibs/R4-4-1", .libPaths()))

path_seurat <- '/project/gilad/awchen55/mechanism/data/diploid_crossfilt_data/'
diploid_crossfilt_data <- readRDS(paste0(path_seurat, 'combined_1kFeatures_integrated_annotated.rds'))

DefaultAssay(diploid_crossfilt_data) <- "RNA"
diploid_crossfilt_data[["sketch"]] <- NULL
diploid_crossfilt_data <- JoinLayers(diploid_crossfilt_data)
```

### DD but not DE

Example gene illustrating differential dispersion in the absence of
differential expression: *SESN3* in Neuroblast. The density plot below
shows the human and chimpanzee expression distributions for this gene
overlapping in location (not DE) while differing in shape/spread (DD) --
i.e. the two species show no detectable shift in mean expression, but a
clear difference in cell-to-cell variability.

```{r}
ct <- "Neuroblast"
# Subset to Neuroblast cells only
ct_obj <- subset(diploid_crossfilt_data, subset = celltype == ct)

# Library size per cell. This object's RNA counts layer is BPCells-backed
# on disk, which supports colSums() natively (streamed, no full
# materialization needed) but doesn't register coercion to every standard
# Matrix class (e.g. as(., "CsparseMatrix") fails with "no method... for
# coercing 'RenameDims'"), and edgeR::cpm() calls min() internally which
# BPCells' matrix class doesn't support either. So we skip edgeR::cpm()
# entirely and compute log2 CPM by hand per gene (below), which only ever
# needs one gene's row -- small enough to materialize directly.
counts <- GetAssayData(ct_obj, assay = "RNA", layer = "counts")
libsize <- Matrix::colSums(counts)
```

Plot distribution of expression.
```{r}
gene <- "SESN3"
ct   <- "Neuroblast"

de_col <- paste0("de.adj.pvalue.diploid.", ct)
dd_col <- paste0("dd.lfsr.diploid.", ct)

vals <- dd_results %>%
  dplyr::filter(gene == !!gene) %>%
  dplyr::select(
    de = !!rlang::sym(de_col),
    dd = !!rlang::sym(dd_col)
  ) %>%
  dplyr::slice(1)

# log2 CPM for this one gene, matching edgeR::cpm(log=TRUE, prior.count=0)
gene_counts <- as.numeric(as.matrix(counts[gene, ]))
expr <- log2(gene_counts / libsize * 1e6)

plot_df <- data.frame(
  expr    = expr,
  species = ct_obj$species
)

# rounding helpers
round_down <- function(x, digits = 1) {
  floor(x * 10^digits) / 10^digits
}

round_up <- function(x, digits = 2) {
  ceiling(x * 10^digits) / 10^digits
}

de_label <- format(round_down(vals$de, 1), nsmall = 1)
# round_up_signif() gives proper significant-figure precision (falling back
# to scientific notation, e.g. "8 %*% 10^{-11}", for very small values)
# instead of always rounding to 2 decimals, which floors tiny LFSRs to "0.01".
# Parsing it into a language object (rather than inserting the string via
# bquote's .()) lets the "%*%"/"^{}" plotmath syntax actually render as a
# multiplication sign and superscript instead of literal text.
dd_label <- parse(text = round_up_signif(vals$dd))[[1]]

# plotmath label
label_expr <- bquote(
  atop(
    bold("DE:") ~ "FDR >" ~ .(de_label),
    bold("DD:") ~ "LFSR <" ~ .(dd_label)
  )
)

dd_ex1_plot <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = expr, color = species, fill = species)
) +
  ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::scale_fill_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::labs(
    x = "Expression (log2 CPM)",
    y = "Density",
    title = bquote(italic(.(gene)))
  ) +
  ggplot2::annotate(
    "label",
    x = Inf, y = Inf,
    label = as.character(as.expression(label_expr)),
    parse = TRUE,
    hjust = 1.05,
    vjust = 1.25,
    size = 4,
    label.size = 0.3,
    label.padding = grid::unit(0.25, "lines"),
    fill = "white",
    color = "black"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title   = ggplot2::element_text(size = 14),
    axis.title   = ggplot2::element_text(size = 14),
    axis.text    = ggplot2::element_text(size = 12),
    axis.line    = ggplot2::element_blank(),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.text  = ggplot2::element_text(size = 12),
    legend.title = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(title = NULL),
    fill  = ggplot2::guide_legend(title = NULL)
  )

dd_ex1_plot
#ggsave(paste0(results_path, 'main/dd_no_de_gene_example.pdf'), plot = dd_ex1_plot, device = "pdf", width = 7, height = 5, units = "in")
```

#### Mean expression and dispersion by species (a second DD-not-DE example)

The density plot above shows one example (*SESN3* in Neuroblast) where the
two species' distributions overlap in location (no DE) but differ in shape
(DD). Here is a second, independent DD-not-DE example -- *SETD2* in
Stromal cells -- shown directly in memento's own per-donor mean and
dispersion estimates rather than as a density plot: these two box plots
show each individual donor's mean expression and dispersion, grouped by
species. As with the density plot above, the mean-expression boxes should
overlap between species (no DE) while the dispersion boxes should not (DD).
```{r}
# Per-celltype memento output (mt/rb genes removed, mean-expression cutoff
# and memento_min_cells_per_group already applied upstream), matching
# dd_results above -- NOT the older all-celltypes-combined files under
# data/memento/, which predate the mean cutoff and the low-cell-count donor
# fix (see dispersion_diagnostics.Rmd / q_threshold_sensitivity.Rmd Section 8b).
memento_q02_dir <- "/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q0.2/memento"

ct <- 'Stromal.cells'
gene <- "SETD2"
ct_file <- gsub("\\.", "_", ct)

diploid_mean_est <- read.csv(file.path(memento_q02_dir, paste0("diploid_memento_mean_", ct_file, ".csv")), check.names = FALSE)
diploid_var_est  <- read.csv(file.path(memento_q02_dir, paste0("diploid_memento_var_",  ct_file, ".csv")), check.names = FALSE)
```

```{r}
# Per-celltype files have one column per donor (celltype is implicit in the
# filename): sg^<species>^<donor>
donor_cols <- setdiff(colnames(diploid_mean_est), "gene")

mean_vals <- as.numeric(diploid_mean_est[diploid_mean_est$gene == gene, donor_cols])
disp_vals <- as.numeric(diploid_var_est[diploid_var_est$gene == gene, donor_cols])
donor_species <- sub("^sg\\^([a-z]+)\\^.*$", "\\1", donor_cols)

donor_plot_df <- data.frame(
  species    = donor_species,
  mean       = mean_vals,
  dispersion = disp_vals
)

species_colors <- c("human" = "lightcoral", "chimp" = "turquoise3")
species_labels <- c("human" = "human", "chimp" = "chimpanzee")

p_mean_by_species <- ggplot2::ggplot(donor_plot_df, ggplot2::aes(x = species, y = mean, fill = species)) +
  ggplot2::geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  ggplot2::scale_fill_manual(values = species_colors, breaks = c("human", "chimp"), labels = species_labels) +
  ggplot2::scale_x_discrete(limits = c("human", "chimp"), labels = species_labels) +
  ggplot2::scale_y_continuous(breaks = seq(-8.5, -7.5, by = 0.5)) +
  ggplot2::coord_cartesian(ylim = c(-8.5, -7.5)) +
  ggplot2::labs(x = NULL, y = "Mean expression", title = bquote(italic(.(gene)))) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "none",
    plot.title = ggplot2::element_text(size = 16),
    axis.title = ggplot2::element_text(size = 18),
    axis.text  = ggplot2::element_text(size = 15)
  )
p_dispersion_by_species <- ggplot2::ggplot(donor_plot_df, ggplot2::aes(x = species, y = dispersion, fill = species)) +
  ggplot2::geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  ggplot2::scale_fill_manual(values = species_colors, breaks = c("human", "chimp"), labels = species_labels) +
  ggplot2::scale_x_discrete(limits = c("human", "chimp"), labels = species_labels) +
    ggplot2::coord_cartesian(ylim = c(-1.2, -0.6)) +
  ggplot2::labs(x = NULL, y = "Dispersion") +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "none",
    axis.title = ggplot2::element_text(size = 18),
    axis.text  = ggplot2::element_text(size = 15)
  )

library(patchwork)
p_boxplots <- p_mean_by_species + p_dispersion_by_species
p_boxplots
# ggplot2::ggsave(filename = paste0(results_path, "main/boxplots.pdf"), plot = p_boxplots,  device = "pdf", width = 9, height = 5, units = "in")
```

### DD and DE Opposite Signs

Example gene illustrating a case where a gene is significant for *both* DD
and DE, but the direction of the effect differs between the two: *MAML2*
in ENS glia. Unlike the DD-not-DE examples above, the human and
chimpanzee distributions for this gene differ in both location (DE) and
shape (DD) -- but the species with the higher mean expression is not the
species with the higher dispersion, i.e. the DD and DE effect sizes point
in opposite directions. This is the kind of gene the "Effect Size Plot"
section below (and its cis/trans effect-size comparisons) is built to
characterize systematically across all cell types, rather than one example
at a time.

```{r}
ct <- "ENS glia"
# Subset to ENS glia cells only
ct_obj <- subset(diploid_crossfilt_data, subset = celltype == ct)

# Library size per cell (see note in the "DD but not DE" section above for
# why we skip edgeR::cpm() and compute log2 CPM by hand per gene instead).
counts <- GetAssayData(ct_obj, assay = "RNA", layer = "counts")
libsize <- Matrix::colSums(counts)
```

Plot distribution of expression.
```{r}
gene <- "MAML2"
ct   <- "ENS.glia"

de_col <- paste0("de.adj.pvalue.diploid.", ct)
dd_col <- paste0("dd.lfsr.diploid.", ct)

vals <- dd_results %>%
  dplyr::filter(gene == !!gene) %>%
  dplyr::select(
    de = !!rlang::sym(de_col),
    dd = !!rlang::sym(dd_col)
  ) %>%
  dplyr::slice(1)

# log2 CPM for this one gene, matching edgeR::cpm(log=TRUE, prior.count=0)
gene_counts <- as.numeric(as.matrix(counts[gene, ]))
expr <- log2(gene_counts / libsize * 1e6)

plot_df <- data.frame(
  expr    = expr,
  species = ct_obj$species
)

de_label <- parse(text = round_up_signif(vals$de))[[1]]
# round_up_signif() gives proper significant-figure precision (falling back
# to scientific notation, e.g. "8 %*% 10^{-11}", for very small values)
# instead of always rounding to 2 decimals, which floors tiny LFSRs to "0.01".
# Parsing it into a language object (rather than inserting the string via
# bquote's .()) lets the "%*%"/"^{}" plotmath syntax actually render as a
# multiplication sign and superscript instead of literal text.
dd_label <- parse(text = round_up_signif(vals$dd))[[1]]

# plotmath label
label_expr <- bquote(
  atop(
    bold("DE:") ~ "FDR <" ~ .(de_label),
    bold("DD:") ~ "LFSR <" ~ .(dd_label)
  )
)

dd_ex2_plot <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = expr, color = species, fill = species)
) +
  ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::scale_fill_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::labs(
    x = "Expression (log2 CPM)",
    y = "Density",
    title = bquote(italic(.(gene)))
  ) +
  ggplot2::annotate(
    "label",
    x = Inf, y = Inf,
    label = as.character(as.expression(label_expr)),
    parse = TRUE,
    hjust = 1.05,
    vjust = 1.25,
    size = 4,
    label.size = 0.3,
    label.padding = grid::unit(0.25, "lines"),
    fill = "white",
    color = "black"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title   = ggplot2::element_text(size = 14),
    axis.title   = ggplot2::element_text(size = 14),
    axis.text    = ggplot2::element_text(size = 12),
    axis.line    = ggplot2::element_blank(),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.text  = ggplot2::element_text(size = 12),
    legend.title = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(title = NULL),
    fill  = ggplot2::guide_legend(title = NULL)
  )

dd_ex2_plot
ggsave(paste0(results_path, 'main/dd_de_gene_example.pdf'), plot = dd_ex2_plot, device = "pdf", width = 7, height = 5, units = "in")
```


## Effect Size Plot

We plot DD effect sizes of human vs chimpanzee comparisons. This includes all cell types together. Light grey represents DD genes in the diploid (LFSR < 0.05) and black shows the subset recapitulated in the allotetraploid (LFSR < 0.05 and same sign of effect).
```{r}
plot_df <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_dip_col  <- paste0("dd.lfsr.diploid.", ct)
  lfsr_allo_col <- paste0("dd.lfsr.allotetraploid.", ct)
  same_col      <- paste0("dd.recapitulated.same.sign.", ct)
  dip_col       <- paste0("dd.logFC.diploid.", ct)
  allo_col      <- paste0("dd.logFC.allotetraploid.", ct)

  dd_results %>%
    dplyr::select(
      gene,
      dplyr::all_of(lfsr_dip_col),
      dplyr::all_of(lfsr_allo_col),
      dplyr::all_of(same_col),
      dplyr::all_of(dip_col),
      dplyr::all_of(allo_col)
    ) %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      diploid_lfsr = as.numeric(.data[[lfsr_dip_col]]),
      allotetraploid_lfsr = as.numeric(.data[[lfsr_allo_col]]),
      same_sign = as.integer(.data[[same_col]]),
      diploid_logFC = as.numeric(.data[[dip_col]]),
      allotetraploid_logFC = as.numeric(.data[[allo_col]])
    ) %>%
    dplyr::mutate(
      sig_dd = diploid_lfsr < 0.001,
      black_point = diploid_lfsr < 0.001 &
        allotetraploid_lfsr < 0.001 &
        same_sign == 1
    ) %>%
    dplyr::filter(
      sig_dd,
      !is.na(diploid_logFC),
      !is.na(allotetraploid_logFC)
    ) %>%
    dplyr::mutate(
      same_sign = dplyr::if_else(is.na(same_sign), 0L, same_sign),
      black_point = dplyr::if_else(is.na(black_point), FALSE, black_point)
    )
}) %>%
  dplyr::distinct(
    gene, celltype, same_sign, black_point,
    diploid_logFC, allotetraploid_logFC
  )

quad_df <- plot_df %>%
  dplyr::mutate(
    quadrant = dplyr::case_when(
      allotetraploid_logFC >= 0 & diploid_logFC >= 0 ~ "TR",
      allotetraploid_logFC <  0 & diploid_logFC >= 0 ~ "TL",
      allotetraploid_logFC <  0 & diploid_logFC <  0 ~ "BL",
      TRUE                                           ~ "BR"
    )
  ) %>%
  dplyr::count(quadrant, name = "n") %>%
  dplyr::mutate(pct = 100 * n / sum(n))

xlim <- c(-2.1, 2.1)
ylim <- c(-2.0, 2.0)

label_pos <- tibble::tibble(
  quadrant = c("TL", "TR", "BL", "BR"),
  x = c(xlim[1] + 0.12, xlim[2] - 0.12, xlim[1] + 0.12, xlim[2] - 0.12),
  y = c(ylim[2] - 0.12, ylim[2] - 0.12, ylim[1] + 0.12, ylim[1] + 0.12),
  hjust = c(0, 1, 0, 1),
  vjust = c(1, 1, 0, 0)
)

quad_labels <- label_pos %>%
  dplyr::left_join(quad_df, by = "quadrant") %>%
  dplyr::mutate(
    label = ifelse(is.na(pct), "0%", paste0(sprintf("%.1f", pct), "%"))
  )

lim <- range(
  c(plot_df$allotetraploid_logFC, plot_df$diploid_logFC),
  na.rm = TRUE
)

short_lim <- as.numeric(
  stats::quantile(lim, probs = c(0.1, 0.75), na.rm = TRUE)
)

pad <- 0.15

blue_start <- short_lim[1] + pad
# Extend the upper end further toward the corner than the 75th percentile
# of the data, but stop short of the "% in this quadrant" label
# (positioned at ylim[2] - 0.12) so the arrowhead doesn't overlap it.
blue_end   <- min(xlim[2], ylim[2]) - 0.4

red_bottom <- ylim[1] + pad
red_top    <- ylim[2] - pad

p_effect_size <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = allotetraploid_logFC, y = diploid_logFC)
) +

  ggplot2::geom_point(
    data = dplyr::filter(plot_df, !black_point),
    color = "grey75",
    alpha = 0.8,
    size = 1
  ) +

  ggplot2::geom_point(
    data = dplyr::filter(plot_df, black_point),
    color = "black",
    alpha = 0.7,
    size = 1
  ) +

  ggplot2::geom_label(
    data = quad_labels,
    ggplot2::aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = quad_labels$hjust,
    vjust = quad_labels$vjust,
    size = 4,
    label.size = 0.25,
    fill = "white",
    color = "black"
  ) +

  ggplot2::scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2),
    expand = ggplot2::expansion(mult = 0.02)
  ) +
  ggplot2::scale_y_continuous(
    breaks = c(-2, -1, 0, 1, 2),
    expand = ggplot2::expansion(mult = 0.02)
  ) +

  ggplot2::coord_equal(xlim = xlim, ylim = ylim, clip = "on") +

  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = 0,
      y = red_bottom,
      yend = red_top
    ),
    inherit.aes = FALSE,
    color = "red",
    linewidth = 0.7,
    arrow = grid::arrow(
      length = grid::unit(0.18, "cm"),
      ends = "both",
      type = "closed"
    )
  ) +

  ggplot2::geom_segment(
    ggplot2::aes(
      x = blue_start,
      y = blue_start,
      xend = blue_end,
      yend = blue_end
    ),
    inherit.aes = FALSE,
    color = "blue",
    linewidth = 0.6,
    arrow = grid::arrow(
      length = grid::unit(0.18, "cm"),
      ends = "both",
      type = "closed"
    )
  ) +

  ggplot2::labs(
    x = "logFC human/chimpanzee in allotetraploid cells",
    y = "logFC human/chimpanzee"
  ) +

  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.grid.major = ggplot2::element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor = ggplot2::element_line(color = "grey90", linewidth = 0.25),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.line = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 10),
    axis.text  = ggplot2::element_text(size = 8)
  )

p_effect_size
ggsave(paste0(results_path, 'main/cis_trans_effect_sizes_0.001.pdf'), plot = p_effect_size, device = "pdf", width = 7, height = 5, units = "in")
```



### Supplementary: Percent DD Genes that have Same Sign
Want to see how many significant DD genes in both diploid and allotetraploid settings also have the same sign of effect.
```{r}

same_sign_df <- purrr::map_dfr(celltypes, function(ct) {

  dip_lfsr_col  <- paste0("dd.lfsr.diploid.", ct)
  allo_lfsr_col <- paste0("dd.lfsr.allotetraploid.", ct)

  dip_logfc_col  <- paste0("dd.logFC.diploid.", ct)
  allo_logfc_col <- paste0("dd.logFC.allotetraploid.", ct)

  dd_results %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      dip_lfsr = .data[[dip_lfsr_col]],
      allo_lfsr = .data[[allo_lfsr_col]],
      dip_logFC = .data[[dip_logfc_col]],
      allo_logFC = .data[[allo_logfc_col]]
    ) %>%
    dplyr::filter(
      !is.na(dip_lfsr),
      !is.na(allo_lfsr),
      dip_lfsr < 0.05,
      allo_lfsr < 0.05,
      !is.na(dip_logFC),
      !is.na(allo_logFC)
    ) %>%
    dplyr::summarise(
      n_sig_both = dplyr::n(),
      n_same_sign = sum(sign(dip_logFC) == sign(allo_logFC)),
      percent_same_sign = 100 * n_same_sign / n_sig_both,
      .groups = "drop"
    ) %>%
    dplyr::mutate(celltype = ct)
})

# replace "." with space for nicer labels
same_sign_df <- same_sign_df %>%
  dplyr::mutate(
    celltype_label = stringr::str_replace_all(celltype, "\\.", " ")
  )

# order by descending percent
same_sign_df <- same_sign_df %>%
  dplyr::arrange(desc(percent_same_sign)) %>%
  dplyr::mutate(
    celltype_label = factor(celltype_label, levels = celltype_label)
  )

p_same_sign <- ggplot(same_sign_df, aes(x = celltype_label, y = percent_same_sign)) +
  geom_col(fill = "steelblue", color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = paste0(round(percent_same_sign, 1), "%\n(n=", n_sig_both, ")")),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    x = "",
    y = "% same sign of effect size",
    title = ""
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
    axis.line = element_blank()
  ) +
  coord_cartesian(
    ylim = c(0, max(same_sign_df$percent_same_sign, na.rm = TRUE) * 1.15)
  )

p_same_sign
#ggsave(paste0(results_path, 'supplement/same_sign_effect_sizes.pdf'), plot = p_same_sign, device = "pdf", width = 7, height = 5, units = "in")
```

## Cis vs Trans Analysis

### Number of cis/trans DD genes

Number of DD genes for each cell type.The overall category using mean cis proportion of cell types DD.
```{r}
# -----------------------------
# Per-cell-type counts
# -----------------------------
celltype_counts <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.category.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.cis.category."),
    names_to = "celltype",
    values_to = "category"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.cis\\.category\\."),
    celltype = stringr::str_replace_all(celltype, "\\.", " "),
    category = factor(category, levels = c("cis", "neither", "trans", "mixed", "not DD"))
  ) %>%
  dplyr::count(celltype, category, .drop = FALSE) %>%
  tidyr::pivot_wider(
    names_from = category,
    values_from = n,
    values_fill = 0
  )
celltype_counts

```



### Effect size comparisons

We compare the DD effect sizes in diploid cells for all cell types together. These meet our cutoff of > 0.7 as cis and < 0.3 is trans. We use a K-S test to test for significant differences in the effect sizes.
```{r}
# -------------------------------
# 1. Cis category → long
# -------------------------------
cat_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.category.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.cis.category."),
    names_to = "celltype",
    values_to = "category"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.cis\\.category\\.")
  )

# -------------------------------
# 2. Effect size → long
# -------------------------------
eff_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.logFC.diploid.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.logFC.diploid."),
    names_to = "celltype",
    values_to = "effect_size"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.logFC\\.diploid\\.")
  )

# -------------------------------
# 3. Combine (gene × cell type instances)
# -------------------------------
ecdf_df <- cat_long %>%
  dplyr::inner_join(eff_long, by = c("gene", "celltype")) %>%
  dplyr::filter(celltype %in% celltypes) %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::mutate(
    category = factor(category, levels = c("cis", "neither", "trans"))
  )

ecdf_df_plot <- ecdf_df %>%
  dplyr::filter(category %in% c("cis", "trans")) %>%
  dplyr::mutate(
    category = factor(category, levels = c("cis", "trans"))
  )


# subset effect sizes
cis_vals <- ecdf_df_plot %>%
  dplyr::filter(category == "cis") %>%
  dplyr::pull(abs(effect_size))

trans_vals <- ecdf_df_plot %>%
  dplyr::filter(category == "trans") %>%
  dplyr::pull(abs(effect_size))

# KS test
ks_res <- ks.test(cis_vals, trans_vals)

ks_res

```

```{r}
# format p-value using your function
p_label <- paste0("P < ", round_up_signif(ks_res$p.value))

ecdf_plot <- ggplot2::ggplot(
  ecdf_df_plot,
  ggplot2::aes(x = abs(effect_size), color = category)
) +
  ggplot2::stat_ecdf(linewidth = 1) +

  ggplot2::annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = p_label,
    hjust = -0.2,
    vjust = 2,
    size = 4,
    parse = TRUE   # needed for %*% and 10^{}
  ) +

  ggplot2::labs(
    x = "Absolute logFC",
    y = "Cumulative frequency",
    color = ""
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "cis"   = "#009E73",
      "trans" = "#D55E00"
    ),
    labels = c(
      cis   = expression(italic("cis")),
      trans = expression(italic("trans"))
    )
  ) +
  ggplot2::coord_cartesian(xlim = c(0, 1.19)) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", linewidth = 1),
    panel.grid.major = ggplot2::element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = c(0.9, 0.05),
    legend.justification = c("right", "bottom"),
    legend.background = ggplot2::element_rect(
      fill = scales::alpha("white", 0.7),
      color = NA
    ),
    legend.key = ggplot2::element_blank(),
    legend.key.height = grid::unit(0.9, "cm"),
    legend.key.width  = grid::unit(1.2, "cm"),
    legend.text  = ggplot2::element_text(size = 14),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12)
  )

ecdf_plot
#ggsave(paste0(results_path, 'main/ecdf_cis_trans.pdf'), plot = ecdf_plot, device = "pdf", width = 7, height = 5, units = "in")
```



### Mean cis proportion of DD genes

We compare the mean cis proportion of cell types that are DD to the number of times a gene is DD in a cell type. This includes all genes that are DD in the diploid.
```{r}
# ----------------------------------
# Build long dataframe: one row per gene x cell type
# ----------------------------------
plot_df_long <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col <- paste0("dd.lfsr.diploid.", ct)
  cis_col  <- paste0("dd.cis.proportion.", ct)

  dd_results %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      dd_lfsr = .data[[lfsr_col]],
      cis_prop = .data[[cis_col]]
    )
})

# ----------------------------------
# Keep only diploid-DD gene/celltype pairs
# ----------------------------------
plot_df_long_sig <- plot_df_long %>%
  dplyr::filter(!is.na(dd_lfsr), dd_lfsr < 0.05, !is.na(cis_prop))

# ----------------------------------
# Collapse to gene-level summary
# ----------------------------------
plot_df <- plot_df_long_sig %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    n_ct = dplyr::n(),
    mean_cis_prop = mean(cis_prop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    n_ct = factor(n_ct, levels = sort(unique(n_ct)))
  )

# ----------------------------------
# Compute p-value: first vs last group
# ----------------------------------
n_ct_levels <- sort(unique(as.numeric(as.character(plot_df$n_ct))))
first_group <- as.character(min(n_ct_levels))
last_group  <- as.character(max(n_ct_levels))

p_val <- wilcox.test(
  mean_cis_prop ~ n_ct,
  data = plot_df %>% dplyr::filter(n_ct %in% c(first_group, last_group))
)$p.value

p_label <- paste0("P < ", round_up_signif(p_val))

# ----------------------------------
# Bracket positioning
# ----------------------------------
y_max <- max(plot_df$mean_cis_prop, na.rm = TRUE)
h     <- 0.03
y_pos <- y_max + 0.055

x_start <- which(levels(plot_df$n_ct) == first_group)
x_end   <- which(levels(plot_df$n_ct) == last_group)
x_mid   <- mean(c(x_start, x_end))

# ----------------------------------
# Plot
# ----------------------------------
p_mean_cis_diploid <- ggpubr::ggboxplot(
  plot_df,
  x = "n_ct",
  y = "mean_cis_prop",
  fill = "n_ct",
  color = "black",
  palette = "npg",
  #add = "jitter",
  add.params = list(width = 0.15, alpha = 0.5, size = 1)
) +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 1) +
#   scale_fill_manual(
#   values = colorRampPalette(c("#6BAED6", "#FB6A4A"))(length(unique(plot_df$n_ct)))
# ) +
  
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.2),
  expand = ggplot2::expansion(mult = 0.02)
) +
  annotate("segment", x = x_start, xend = x_end, y = y_pos, yend = y_pos) +
  annotate("segment", x = x_start, xend = x_start, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = x_end, xend = x_end, y = y_pos, yend = y_pos - h) +
  annotate("text", x = x_mid, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +
  labs(
    x = "Number of cell types DD in diploid",
    y = expression(paste("Mean ", italic("cis"), " proportion"))
  ) +
  ggpubr::theme_pubr() +
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.line = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  ) +
  coord_cartesian(
    ylim = c(0, y_pos + 0.05),
    clip = "off"
  )

p_mean_cis_diploid
# ggsave(paste0(results_path, 'main/mean_cis_by_n_celltype_dd.pdf'), plot = p_mean_cis_diploid, device = "pdf", width = 7, height = 5, units = "in")
```


#### Robustness: effect size vs. number of DD cell types

The plot above shows mean *cis* proportion increasing with the number of
cell types a gene is DD in. A possible confound: if genes DD in more cell
types also tend to have larger dispersion effect sizes, and effect size
itself is related to *cis*/*trans* classification, that relationship could
be driving the pattern above rather than the number of cell types per se.
This plot checks whether absolute DD effect size (|logFC|) is in fact
roughly stable across the same number-of-cell-types groups.

```{r}
plot_df_long <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col  <- paste0("dd.lfsr.diploid.", ct)
  logfc_col <- paste0("dd.logFC.diploid.", ct)

  dd_results %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      dd_lfsr = .data[[lfsr_col]],
      dd_logFC = .data[[logfc_col]]
    )
})

plot_df <- plot_df_long %>%
  dplyr::filter(
    !is.na(dd_lfsr),
    dd_lfsr < 0.05,
    !is.na(dd_logFC)
  ) %>%
  dplyr::group_by(gene) %>%
  dplyr::mutate(
    n_ct = dplyr::n()
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    abs_effect_size = abs(dd_logFC),
    n_ct = factor(n_ct, levels = sort(unique(n_ct)))
  )

# Wilcoxon test: first vs last group
n_ct_levels <- sort(unique(as.numeric(as.character(plot_df$n_ct))))
first_group <- as.character(min(n_ct_levels))
last_group  <- as.character(max(n_ct_levels))

p_val <- wilcox.test(
  abs_effect_size ~ n_ct,
  data = plot_df %>% dplyr::filter(n_ct %in% c(first_group, last_group))
)$p.value

p_label <- paste0("P < ", round_up_signif(p_val))

# bracket positioning
y_max <- max(plot_df$abs_effect_size, na.rm = TRUE)
h     <- 0.03 * y_max
y_pos <- y_max + 0.08 * y_max

x_start <- which(levels(plot_df$n_ct) == first_group)
x_end   <- which(levels(plot_df$n_ct) == last_group)
x_mid   <- mean(c(x_start, x_end))

p_abs_effect_size_diploid <- ggpubr::ggboxplot(
  plot_df,
  x = "n_ct",
  y = "abs_effect_size",
  fill = "n_ct",
  color = "black",
  # no jitter
) +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 1) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 6),
    expand = ggplot2::expansion(mult = 0.02)
  ) +
  annotate("segment", x = x_start, xend = x_end, y = y_pos, yend = y_pos) +
  annotate("segment", x = x_start, xend = x_start, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = x_end, xend = x_end, y = y_pos, yend = y_pos - h) +
  annotate("text", x = x_mid, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +
  labs(
    x = "Number of cell types DD in diploid",
    y = "abs(logFC)"
  ) +
  ggpubr::theme_pubr() +
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.line = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  ) +
  coord_cartesian(
    ylim = c(0, y_pos + 2 * h),
    clip = "off"
  )

p_abs_effect_size_diploid
# ggsave(paste0(results_path, 'supplement/robust_mean_cis_proportion/logFC_by_n_celltype_dd.pdf'), plot = p_abs_effect_size_diploid, device = "pdf", width = 7, height = 5, units = "in")
```

### LOEUF

Load LOF data.
```{r}
# Load data to get LOF scores
path_lof <- '/project/gilad/brendan/dispersion/pilot/cHDC_data/cellranger_cluster-mode_trial/analysis/datasets/evo_constraint/'
lof_scores <- read.table(paste0(path_lof,'gnomad.v4.1.constraint_metrics.tsv'), header = TRUE)
lof_scores <- lof_scores %>% filter(mane_select == "true")
```

We compare the loss of function in cis and trans genes. We use our overall definition of cis/trans genes by looking at mean cis proportion of cell type DD and determine cis (>0.7) or trans (<0.3).
```{r}
lof_cutoff <- function(dd_results, lof_scores, cis_cutoff, trans_cutoff) {

  plot_df2 <- dd_results %>%
    dplyr::select(gene, dd.in.diploid.mean.cis.proportion) %>%
    dplyr::mutate(
      combined_category = dplyr::case_when(
        is.na(dd.in.diploid.mean.cis.proportion) ~ "not DD",
        dd.in.diploid.mean.cis.proportion > cis_cutoff   ~ "cis",
        dd.in.diploid.mean.cis.proportion < trans_cutoff ~ "trans",
        TRUE                                             ~ "neither"
      )
    ) %>%
    dplyr::inner_join(
      lof_scores %>%
        dplyr::select(gene, lof.oe),
      by = "gene"
    ) %>%
    dplyr::filter(combined_category != "not DD") %>%
    dplyr::mutate(
      cat = factor(
        combined_category,
        levels = c("cis", "neither", "trans")
      )
    )

  # Wilcoxon: cis vs trans
  pval <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans")) %>%
    dplyr::summarise(p = wilcox.test(lof.oe ~ cat)$p.value) %>%
    dplyr::pull(p)

  p_label <- paste0(
    "P < ",
    round_up_signif(pval)
  )

  y_pos <- min(1.6, max(plot_df2$lof.oe, na.rm = TRUE) * 1.01)
  h <- 0.05
  # uniform clearance above the bracket for every label (plain-decimal and
  # scientific-notation alike), so labels line up consistently across plots.
  label_gap <- 0.16
  label_size <- 6

  p_lof <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = lof.oe, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    # bracket spanning cis (1) to trans (3)
    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "trans"   = "#D55E00",
      "neither" = "#0072B2"
    )) +
    ggplot2::labs(x = "", y = "LOEUF") +
    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 18),
      axis.text = ggplot2::element_text(size = 16)
    ) +
    ggplot2::coord_cartesian(ylim = c(0, 1.8)) +
    ggplot2::scale_x_discrete(labels = c(
      "cis"     = expression(italic("cis")),
      "trans"   = expression(italic("trans")),
      "neither" = "neither"
    ))

  # add p-value label conditionally
  if (pval >= 1e-4) {
    p_lof <- p_lof +
      ggplot2::annotate(
        "text",
        x = 2,
        y = y_pos + label_gap,
        label = p_label,
        size = label_size
      )
  } else {
    p_lof <- p_lof +
      ggplot2::annotate(
        "text",
        x = 2,
        y = y_pos + label_gap,
        label = p_label,
        size = label_size,
        parse = TRUE
      )
  }

  return(p_lof)
}

p_0.7_0.3lof <-lof_cutoff(dd_results,lof_scores,0.7,0.3)
p_0.7_0.3lof

#ggsave(paste0(results_path, 'main/lof_cis_trans.pdf'), plot = p_0.7_0.3lof, device = "pdf", width = 7, height = 5, units = "in")
```


#### Supplementary varying cis/trans cutoff

Robustness check: does the LOEUF (loss-of-function intolerance) difference
between *cis* and *trans* DD genes shown above hold up across a range of
*cis*/*trans* cutoff choices, not just the 0.7/0.3 split used as the main
cutoff? Six panels below sweep the cutoff pair from the most permissive
(0.6/0.4) to the strictest (0.85/0.15).
```{r fig.width=13, fig.height=8}
p_0.6_0.4lof <-lof_cutoff(dd_results,lof_scores,0.6,0.4)
p_0.65_0.35lof <-lof_cutoff(dd_results,lof_scores,0.65,0.35)
p_0.7_0.3lof <-lof_cutoff(dd_results,lof_scores,0.7,0.3)
p_0.75_0.25lof <-lof_cutoff(dd_results,lof_scores,0.75,0.25)
p_0.8_0.2lof <-lof_cutoff(dd_results,lof_scores,0.8,0.2)
p_0.85_0.15lof <-lof_cutoff(dd_results,lof_scores,0.85,0.15)

(p_0.6_0.4lof + p_0.65_0.35lof + p_0.7_0.3lof + p_0.75_0.25lof + p_0.8_0.2lof + p_0.85_0.15lof) +
  patchwork::plot_layout(ncol = 3)
```


### Percent coding identity

We load percent coding identity. We take more conservative estimate of coding divergence by taking the minimum difference between human to chimp and chimp to human.
```{r}
mart <- read.delim(
  gzfile("/project2/gilad/kenneth/Analysis/mart_export.txt.gz"),
  header = TRUE,
  stringsAsFactors = FALSE
)
mart <- mart %>%
  dplyr::mutate(
    percent_identity = pmin(
      #as.numeric(X.id..query.gene.identical.to.target.Chimpanzee.gene),
      as.numeric(X.id..target.Chimpanzee.gene.identical.to.query.gene),
      na.rm = TRUE
    )
  )

mart <- mart %>% filter(Chimpanzee.orthology.confidence..0.low..1.high. ==1)
```

We compare the percent coding identity in cis and trans genes. We use our overall definition of cis/trans genes by looking at mean cis proportion of cell type DD and determine cis (>0.7) or trans (<0.3).
```{r}
pci_cutoff <- function(dd_results, mart, cis_cutoff, trans_cutoff) {

  plot_df <- dd_results %>%
    dplyr::select(gene, dd.in.diploid.mean.cis.proportion) %>%
    dplyr::mutate(
      combined_category = dplyr::case_when(
        is.na(dd.in.diploid.mean.cis.proportion) ~ "not DD",
        dd.in.diploid.mean.cis.proportion > cis_cutoff   ~ "cis",
        dd.in.diploid.mean.cis.proportion < trans_cutoff ~ "trans",
        TRUE                                             ~ "neither"
      )
    ) %>%
    dplyr::left_join(
      mart %>%
        dplyr::select(Gene.name, percent_identity) %>%
        dplyr::rename(gene = Gene.name) %>%
        dplyr::mutate(percent_identity = as.numeric(percent_identity)),
      by = "gene"
    )

  plot_df2 <- plot_df %>%
    dplyr::filter(combined_category != "not DD") %>%
    dplyr::mutate(
      cat = factor(
        combined_category,
        levels = c("cis", "neither", "trans")
      )
    ) %>%
    dplyr::filter(!is.na(percent_identity), is.finite(percent_identity))

  # cis vs trans only
  cis_trans <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans")) %>%
    dplyr::mutate(cat = droplevels(cat))

  pval <- wilcox.test(percent_identity ~ cat, data = cis_trans)$p.value
  p_label <- paste0("P < ", round_up_signif(pval))

  ymin <- 96
  # Lowered from 102 -- percent identity can't exceed 100, so a box/axis
  # extending to 102 looked wrong. 101.8 still clears the label (checked
  # below) without showing a nonsensical >100 tick.
  ymax <- 101.8

  y_pos <- 100.75
  h <- (ymax - ymin) * 0.03
  # uniform clearance above the bracket for every label (plain-decimal and
  # scientific-notation alike), so labels line up consistently across plots.
  # (previously ymax=100.8 was actually *below* y_pos+h, clipping this label)
  label_gap <- h * 3

  p <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = percent_identity, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +
    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("text", x = 2, y = y_pos + label_gap, label = p_label, size = 6, parse = TRUE) +
    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "neither" = "#0072B2",
      "trans"   = "#D55E00"
    )) +
    ggplot2::labs(x = "", y = "Percent coding identity") +
    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", linewidth = 1, fill = NA),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 18),
      axis.text  = ggplot2::element_text(size = 16)
    ) +
    ggplot2::coord_cartesian(ylim = c(ymin, ymax)) +
    ggplot2::scale_y_continuous(breaks = c(96, 98, 100)) +
    ggplot2::scale_x_discrete(labels = c(
      cis     = expression(italic("cis")),
      neither = "neither",
      trans   = expression(italic("trans"))
    ))

  return(p)
}

p_0.7_0.3_pci  <- pci_cutoff(dd_results, mart, 0.7, 0.3)
p_0.7_0.3_pci
# ggsave(paste0(results_path, 'main/pci_cis_trans.pdf'), plot = p_0.7_0.3_pci, device = "pdf", width = 7, height = 5, units = "in")
```

#### Supplementary varying cis/trans

Same robustness check as above, for percent coding identity instead of
LOEUF: does the *cis* vs. *trans* difference hold across the same sweep of
cutoff choices (0.6/0.4 through 0.85/0.15)?
```{r fig.width=13, fig.height=8}
p_0.6_0.4_pci  <- pci_cutoff(dd_results, mart, 0.6, 0.4)
p_0.65_0.35_pci <- pci_cutoff(dd_results, mart, 0.65, 0.35)
p_0.7_0.3_pci  <- pci_cutoff(dd_results, mart, 0.7, 0.3)
p_0.75_0.25_pci <- pci_cutoff(dd_results, mart, 0.75, 0.25)
p_0.8_0.2_pci  <- pci_cutoff(dd_results, mart, 0.8, 0.2)
p_0.85_0.15_pci <- pci_cutoff(dd_results, mart, 0.85, 0.15)

(p_0.6_0.4_pci + p_0.65_0.35_pci + p_0.7_0.3_pci + p_0.75_0.25_pci + p_0.8_0.2_pci + p_0.85_0.15_pci) +
  patchwork::plot_layout(ncol = 3)
```


### Supplementary: Determining cis/trans cutoffs

All 20 cell types.

#### Number of DD Genes

Using cis cutoffs.
```{r fig.width=10, fig.height=6.5}
celltypes_demo <- celltypes  # all 20 cell types
cutoffs   <- seq(0.5, 1, by = 0.1)

# ----------------------------------------------
# 1. Wide -> long for cis proportion
#    dd.cis.proportion.<celltype>
# ----------------------------------------------
cis_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.proportion.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.cis.proportion."),
    names_to  = "celltype",
    values_to = "cis_prop"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.cis\\.proportion\\.", "", celltype)
  )

# ----------------------------------------------
# 2. Wide -> long for diploid DD significance
#    dd.lfsr.diploid.<celltype>
# ----------------------------------------------
lfsr_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.lfsr.diploid.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.lfsr.diploid."),
    names_to  = "celltype",
    values_to = "dd_lfsr_diploid"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.lfsr\\.diploid\\.", "", celltype)
  )

# ----------------------------------------------
# 3. Merge and filter to diploid DD genes only
#    condition: dd.lfsr.diploid < 0.05
# ----------------------------------------------
df_dd <- cis_long %>%
  dplyr::inner_join(lfsr_long, by = c("gene", "celltype")) %>%
  dplyr::filter(
    celltype %in% celltypes_demo,
    !is.na(dd_lfsr_diploid),
    dd_lfsr_diploid < 0.05
  )

# ----------------------------------------------
# 4. Count cis genes passing each cutoff by cell type
#    condition: cis_prop >= cutoff
# ----------------------------------------------
summary_plot_dd <- tidyr::expand_grid(
  celltype = celltypes_demo,
  cutoff   = cutoffs
) %>%
  dplyr::left_join(df_dd, by = "celltype") %>%
  dplyr::group_by(celltype, cutoff) %>%
  dplyr::summarise(
    n_cis_genes = sum(cis_prop >= cutoff, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------
# 4b. Also compute the trans-cutoff counts here (recomputed identically in
#     the trans chunk below) purely so the cis and trans "number of DD
#     genes" plots can share a common y-axis range -- see shared_max_dd_genes.
# ----------------------------------------------
cutoffs_trans_preview <- seq(0, 0.5, by = 0.1)
summary_plot_trans_preview <- tidyr::expand_grid(
  celltype = celltypes_demo,
  cutoff   = cutoffs_trans_preview
) %>%
  dplyr::left_join(df_dd, by = "celltype") %>%
  dplyr::group_by(celltype, cutoff) %>%
  dplyr::summarise(
    n_trans_genes = sum(cis_prop <= cutoff, na.rm = TRUE),
    .groups = "drop"
  )

shared_max_dd_genes <- max(summary_plot_dd$n_cis_genes, summary_plot_trans_preview$n_trans_genes, na.rm = TRUE)

# ----------------------------------------------
# 5. Plot -- shared y-axis range with the trans-cutoff plot below (was
#    independently scaled to each plot's own data, which made the two hard
#    to compare directly)
# ----------------------------------------------
p_cis_cutoff_genes <- ggplot2::ggplot(
  summary_plot_dd,
  ggplot2::aes(x = cutoff, y = n_cis_genes, color = celltype)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::coord_cartesian(ylim = c(0, shared_max_dd_genes * 1.05)) +
  ggplot2::scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::scale_x_continuous(breaks = cutoffs, limits = c(0.5, 1)) +
  ggplot2::scale_color_discrete(labels = function(x) gsub("\\.", " ", x)) +
  ggplot2::labs(
    x = expression(italic("cis") ~ "proportion"),
    y = expression("#" ~ italic("cis") ~ "DD genes"),
    color = "Cell Type"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(
      color = "black",
      fill  = NA,
      linewidth = 1
    ),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 9),
    legend.key.size = grid::unit(0.4, "cm")
  )

p_cis_cutoff_genes
# ggsave(paste0(results_path, 'supplement/cis_trans_cutoffs/cis_cutoff_num_dd_genes.pdf'), plot = p_cis_cutoff_genes, device = "pdf", width = 10, height = 6.5, units = "in")
```

Using trans cutoffs.
```{r fig.width=10, fig.height=6.5}
cutoffs_trans <- seq(0, 0.5, by = 0.1)

# ----------------------------------------------
# 1. Wide -> long for cis proportion
#    dd.cis.proportion.<celltype>
# ----------------------------------------------
cis_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.proportion.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.cis.proportion."),
    names_to  = "celltype",
    values_to = "cis_prop"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.cis\\.proportion\\.", "", celltype)
  )

# ----------------------------------------------
# 2. Wide -> long for diploid DD significance
#    dd.lfsr.diploid.<celltype>
# ----------------------------------------------
lfsr_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.lfsr.diploid.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.lfsr.diploid."),
    names_to  = "celltype",
    values_to = "dd_lfsr_diploid"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.lfsr\\.diploid\\.", "", celltype)
  )

# ----------------------------------------------
# 3. Merge and filter to diploid DD genes only
#    condition: dd.lfsr.diploid < 0.05
# ----------------------------------------------
df_dd <- cis_long %>%
  dplyr::inner_join(lfsr_long, by = c("gene", "celltype")) %>%
  dplyr::filter(
    celltype %in% celltypes_demo,
    !is.na(dd_lfsr_diploid),
    dd_lfsr_diploid < 0.05
  )

# ----------------------------------------------
# 4. Count trans genes at each cutoff by cell type
#    Define trans as: cis_prop <= cutoff
# ----------------------------------------------
summary_plot_trans <- tidyr::expand_grid(
  celltype = celltypes_demo,
  cutoff   = cutoffs_trans
) %>%
  dplyr::left_join(df_dd, by = "celltype") %>%
  dplyr::group_by(celltype, cutoff) %>%
  dplyr::summarise(
    n_trans_genes = sum(cis_prop <= cutoff, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------
# 5. Plot -- shared y-axis range with the cis-cutoff plot above
#    (shared_max_dd_genes computed in that chunk)
# ----------------------------------------------
p_trans_cutoff_genes <- ggplot2::ggplot(
  summary_plot_trans,
  ggplot2::aes(x = cutoff, y = n_trans_genes, color = celltype)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::coord_cartesian(ylim = c(0, shared_max_dd_genes * 1.05)) +
  ggplot2::scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::scale_x_continuous(breaks = cutoffs_trans, limits = c(0, 0.5)) +
  ggplot2::scale_color_discrete(
    labels = function(x) gsub("\\.", " ", x)
  ) +
  ggplot2::labs(
    x = expression(italic("cis") ~ "proportion"),
    y = expression("#" ~ italic("trans") ~ "DD genes"),
    title = "",
    color = "Cell Type"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(
      color = "black",
      fill  = NA,
      linewidth = 1
    ),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 9),
    legend.key.size = grid::unit(0.4, "cm")
  )

p_trans_cutoff_genes
# ggsave(paste0(results_path, 'supplement/cis_trans_cutoffs/trans_cutoff_num_dd_genes.pdf'), plot = p_trans_cutoff_genes, device = "pdf", width = 10, height = 6.5, units = "in")
```

#### DD genes that exist in multiple cell types

With all 20 cell types, showing one bar per exact count (1, 2, 3, ..., 20)
is too sparse -- most of the high counts have very few (or zero) genes.
Instead, `n_celltypes_{cis,trans}` is bucketed into 1, 2, ..., 9, 10+
and gene counts summed within each bucket.

```{r}
# Buckets n (1..max_n) into 1, 2, ..., 9, 10+ (the last bucket spans
# 10..max_n but is always labeled "10+").
bucket_n_celltypes <- function(n, max_n) {
  breaks <- unique(pmin(c(0:9, max_n), max_n))
  labs   <- c(as.character(1:9), "10+")[seq_len(length(breaks) - 1)]
  cut(n, breaks = breaks, labels = labs, include.lowest = TRUE)
}
```

Using cis cutoffs.
```{r fig.width=10, fig.height=6.5}
N_ct <- length(celltypes_demo)
bucket_levels <- levels(bucket_n_celltypes(1:N_ct, N_ct))

# For each cutoff, count how many cell types each gene is cis-like in
# (defined as dd.cis.proportion >= cutoff among diploid DD genes)
gene_ct_counts_cis_dd <- df_dd %>%
  dplyr::filter(celltype %in% celltypes_demo) %>%
  tidyr::expand_grid(cutoff = cutoffs) %>%
  dplyr::filter(cis_prop >= cutoff) %>%
  dplyr::distinct(gene, celltype, cutoff) %>%
  dplyr::count(cutoff, gene, name = "n_celltypes_cis")

# Bucketed histogram: sum of genes within each bucket of cell-type counts
plot_df_cis_dd <- gene_ct_counts_cis_dd %>%
  dplyr::filter(n_celltypes_cis >= 1) %>%
  dplyr::mutate(bucket = bucket_n_celltypes(n_celltypes_cis, N_ct)) %>%
  dplyr::count(cutoff, bucket, name = "n_genes", .drop = FALSE) %>%
  tidyr::complete(
    cutoff,
    bucket = factor(bucket_levels, levels = bucket_levels),
    fill = list(n_genes = 0)
  ) %>%
  dplyr::mutate(
    cutoff = factor(cutoff, levels = sort(unique(cutoffs)))
  )

# Also compute the trans-cutoff bucketed counts here (recomputed identically
# in the trans chunk below) purely so the cis and trans "# DD genes in N cell
# types" plots can share a common y-axis range.
plot_df_trans_dd_preview <- gene_ct_counts_trans_dd_preview <- df_dd %>%
  dplyr::filter(celltype %in% celltypes_demo) %>%
  tidyr::expand_grid(cutoff = cutoffs_trans) %>%
  dplyr::filter(cis_prop <= cutoff) %>%
  dplyr::distinct(gene, celltype, cutoff) %>%
  dplyr::count(cutoff, gene, name = "n_celltypes_trans") %>%
  dplyr::filter(n_celltypes_trans >= 1) %>%
  dplyr::mutate(bucket = bucket_n_celltypes(n_celltypes_trans, N_ct)) %>%
  dplyr::count(cutoff, bucket, name = "n_genes", .drop = FALSE)

shared_max_dd_genes_per_ct <- max(plot_df_cis_dd$n_genes, plot_df_trans_dd_preview$n_genes, na.rm = TRUE)

# Plot
p_cis_dd_celltypes <- ggplot2::ggplot(
  plot_df_cis_dd,
  ggplot2::aes(x = bucket, y = n_genes)
) +
  ggplot2::geom_col(fill = "#619CFF") +
  ggplot2::geom_text(
    ggplot2::aes(label = n_genes),
    vjust = -0.3,
    size = 3
  ) +
  ggplot2::facet_wrap(~ cutoff, ncol = 3) +
  ggplot2::coord_cartesian(ylim = c(0, shared_max_dd_genes_per_ct * 1.15)) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.15))
  ) +
  ggplot2::labs(
    x = expression("# cell types where" ~ italic("cis") ~ "DD genes exist"),
    y = "# DD genes",
    title = ""
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    ),
    axis.line = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 9),
    strip.background = ggplot2::element_rect(
      fill = "grey88",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = ggplot2::element_text(face = "bold"),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.3)
  )

p_cis_dd_celltypes
# ggsave(paste0(results_path, 'supplement/cis_trans_cutoffs/cis_cutoff_num_dd_genes_per_celltype.pdf'), plot = p_cis_dd_celltypes, device = "pdf", width = 10, height = 6.5, units = "in")
```

Using trans cutoffs.
```{r fig.width=10, fig.height=6.5}
# For each cutoff, count how many cell types each gene is trans-like in
# (defined as dd.cis.proportion <= cutoff among diploid DD genes)
gene_ct_counts_trans_dd <- df_dd %>%
  dplyr::filter(celltype %in% celltypes_demo) %>%
  tidyr::expand_grid(cutoff = cutoffs_trans) %>%
  dplyr::filter(cis_prop <= cutoff) %>%
  dplyr::distinct(gene, celltype, cutoff) %>%
  dplyr::count(cutoff, gene, name = "n_celltypes_trans")

# Bucketed histogram: sum of genes within each bucket of cell-type counts
plot_df_trans_dd <- gene_ct_counts_trans_dd %>%
  dplyr::filter(n_celltypes_trans >= 1) %>%
  dplyr::mutate(bucket = bucket_n_celltypes(n_celltypes_trans, N_ct)) %>%
  dplyr::count(cutoff, bucket, name = "n_genes", .drop = FALSE) %>%
  tidyr::complete(
    cutoff,
    bucket = factor(bucket_levels, levels = bucket_levels),
    fill = list(n_genes = 0)
  ) %>%
  dplyr::mutate(
    cutoff = factor(cutoff, levels = sort(unique(cutoffs_trans)))
  )

# Plot
p_trans_cutoff_celltype <- ggplot2::ggplot(
  plot_df_trans_dd,
  ggplot2::aes(x = bucket, y = n_genes)
) +
  ggplot2::geom_col(fill = "#619CFF") +
  ggplot2::geom_text(
    ggplot2::aes(label = n_genes),
    vjust = -0.3,
    size = 3
  ) +
  ggplot2::facet_wrap(~ cutoff, ncol = 3) +
  ggplot2::coord_cartesian(ylim = c(0, shared_max_dd_genes_per_ct * 1.15)) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.15))
  ) +
  ggplot2::labs(
    x = expression("# cell types where" ~ italic("trans") ~ "DD genes exist"),
    y = "# DD genes",
    title = ""
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    ),
    axis.line = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 9),
    strip.background = ggplot2::element_rect(
      fill = "grey88",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = ggplot2::element_text(face = "bold"),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.3)
  )

p_trans_cutoff_celltype
# ggsave(paste0(results_path, 'supplement/cis_trans_cutoffs/trans_cutoff_num_dd_genes_per_celltype.pdf'), plot = p_trans_cutoff_celltype, device = "pdf", width = 10, height = 6.5, units = "in")
```

### Supplementary: Cis/Trans Replication at a Stricter Mean Expression Cutoff (-7)

Everywhere above uses `dd_results`, which was already built with memento's
`mean_expr_cutoff = -6` (per gene x cell type, matching the convention in
`de_dd_snakemake/scripts/memento_diploid.py`: a gene is retained for a cell
type only if *all* its `(species, donor_id)` group mean estimates are
`<= -6`). This section checks whether the cis/trans results replicate under
a **stricter -7 cutoff** -- i.e. removing genes that passed -6 but wouldn't
pass -7 -- using the same cis (> 0.7) / trans (< 0.3) mean cis-proportion
definition as the main analysis throughout this document.

```{r}
# Per-donor mean estimates, per cell type -- the same per-celltype memento
# output files dd_results was built from (mt/rb genes removed,
# memento_min_cells_per_group already applied), NOT the older
# all-celltypes-combined file under data/memento/, which predates both fixes
# (it still includes MT genes and low-cell-count donors like C3651 that the
# production pipeline excludes -- see q_threshold_sensitivity.Rmd Section 8b).
#
# For each cell type, find genes failing the stricter -7 cutoff -- mirroring
# memento_diploid.py's own retained_genes = mean.index[(mean <= cutoff).all
# (axis=1)]: a gene fails if *any* of its (species, donor_id) group means for
# that cell type is missing or > -7.
fail_minus7_by_ct <- purrr::map(celltypes, function(ct) {
  ct_file <- gsub("\\.", "_", ct)
  mean_f  <- file.path(memento_q02_dir, paste0("diploid_memento_mean_", ct_file, ".csv"))
  if (!file.exists(mean_f)) return(character(0))
  mean_df <- read.csv(mean_f, check.names = FALSE)
  cols    <- setdiff(colnames(mean_df), "gene")
  sub_mat <- as.matrix(mean_df[, cols])
  pass    <- apply(sub_mat, 1, function(x) all(!is.na(x)) && all(x <= -7))
  mean_df$gene[!pass]
})
names(fail_minus7_by_ct) <- celltypes
```

```{r}
# Working copy of dd_results with the -7-failing genes NA'd out, per cell
# type -- same effect as if -7 had been the pipeline's cutoff from the start,
# without re-running memento.
dd_results_m7 <- dd_results
for (ct in celltypes) {
  fail_genes <- fail_minus7_by_ct[[ct]]
  idx <- dd_results_m7$gene %in% fail_genes
  dd_results_m7[idx, paste0("dd.lfsr.diploid.", ct)]     <- NA
  dd_results_m7[idx, paste0("dd.cis.proportion.", ct)]   <- NA
  dd_results_m7[idx, paste0("dd.cis.category.", ct)]     <- NA
}

# Recompute the overall per-gene mean cis proportion (used by the LOEUF/PCI
# panels below) on the -7-masked data, using the same aggregation as the
# "Mean cis proportion of DD genes" section above: average dd.cis.proportion
# across cell types where the gene is diploid-DD (lfsr < 0.05).
plot_df_long_m7 <- purrr::map_dfr(celltypes, function(ct) {
  lfsr_col <- paste0("dd.lfsr.diploid.", ct)
  cis_col  <- paste0("dd.cis.proportion.", ct)
  dd_results_m7 %>%
    dplyr::transmute(gene = gene, celltype = ct,
                      dd_lfsr = .data[[lfsr_col]], cis_prop = .data[[cis_col]])
})
plot_df_long_sig_m7 <- plot_df_long_m7 %>%
  dplyr::filter(!is.na(dd_lfsr), dd_lfsr < 0.05, !is.na(cis_prop))

mean_cis_m7 <- plot_df_long_sig_m7 %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(dd.in.diploid.mean.cis.proportion.m7 = mean(cis_prop, na.rm = TRUE), .groups = "drop")

dd_results_m7 <- dd_results_m7 %>%
  dplyr::select(-dd.in.diploid.mean.cis.proportion) %>%
  dplyr::left_join(mean_cis_m7, by = "gene") %>%
  dplyr::rename(dd.in.diploid.mean.cis.proportion = dd.in.diploid.mean.cis.proportion.m7)

cat("Genes with non-NA overall mean cis proportion -- original (-6):",
    sum(!is.na(dd_results$dd.in.diploid.mean.cis.proportion)),
    " | at -7:", sum(!is.na(dd_results_m7$dd.in.diploid.mean.cis.proportion)), "\n")
```

**Upper left panel -- cis vs trans ECDF** (same construction as the "Effect
size comparisons" section above, on `dd_results_m7`).
```{r}
cat_long_m7 <- dd_results_m7 %>%
  dplyr::select(gene, starts_with("dd.cis.category.")) %>%
  tidyr::pivot_longer(cols = starts_with("dd.cis.category."), names_to = "celltype", values_to = "category") %>%
  dplyr::mutate(celltype = stringr::str_remove(celltype, "^dd\\.cis\\.category\\."))

eff_long_m7 <- dd_results_m7 %>%
  dplyr::select(gene, starts_with("dd.logFC.diploid.")) %>%
  tidyr::pivot_longer(cols = starts_with("dd.logFC.diploid."), names_to = "celltype", values_to = "effect_size") %>%
  dplyr::mutate(celltype = stringr::str_remove(celltype, "^dd\\.logFC\\.diploid\\."))

ecdf_df_m7 <- cat_long_m7 %>%
  dplyr::inner_join(eff_long_m7, by = c("gene", "celltype")) %>%
  dplyr::filter(celltype %in% celltypes) %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::mutate(category = factor(category, levels = c("cis", "neither", "trans")))

ecdf_df_plot_m7 <- ecdf_df_m7 %>%
  dplyr::filter(category %in% c("cis", "trans")) %>%
  dplyr::mutate(category = factor(category, levels = c("cis", "trans")))

cis_vals_m7   <- ecdf_df_plot_m7 %>% dplyr::filter(category == "cis")   %>% dplyr::pull(abs(effect_size))
trans_vals_m7 <- ecdf_df_plot_m7 %>% dplyr::filter(category == "trans") %>% dplyr::pull(abs(effect_size))
ks_res_m7 <- ks.test(cis_vals_m7, trans_vals_m7)
ks_res_m7
p_label_m7 <- paste0("P < ", round_up_signif(ks_res_m7$p.value))

ecdf_plot_m7 <- ggplot2::ggplot(ecdf_df_plot_m7, ggplot2::aes(x = abs(effect_size), color = category)) +
  ggplot2::stat_ecdf(linewidth = 1) +
  ggplot2::annotate("text", x = -Inf, y = Inf, label = p_label_m7, hjust = -0.2, vjust = 2, size = 4, parse = TRUE) +
  ggplot2::labs(x = "Absolute logFC", y = "Cumulative frequency", color = "") +
  ggplot2::scale_color_manual(
    values = c("cis" = "#009E73", "trans" = "#D55E00"),
    labels = c(cis = expression(italic("cis")), trans = expression(italic("trans")))
  ) +
  ggplot2::coord_cartesian(xlim = c(0, 1.19)) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", linewidth = 1),
    panel.grid.major = ggplot2::element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = c(0.9, 0.05),
    legend.justification = c("right", "bottom"),
    legend.background = ggplot2::element_rect(fill = scales::alpha("white", 0.7), color = NA),
    legend.key = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12)
  )
```

**Upper right panel -- mean *cis* proportion vs. number of cell types DD**
(same construction as the "Mean cis proportion of DD genes" section above,
on `dd_results_m7`).
```{r}
plot_df_b_m7 <- plot_df_long_sig_m7 %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(n_ct = dplyr::n(), mean_cis_prop = mean(cis_prop, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(n_ct = factor(n_ct, levels = sort(unique(n_ct))))

n_ct_levels_m7 <- sort(unique(as.numeric(as.character(plot_df_b_m7$n_ct))))
first_group_m7 <- as.character(min(n_ct_levels_m7))
last_group_m7  <- as.character(max(n_ct_levels_m7))

p_val_b_m7 <- wilcox.test(
  mean_cis_prop ~ n_ct,
  data = plot_df_b_m7 %>% dplyr::filter(n_ct %in% c(first_group_m7, last_group_m7))
)$p.value
p_label_b_m7 <- paste0("P < ", round_up_signif(p_val_b_m7))

y_max_b_m7 <- max(plot_df_b_m7$mean_cis_prop, na.rm = TRUE)
h_b_m7 <- 0.03
y_pos_b_m7 <- y_max_b_m7 + 0.055
x_start_b_m7 <- which(levels(plot_df_b_m7$n_ct) == first_group_m7)
x_end_b_m7   <- which(levels(plot_df_b_m7$n_ct) == last_group_m7)
x_mid_b_m7   <- mean(c(x_start_b_m7, x_end_b_m7))

p_mean_cis_m7 <- ggpubr::ggboxplot(
  plot_df_b_m7, x = "n_ct", y = "mean_cis_prop", fill = "n_ct", color = "black",
  palette = "npg", add.params = list(width = 0.15, alpha = 0.5, size = 1)
) +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 1) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.2), expand = ggplot2::expansion(mult = 0.02)) +
  annotate("segment", x = x_start_b_m7, xend = x_end_b_m7, y = y_pos_b_m7, yend = y_pos_b_m7) +
  annotate("segment", x = x_start_b_m7, xend = x_start_b_m7, y = y_pos_b_m7, yend = y_pos_b_m7 - h_b_m7) +
  annotate("segment", x = x_end_b_m7, xend = x_end_b_m7, y = y_pos_b_m7, yend = y_pos_b_m7 - h_b_m7) +
  annotate("text", x = x_mid_b_m7, y = y_pos_b_m7 + h_b_m7, label = p_label_b_m7, size = 4, parse = TRUE) +
  labs(
    x = "Number of cell types DD in diploid",
    y = expression(paste("Mean ", italic("cis"), " proportion"))
  ) +
  ggpubr::theme_pubr() +
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.line = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  ) +
  coord_cartesian(ylim = c(0, y_pos_b_m7 + 0.05), clip = "off")
```

**Lower left panel -- LOEUF** and **lower right panel -- percent coding
identity**, reusing the
`lof_cutoff()`/`pci_cutoff()` functions defined earlier in this document
(they take `dd_results` as a parameter, so no changes were needed -- just
called on `dd_results_m7` instead) at the same 0.7/0.3 cutoff as the main
analysis.
```{r}
p_lof_m7 <- lof_cutoff(dd_results_m7, lof_scores, 0.7, 0.3)
p_pci_m7 <- pci_cutoff(dd_results_m7, mart, 0.7, 0.3)
```

```{r, fig.width=12, fig.height=10}
library(patchwork)
p_cistrans_m7 <- (ecdf_plot_m7 + p_mean_cis_m7) / (p_lof_m7 + p_pci_m7)
p_cistrans_m7
# ggsave(paste0(results_path, 'supplement/cis_trans_mean_cutoff_m7/cis_trans_replication_m7.pdf'), plot = p_cistrans_m7, device = "pdf", width = 12, height = 10, units = "in")
```

At a stricter -7 mean-expression cutoff, the cis/trans results replicate:
the ECDF separation between *cis* and *trans* effect sizes remains highly
significant (upper left), mean *cis* proportion still rises with the number
of cell types a gene is DD in (upper right), and the LOEUF/percent-coding-
identity comparisons between *cis* and *trans* genes (bottom row) retain the
same direction and remain significant. Removing the -6-to-7 boundary genes
(`r sum(sapply(fail_minus7_by_ct, length))` gene x cell type instances
removed in total, `r sum(!is.na(dd_results$dd.in.diploid.mean.cis.proportion)) - sum(!is.na(dd_results_m7$dd.in.diploid.mean.cis.proportion))`
fewer genes with a defined overall mean *cis* proportion) doesn't change any
of the qualitative conclusions.


## Supplementary: Cell Counts

Updated to the current diploid/allotetraploid annotation dataset
(`data/cell_annotations/diploid_metadata.csv`, `tetraploid_metadata.csv` --
see `cell_annotations_diploid_lines.Rmd`), covering all 20 cell types used
throughout this document (previously only 5).

Cell counts here **exclude the same donor x cell-type combinations dropped
by the `memento_min_cells_per_group` (100-cell) filter** used upstream in
the DD/DE pipeline (see `q_threshold_sensitivity.Rmd` Section 8b for the
original diagnosis, and `diploid_pseudobulk_ct.R` / `memento_diploid.py` for
where the filter is applied). Without this, the cell counts shown here
would imply more replication than the DD/DE numbers actually reflect.

```{r}
diploid_meta_path <- '/project/gilad/awchen55/mechanism/data/cell_annotations/diploid_metadata.csv'

diploid_metadata_raw <- readr::read_csv(
  diploid_meta_path,
  col_select = c(cell_barcode, donor_id, species, celltype),
  show_col_types = FALSE
) %>%
  dplyr::filter(celltype %in% celltypes_display, !is.na(species), !is.na(donor_id))

# Donor x cell-type combinations excluded from DD/DE (n_cells < 100), computed
# directly from the diploid metadata at the same threshold as
# memento_min_cells_per_group (see memento_diploid.py / config.yaml), rather
# than a manually-maintained table.
excluded_donor_celltypes <- diploid_metadata_raw %>%
  dplyr::count(celltype, species, donor_id, name = "n_cells") %>%
  dplyr::filter(n_cells < 100)

cat("Excluded donor x cell-type combinations:", nrow(excluded_donor_celltypes), "\n")
```

```{r}
diploid_metadata <- diploid_metadata_raw %>%
  dplyr::anti_join(excluded_donor_celltypes, by = c("celltype", "species", "donor_id"))

cat("Diploid cells after excluding low-count donors:", nrow(diploid_metadata), "\n")
```

```{r fig.width=11, fig.height=6}
diploid_counts <- diploid_metadata %>%
  dplyr::count(celltype, species, name = "cell_count") %>%
  dplyr::mutate(species = dplyr::recode(species, chimp = "Chimpanzee", human = "Human"))

# Allotetraploid: one genome copy per cell (species == "human" barcodes),
# matching the convention already used elsewhere in this document.
allo_counts <- metadata_tetraploid %>%
  dplyr::filter(
    stringr::str_starts(cell_barcode, "human_"),
    celltype != "unassigned",
    celltype %in% celltypes_display
  ) %>%
  dplyr::count(celltype, name = "cell_count") %>%
  dplyr::mutate(species = "Allotetraploid")

plot_df <- dplyr::bind_rows(diploid_counts, allo_counts)

celltype_order_counts <- plot_df %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(total = sum(cell_count), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(total)) %>%
  dplyr::pull(celltype)

plot_df <- plot_df %>%
  dplyr::mutate(
    celltype = factor(celltype, levels = celltype_order_counts),
    species  = factor(species, levels = c("Human", "Chimpanzee", "Allotetraploid"))
  )

dodge <- position_dodge(width = 0.9)

p_cell_counts <- ggplot(plot_df, aes(x = celltype, y = cell_count, fill = species)) +
  geom_col(
    position = dodge,
    width = 0.75,
    color = "black",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = scales::comma(cell_count)),
    position = dodge,
    vjust = 0.5,
    hjust = -0.2,
    angle = 90,
    size = 4
  ) +
  scale_fill_manual(
    values = c(
      "Human" = "#D55E00",
      "Chimpanzee" = "#0072B2",
      "Allotetraploid" = "#009E73"
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    x = "",
    y = "Number of cells",
    fill = ""
  ) +
  coord_cartesian(ylim = c(0, max(plot_df$cell_count, na.rm = TRUE) * 1.1), clip = "off") +
  theme_classic(base_size = 12) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.line = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 15),
    legend.position = "top",
    plot.margin = margin(10,10,10,10)
  )

p_cell_counts
# ggsave(paste0(results_path, 'supplement/cell_type_proportions/cellcounts.pdf'), plot = p_cell_counts, device = "pdf", width = 20, height = 7, units = "in")
```

### Cell counts by individual donor

Total cell count per individual donor (low-count donor x cell-type
combinations excluded, restricted to the 20 cell types used throughout this
analysis), plus one combined bar for the allotetraploid line. The
allotetraploid has no separate donors/individuals -- it's a single hybrid
line -- and each hybrid cell contributes both a human-allele and a
chimp-allele row, so we count only the human-allele rows (same convention
used for allotetraploid cell counts elsewhere in this document) to avoid
double-counting. Bars are ordered allotetraploid first, then chimp
individuals, then human individuals, descending by cell count within each
group.

```{r}
donor_counts <- diploid_metadata %>%
  dplyr::count(species, donor_id, name = "n_cells") %>%
  dplyr::mutate(species = dplyr::recode(species, chimp = "Chimpanzee", human = "Human"))

allo_total <- metadata_tetraploid %>%
  dplyr::filter(
    stringr::str_starts(cell_barcode, "human_"),
    celltype != "unassigned",
    celltype %in% celltypes_display
  ) %>%
  nrow()

donor_counts <- dplyr::bind_rows(
  donor_counts,
  # HLI-25: the allotetraploid hybrid line ID (not present in the metadata,
  # added manually).
  tibble::tibble(species = "Allotetraploid", donor_id = "HLI-25", n_cells = allo_total)
)

order_within <- function(sp) {
  donor_counts %>%
    dplyr::filter(species == sp) %>%
    dplyr::arrange(dplyr::desc(n_cells)) %>%
    dplyr::pull(donor_id)
}

donor_order <- c("HLI-25", order_within("Chimpanzee"), order_within("Human"))

donor_counts
```

```{r fig.width=15, fig.height=6}

donor_counts <- donor_counts %>%
  dplyr::mutate(
    donor_id = factor(donor_id, levels = donor_order),
    species  = factor(species, levels = c("Allotetraploid", "Chimpanzee", "Human"))
  )

p_donor_counts <- ggplot(
  donor_counts,
  aes(x = donor_id, y = n_cells, fill = species)
) +
  geom_col(color = "black", linewidth = 0.3) +
  geom_text(aes(label = scales::comma(n_cells)), vjust = -0.3, size = 6) +
  scale_fill_manual(values = c(
    "Human" = "#D55E00", "Chimpanzee" = "#0072B2", "Allotetraploid" = "#009E73"
  )) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(x = "", y = "Number of cells", fill = "") +
  theme_classic(base_size = 14) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.ticks = element_line(linewidth = 0.6),
    axis.ticks.length = unit(6, "pt"),
    legend.position = "top",
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  )

p_donor_counts
# ggsave(paste0(results_path, 'supplement/cell_type_proportions/donor_cellcounts.pdf'), plot = p_donor_counts, device = "pdf", width = 15, height = 6, units = "in")
```

### Cell type proportions by individual donor

For each of the 20 cell types, what fraction of that cell type's cells come from each
individual donor -- but using the filtered data above (excluded low-count
donors removed) and all 20 cell types rather than the full annotation set.

```{r fig.width=8, fig.height=7}
shared_celltype_order <- diploid_metadata %>%
  dplyr::count(celltype, name = "total_cells") %>%
  dplyr::arrange(dplyr::desc(total_cells)) %>%
  dplyr::pull(celltype)

make_species_prop_plot <- function(metadata_df, species_name, title, celltype_order) {
  plot_df <- metadata_df %>%
    dplyr::filter(species == species_name, celltype %in% celltype_order) %>%
    dplyr::count(celltype, donor_id, name = "cell_count") %>%
    dplyr::group_by(celltype) %>%
    dplyr::mutate(proportion = cell_count / sum(cell_count)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(celltype = factor(celltype, levels = rev(celltype_order)))

  ggplot(plot_df, aes(x = proportion, y = celltype, fill = donor_id)) +
    geom_col(width = 0.8) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
    labs(title = title, x = "Proportion of cells", y = "", fill = "Individual") +
    theme_classic() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
      axis.text.y   = element_text(size = 9),
      legend.position = "right"
    )
}

p_human_prop <- make_species_prop_plot(diploid_metadata, "human", "Human",      shared_celltype_order)
p_chimp_prop <- make_species_prop_plot(diploid_metadata, "chimp", "Chimpanzee", shared_celltype_order)

p_human_prop
p_chimp_prop
ggsave(paste0(results_path, 'supplement/cell_type_proportions/human_diploid_celltype_prop.pdf'), plot = p_human_prop, device = "pdf", width = 8, height = 7, units = "in")
ggsave(paste0(results_path, 'supplement/cell_type_proportions/chimp_diploid_celltype_prop.pdf'), plot = p_chimp_prop,  device = "pdf", width = 8, height = 7, units = "in")
```





