---
title: "Sensitivity of DD Gene Calls to Memento Capture-Efficiency (q)"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
date: "2026-07-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

`q` is the assumed UMI capture efficiency passed to `memento` (see
`memento_q` in `de_dd_snakemake/config.yaml`). Each value of `q` produces its
own `DD_DE_interspecies_results.csv` under
`data/dd_de_results/q_thresholds/q{q}/`, with its own gene universe (memento's
internal filtering means more/fewer genes pass through as `q` changes).

This notebook asks: **how sensitive are the diploid DD gene calls, and the
recapitulated DD gene calls, to the choice of q?** For each cell type we
compare the DD gene sets across `q = 0.05, 0.1, 0.15, 0.2` and report what
fraction of genes are called consistently.

- **Diploid DD**: `dd.lfsr.diploid.{ct} < 0.05`
- **Recapitulated DD**: diploid DD AND `dd.lfsr.allotetraploid.{ct} < 0.05`
  AND same-sign logFC (diploid vs. allotetraploid)

Because the gene universe differs across `q`, all comparisons are done by
matching on gene name, not row position.

---

## Load Data

Reads all four `q` versions of the master DD/DE table into one named list
(`dd_tables[["0.05"]]`, `dd_tables[["0.1"]]`, ...) and sets up shared paths
and helper functions used throughout the rest of the document.

```{r load}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(tibble)

q_values <- c("0.05", "0.1", "0.15", "0.2")

# Master DD/DE table location for a given q (mt/rb genes removed, mean-
# expression cutoff already applied upstream -- see
# code/de_dd_snakemake_mt_rb_mean_filtered/README.md).
path_for_q <- function(q) {
  sprintf("/project/gilad/awchen55/mechanism/data/dd_de_results/mt_rb_mean_filtered_q_thresholds/q%s/DD_DE_interspecies_results.csv", q)
}

dd_tables <- setNames(lapply(q_values, function(q) read.csv(path_for_q(q), check.names = FALSE)), q_values)

# Sanity check: gene universe size should grow (or stay flat) as q becomes
# more permissive.
for (q in q_values) {
  cat(sprintf("q = %-5s : %d genes x %d columns\n", q, nrow(dd_tables[[q]]), ncol(dd_tables[[q]])))
}

results_path <- '/project/gilad/awchen55/mechanism/results/dispersion_paper/'

# All retention/fate plots (Section 4 below) are saved here
q_sensitivity_path <- paste0(results_path, "supplement/q_sensitivity/")
dir.create(q_sensitivity_path, recursive = TRUE, showWarnings = FALSE)

# Consistent naming for every retention/fate plot: which gene set, which
# baseline q, which lfsr significance threshold.
retention_fname <- function(kind, baseline_q, lfsr) {
  sprintf("%s_dd_retained_baseline_q%s_lfsr%s.pdf", kind, baseline_q, lfsr)
}
```

Determines which cell types are available for diploid DD, allotetraploid DD,
and their intersection (the set eligible for recapitulation), from the
column names of the q=0.2 table.

```{r celltypes}
get_celltypes <- function(df, prefix) {
  cols <- colnames(df)[grepl(paste0("^", prefix, "\\."), colnames(df))]
  unique(sub(paste0("^", prefix, "\\."), "", cols))
}

# Cell type sets are identical across q values (verified), so just use q[1]
celltypes_dip  <- get_celltypes(dd_tables[[q_values[1]]], "dd.lfsr.diploid")
celltypes_allo <- get_celltypes(dd_tables[[q_values[1]]], "dd.lfsr.allotetraploid")
celltypes_recap <- intersect(celltypes_dip, celltypes_allo)

cat("Diploid cell types:  ", length(celltypes_dip), "\n")
cat("Allo cell types:     ", length(celltypes_allo), "\n")
cat("Shared (for recap):  ", length(celltypes_recap), "\n")

# Cell type names come from column names, where spaces were converted to
# periods (e.g. "Cardiac mesoderm cells" -> "Cardiac.mesoderm.cells"). Use
# the dotted form for column lookups, and pretty_ct() to convert back to
# spaces for anything shown in a table or plot.
pretty_ct <- function(x) gsub("\\.", " ", x, fixed = FALSE)
```

---

## 1. Gene Universe Size per q

How many genes pass memento's internal filtering at each `q`, before any DD
significance test is applied -- the denominator that everything else in
this document is drawn from.

```{r universe_size}
universe_df <- tibble(
  q     = q_values,
  genes = sapply(q_values, function(q) nrow(dd_tables[[q]]))
)
print(universe_df)

ggplot(universe_df, aes(x = q, y = genes)) +
  geom_col(fill = "#4575b4", width = 0.6) +
  labs(title = "Gene universe size per q", x = "q (memento capture efficiency)",
       y = "Number of genes in table") +
  theme_bw()
```

---

## 2. Diploid DD Gene Sets

Builds the diploid DD gene set (`dd.lfsr.diploid.{ct} < 0.05`) for every
cell type at every `q`, and tabulates how the count changes with `q`.

```{r dip_dd_sets}
get_dd_genes <- function(df, ct) {
  col <- paste0("dd.lfsr.diploid.", ct)
  if (!col %in% colnames(df)) return(character(0))
  df$gene[!is.na(df[[col]]) & df[[col]] < 0.05]
}

# Named list: dip_dd_sets[[q]][[ct]] -> character vector of gene names
dip_dd_sets <- setNames(lapply(q_values, function(q) {
  setNames(lapply(celltypes_dip, function(ct) get_dd_genes(dd_tables[[q]], ct)), celltypes_dip)
}), q_values)

dip_dd_counts <- expand_grid(q = q_values, celltype = celltypes_dip) |>
  rowwise() |>
  mutate(n_dd = length(dip_dd_sets[[q]][[celltype]])) |>
  ungroup() |>
  mutate(celltype = pretty_ct(celltype))

dip_dd_counts |>
  pivot_wider(names_from = q, values_from = n_dd, names_prefix = "q") |>
  print(n = Inf)
```

### DD gene count vs. q, per cell type

Same table as above, plotted as a line per cell type so the direction and
size of the trend (almost always: more permissive q -> more DD genes) is
visible at a glance.

```{r dip_dd_count_plot, fig.width=10, fig.height=7}
ggplot(dip_dd_counts, aes(x = as.numeric(q), y = n_dd, group = celltype, colour = celltype)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Diploid DD gene count vs. q", x = "q", y = "Number of DD genes") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~celltype, scales = "free_y", ncol = 5)
```

---


## 3. Recapitulated DD Gene Sets

Same idea as Section 2, but for the recapitulated DD gene set: diploid DD
AND allotetraploid DD (`lfsr < 0.05` in both) with matching effect-size
sign, at every `q`.

```{r recap_dd_sets}
get_recap_genes <- function(df, ct) {
  lfsr_dip  <- paste0("dd.lfsr.diploid.", ct)
  lfsr_allo <- paste0("dd.lfsr.allotetraploid.", ct)
  lfc_dip   <- paste0("dd.logFC.diploid.", ct)
  lfc_allo  <- paste0("dd.logFC.allotetraploid.", ct)

  missing <- setdiff(c(lfsr_dip, lfsr_allo, lfc_dip, lfc_allo), colnames(df))
  if (length(missing) > 0) return(character(0))

  dd_flag   <- !is.na(df[[lfsr_dip]])  & df[[lfsr_dip]]  < 0.05
  allo_flag <- !is.na(df[[lfsr_allo]]) & df[[lfsr_allo]] < 0.05
  sign_flag <- sign(df[[lfc_dip]]) == sign(df[[lfc_allo]])

  df$gene[dd_flag & allo_flag & sign_flag]
}

recap_dd_sets <- setNames(lapply(q_values, function(q) {
  setNames(lapply(celltypes_recap, function(ct) get_recap_genes(dd_tables[[q]], ct)), celltypes_recap)
}), q_values)

recap_dd_counts <- expand_grid(q = q_values, celltype = celltypes_recap) |>
  rowwise() |>
  mutate(n_recap = length(recap_dd_sets[[q]][[celltype]])) |>
  ungroup() |>
  mutate(celltype = pretty_ct(celltype))

recap_dd_counts |>
  pivot_wider(names_from = q, values_from = n_recap, names_prefix = "q") |>
  print(n = Inf)
```

### Recapitulated DD gene count vs. q, per cell type

Same trend line as the diploid version above, for the recapitulated gene
set.

```{r recap_dd_count_plot, fig.width=10, fig.height=7}
ggplot(recap_dd_counts, aes(x = as.numeric(q), y = n_recap, group = celltype, colour = celltype)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Recapitulated DD gene count vs. q", x = "q", y = "Number of recapitulated DD genes") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~celltype, scales = "free_y", ncol = 5)
```

---


## 4. Retention of q = 0.2 DD Genes as q Decreases

Uses `q = 0.2` (the most permissive value, largest gene universe) as the
baseline. For each baseline DD gene, at each stricter q it falls into
exactly one of three outcomes:

- **Retained** — still significant DD (`lfsr < 0.05`) at that q
- **Lost — no longer significant** — the gene is still in that q's gene
  table (it was tested) but is no longer significant
- **Lost — dropped from smaller universe** — the gene isn't in that q's
  gene table at all (memento's filtering excluded it before any
  significance test could run)

These three percentages always sum to 100% of the baseline gene set, so
they're shown stacked on the same chart: how much of the loss going from
q = 0.2 down to q = 0.05 is "real" (no longer significant) vs. an
artifact of the shrinking gene universe.

```{r retention_setup}
baseline_q <- "0.2"
other_qs   <- setdiff(q_values, baseline_q)

# For one gene set (diploid or recapitulated) and one baseline q, classifies
# every baseline-q DD gene's fate at each stricter q into Retained / Lost
# (no longer significant) / Dropped (outside that q's gene universe), then
# reshapes to one row per (celltype, q) with both counts and percentages.
retention_breakdown <- function(sets_by_q, tables, celltypes, baseline_q, other_qs) {
  map_dfr(celltypes, function(ct) {
    baseline_genes <- sets_by_q[[baseline_q]][[ct]]
    n_baseline     <- length(baseline_genes)

    bind_rows(
      tibble(celltype = ct, q = baseline_q, n_baseline = n_baseline,
             n_retained = n_baseline, n_lost_not_sig = 0L, n_lost_dropped = 0L,
             retention_pct = 100, lost_not_sig_pct = 0, lost_dropped_pct = 0),
      map_dfr(other_qs, function(q) {
        universe_q      <- tables[[q]]$gene
        retained        <- intersect(baseline_genes, sets_by_q[[q]][[ct]])
        dropped         <- setdiff(baseline_genes, universe_q)
        lost_not_sig    <- setdiff(setdiff(baseline_genes, retained), dropped)

        tibble(
          celltype         = ct, q = q, n_baseline = n_baseline,
          n_retained       = length(retained),
          n_lost_not_sig   = length(lost_not_sig),
          n_lost_dropped   = length(dropped),
          retention_pct    = if (n_baseline > 0) 100 * length(retained)     / n_baseline else NA_real_,
          lost_not_sig_pct = if (n_baseline > 0) 100 * length(lost_not_sig) / n_baseline else NA_real_,
          lost_dropped_pct = if (n_baseline > 0) 100 * length(dropped)      / n_baseline else NA_real_
        )
      })
    )
  })
}

dip_retention   <- retention_breakdown(dip_dd_sets,   dd_tables, celltypes_dip,   baseline_q, other_qs) |>
  mutate(celltype = pretty_ct(celltype))
recap_retention <- retention_breakdown(recap_dd_sets, dd_tables, celltypes_recap, baseline_q, other_qs) |>
  mutate(celltype = pretty_ct(celltype))

celltype_order_by_n <- function(retention_df) {
  retention_df |> distinct(celltype, n_baseline) |> arrange(desc(n_baseline)) |> pull(celltype)
}

to_long <- function(retention_df) {
  ct_order <- celltype_order_by_n(retention_df)

  # Label each cell type's facet with its q=0.2 baseline gene count (e.g.
  # "Cardiomyocytes (n=45)") so the baseline denominator behind the % values
  # is visible directly on the plot.
  label_lookup <- retention_df |> distinct(celltype, n_baseline)
  label_map <- setNames(sprintf("%s (n=%d)", label_lookup$celltype, label_lookup$n_baseline),
                         label_lookup$celltype)
  ct_order_labeled <- unname(label_map[ct_order])

  retention_df |>
    select(celltype, q, retention_pct, lost_not_sig_pct, lost_dropped_pct) |>
    mutate(celltype = unname(label_map[celltype])) |>
    pivot_longer(cols = c(retention_pct, lost_not_sig_pct, lost_dropped_pct),
                 names_to = "status", values_to = "pct") |>
    mutate(status = recode(status,
             retention_pct    = "Retained",
             lost_not_sig_pct = "No Longer Significant",
             lost_dropped_pct = "Dropped"),
           status = factor(status, levels = c("Dropped",
                                               "No Longer Significant",
                                               "Retained")),
           q = factor(q, levels = q_values),
           # Facets ordered by number of baseline DD genes (descending), not alphabetically
           celltype = factor(celltype, levels = ct_order_labeled))
}

status_colours <- c("Retained"         = "#4575b4",
                     "No Longer Significant"         = "#fdae61",
                     "Dropped" = "grey70")

# Facet strip labels include a "(n=NNN)" suffix and, for several cell types,
# a long name (e.g. "Bronchiolar and alveolar epithelial cells") -- with
# ncol=5 each facet panel is narrow, so wrap onto at most two lines (width
# chosen so even the longest label -- "Bronchiolar and alveolar epithelial
# cells (n=NNNN)" -- fits in two) rather than letting ggplot truncate the
# label or wrap it across three or more lines.
wrap_ct_label <- function(x, width = 26) stringr::str_wrap(x, width = width)

cat("=== Diploid DD — retention breakdown relative to q=0.2 baseline ===\n")
dip_retention |> arrange(celltype, desc(as.numeric(q))) |> print(n = Inf)
```

### Number of diploid DD genes at q = 0.2 (baseline)

The baseline gene set size for each cell type -- the denominator behind
every retention percentage below. Cell types with a small baseline set will
naturally show noisier retention percentages than large ones.

```{r retention_baseline_bar, fig.width=9, fig.height=5}
dip_baseline_counts <- dip_retention |> filter(q == baseline_q) |> distinct(celltype, n_baseline)

ggplot(dip_baseline_counts, aes(x = reorder(celltype, -n_baseline), y = n_baseline)) +
  geom_col(fill = "#4575b4", width = 0.75) +
  labs(title = "Number of diploid DD genes at q = 0.2 (baseline)",
       x = NULL, y = "Number of DD genes") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

### Fate of q = 0.2 DD genes vs. q, per cell type (diploid DD)

The Retained / No Longer Significant / Dropped breakdown from the setup
chunk above, one stacked bar per q, faceted by cell type (ordered by
descending baseline gene count).

```{r retention_stack_dip, fig.width=11, fig.height=8}
dip_retention_plot <- ggplot(to_long(dip_retention), aes(x = q, y = pct, fill = status)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = status_colours) +
  labs(title = "",
       x = "q", y = "% of diploid DD genes (q=0.2)", fill = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8)) +
  facet_wrap(~celltype, ncol = 5, labeller = labeller(celltype = wrap_ct_label))
dip_retention_plot
# ggplot2::ggsave(filename = paste0(q_sensitivity_path, retention_fname("diploid", "0.2", "0.05")),
#                 plot = dip_retention_plot, device = "pdf", width = 14, height = 7, units = "in")
```

### Recapitulated DD — same analysis

The same two plots (baseline count, then Retained/Lost/Dropped fate) for
the recapitulated gene set instead of diploid DD.

```{r retention_baseline_bar_recap, fig.width=9, fig.height=5}
recap_baseline_counts <- recap_retention |> filter(q == baseline_q) |> distinct(celltype, n_baseline)

ggplot(recap_baseline_counts, aes(x = reorder(celltype, -n_baseline), y = n_baseline)) +
  geom_col(fill = "#d73027", width = 0.75) +
  labs(title = "Number of recapitulated DD genes at q = 0.2 (baseline)",
       x = NULL, y = "Number of DD genes") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

```{r retention_stack_recap, fig.width=11, fig.height=8}
allo_retention_plot <- ggplot(to_long(recap_retention), aes(x = q, y = pct, fill = status)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = status_colours) +
  labs(title = "",
       x = "q", y = "% of recapitulated DD genes (q=0.2)", fill = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8)) +
  facet_wrap(~celltype, ncol = 5, labeller = labeller(celltype = wrap_ct_label))
allo_retention_plot
# ggplot2::ggsave(filename = paste0(q_sensitivity_path, retention_fname("recap", "0.2", "0.05")),
# plot = allo_retention_plot, device = "pdf", width = 14, height = 7, units = "in")
```



### Same analysis, baseline q = 0.05

Same retention breakdown as above (reusing `retention_breakdown()`), but
with `q = 0.05` (the strictest value, smallest gene universe) as the
baseline instead of `q = 0.2`. Because the gene universe *grows* at more
permissive q, genes can also be called DD at a larger q that weren't even
part of the q=0.05 baseline set ("Gained"). Unlike the q=0.2 plots above,
these show the **nominal number of genes** on the y-axis (not %), stacked as
Retained / No Longer Significant / Gained — so the bar height is the size of
the union of the q=0.05 baseline set and each q's DD set, split by which of
the three categories each gene falls into. "Dropped" is omitted here since
q=0.05 is already the smallest universe, so it is expected to be
rare/absent.

```{r retention_setup_q005}
baseline_q_005 <- "0.05"
other_qs_005   <- setdiff(q_values, baseline_q_005)

dip_retention_q005   <- retention_breakdown(dip_dd_sets,   dd_tables, celltypes_dip,   baseline_q_005, other_qs_005) |>
  mutate(celltype = pretty_ct(celltype))
recap_retention_q005 <- retention_breakdown(recap_dd_sets, dd_tables, celltypes_recap, baseline_q_005, other_qs_005) |>
  mutate(celltype = pretty_ct(celltype))

# Genes called DD at a given (more permissive) q that were NOT in the q=0.05
# baseline DD set at all -- i.e. gained on top of the baseline, not part of
# its retained/lost breakdown (which sums to 100% of the baseline set).
genes_gained <- function(sets_by_q, celltypes, baseline_q, other_qs) {
  map_dfr(celltypes, function(ct) {
    baseline_genes <- sets_by_q[[baseline_q]][[ct]]
    n_baseline     <- length(baseline_genes)

    bind_rows(
      tibble(celltype = ct, q = baseline_q, n_gained = 0L, gained_pct = 0),
      map_dfr(other_qs, function(q) {
        gained <- setdiff(sets_by_q[[q]][[ct]], baseline_genes)
        tibble(celltype   = ct, q = q,
               n_gained   = length(gained),
               gained_pct = if (n_baseline > 0) 100 * length(gained) / n_baseline else NA_real_)
      })
    )
  })
}

dip_gained_q005   <- genes_gained(dip_dd_sets,   celltypes_dip,   baseline_q_005, other_qs_005) |>
  mutate(celltype = pretty_ct(celltype))
recap_gained_q005 <- genes_gained(recap_dd_sets, celltypes_recap, baseline_q_005, other_qs_005) |>
  mutate(celltype = pretty_ct(celltype))

cat("=== Diploid DD — retention breakdown relative to q=0.05 baseline ===\n")
dip_retention_q005 |> arrange(celltype, as.numeric(q)) |> print(n = Inf)

cat("\n=== Diploid DD — genes gained relative to q=0.05 baseline ===\n")
dip_gained_q005 |> arrange(celltype, as.numeric(q)) |> print(n = Inf)

# Nominal gene counts (not %), stacked as Retained / No Longer Significant /
# Gained. Bar height = size of the union of the q=0.05 baseline DD set and
# the current q's DD set.
gene_count_colours <- c("Retained"              = "#4575b4",
                         "No Longer Significant"  = "#fdae61",
                         "Gained"                = "#1a9850",
                         "Dropped"               = "grey70")

to_counts_long <- function(retention_df, gained_df = NULL) {
  ct_order <- celltype_order_by_n(retention_df)

  if (is.null(gained_df)) {
    # No "Gained" category -- appropriate for baseline q=0.2 (the most
    # permissive baseline), where every other q is stricter and Section 7's
    # original design only ever tracked the fate of the baseline genes
    # themselves: Retained / No Longer Significant / Dropped
    # (the last of which is non-trivial here, unlike for the q=0.05 baseline,
    # since stricter q's genuinely have smaller gene universes).
    out <- retention_df |>
      select(celltype, q, n_retained, n_lost_not_sig, n_lost_dropped)
    long_cols <- c("n_retained", "n_lost_not_sig", "n_lost_dropped")
    status_levels <- c("Dropped", "No Longer Significant", "Retained")
  } else {
    out <- retention_df |>
      select(celltype, q, n_retained, n_lost_not_sig) |>
      left_join(gained_df |> select(celltype, q, n_gained), by = c("celltype", "q"))
    long_cols <- c("n_retained", "n_lost_not_sig", "n_gained")
    status_levels <- c("No Longer Significant", "Gained", "Retained")
  }

  out |>
    pivot_longer(cols = all_of(long_cols), names_to = "status", values_to = "n_genes") |>
    mutate(status = recode(status,
             n_retained     = "Retained",
             n_lost_not_sig = "No Longer Significant",
             n_gained       = "Gained",
             n_lost_dropped = "Dropped"),
           status = factor(status, levels = status_levels),
           q = factor(q, levels = q_values),
           celltype = factor(celltype, levels = ct_order))
}
```

### Fate of q = 0.05 DD genes vs. q, per cell type (diploid DD)

Retained / No Longer Significant / Gained, as nominal gene counts, one
stacked bar per q, faceted by cell type.

```{r retention_stack_dip_q005, fig.width=11, fig.height=8}
dip_retention_plot_q005 <- ggplot(to_counts_long(dip_retention_q005, dip_gained_q005),
                                   aes(x = q, y = n_genes, fill = status)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = gene_count_colours) +
  labs(title = "",
       x = "q", y = "# of diploid DD genes", fill = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8)) +
  facet_wrap(~celltype, ncol = 5, labeller = labeller(celltype = wrap_ct_label))
dip_retention_plot_q005
# ggplot2::ggsave(filename = paste0(q_sensitivity_path, retention_fname("diploid", "0.05", "0.05")),plot = dip_retention_plot_q005, device = "pdf", width = 14, height = 7, units = "in")
```

### Recapitulated DD — same analysis (q = 0.05 baseline)

Same q=0.05-baseline fate breakdown, for the recapitulated gene set.

```{r retention_stack_recap_q005, fig.width=11, fig.height=8}
allo_retention_plot_q005 <- ggplot(to_counts_long(recap_retention_q005, recap_gained_q005),
                                    aes(x = q, y = n_genes, fill = status)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = gene_count_colours) +
  labs(title = "",
       x = "q", y = "# of recapitulated DD genes", fill = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8)) +
  facet_wrap(~celltype, ncol = 5, labeller = labeller(celltype = wrap_ct_label))
allo_retention_plot_q005
# ggplot2::ggsave(filename = paste0(q_sensitivity_path, retention_fname("recap", "0.05", "0.05")), plot = allo_retention_plot_q005, device = "pdf", width = 14, height = 7, units = "in")
```

---

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```

