---
title: "DD Interspecies Analysis"
output: html_document
date: "2026-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Results and Packages

```{r}
library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(Matrix)
library(purrr)
library(ggpubr)
library(stringr)
library(scales)
```

We load DD results.
```{r}
path_data <- '/project/gilad/awchen55/differentialDispersion/results/variability_paper/data/'
results_path <- '/project/gilad/awchen55/differentialDispersion/results/variability_paper/'

dd_results <- read.csv(paste0(path_data,"DD_interspecies_results_final.csv"))[,-1]
head(dd_results)

round_up_signif <- function(p, digits = 1) {
  
  if (p < 2e-16) {
    return("2 %*% 10^{-16}")
  }
  
  # round up
  if (p != 0) {
    expo  <- floor(log10(p))
    scale <- 10^(expo - digits + 1)
    p <- ceiling(p / scale) * scale
  }
  
  # force decimal formatting
  if (p >= 1e-4) {
    dec <- min(max(0, -floor(log10(p))), 4)
    return(formatC(p, format = "f", digits = dec))
  }
  
  # scientific notation
  expo <- floor(log10(p))
  base <- round(p / 10^expo, digits)
  
  paste0(base, " %*% 10^{", expo, "}")
}
```


## Summary Tables

### Number of DD Genes

```{r}

# vector of cell types
celltypes <- c( "Neuroblast","ENS.glia","Radial.glia","Neuronal.IPC","Glioblast")

summary_table <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col  <- paste0("dd.lfsr.diploid.", ct)
  recap_col <- paste0("dd.recapitulated.same.sign.", ct)

  dd_flag <- dd_results[[lfsr_col]] < 0.05

  dd_total <- sum(dd_flag, na.rm = TRUE)
  recap_total <- sum(dd_flag & dd_results[[recap_col]] == 1, na.rm = TRUE)

  tibble(
    CellType = gsub("\\.", " ", ct),
    `DD Genes` = dd_total,
    Recapitulated = recap_total,
    `% Recapitulated` = recap_total / dd_total
  )
})

summary_table
```

### Plot of DD Genes

```{r}

df_totals <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col  <- paste0("dd.lfsr.diploid.", ct)
  recap_col <- paste0("dd.recapitulated.same.sign.", ct)

  # DD genes in diploid
  dd_flag <- !is.na(dd_results[[lfsr_col]]) & dd_results[[lfsr_col]] < 0.05

  # Recapitulated among DD genes
  recap_flag <- dd_flag & !is.na(dd_results[[recap_col]]) & dd_results[[recap_col]] == 1

  n_dd    <- sum(dd_flag, na.rm = TRUE)
  n_recap <- sum(recap_flag, na.rm = TRUE)
  n_not   <- n_dd - n_recap

  tibble::tibble(
    CellType = ct,
    Not_Recapitulated = n_not,
    Recapitulated     = n_recap
  )
}) %>%
  dplyr::mutate(
    CellType_label = gsub("\\.", " ", CellType),
    CellType_label = factor(
      CellType_label,
      levels = gsub("\\.", " ", celltypes)
    )
  ) %>%
  tidyr::pivot_longer(
    cols = c(Not_Recapitulated, Recapitulated),
    names_to = "List",
    values_to = "DV_Genes"
  ) %>%
  dplyr::mutate(
    List = factor(List, levels = c("Not_Recapitulated", "Recapitulated"))
  )

dd_gene_plot <- ggplot2::ggplot(
  df_totals,
  ggplot2::aes(x = CellType_label, y = DV_Genes, fill = List)
) +
  ggplot2::geom_col() +
  ggplot2::scale_y_continuous(
    limits = c(0, 920),
    breaks = c(0, 200, 400, 600, 800),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    x = "",
    y = "# DD genes (LFSR < 0.05)",
    title = ""
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 15, face = "bold", hjust = 0.5),
    panel.grid = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.4),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.8),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12),
    legend.position = c(0.01, 0.99),
    legend.justification = c("left", "top"),
    legend.title = ggplot2::element_blank(),
    legend.background = ggplot2::element_rect(fill = NA, color = NA),
    legend.text = ggplot2::element_text(size = 12),
    legend.key.size = grid::unit(0.4, "cm")
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Not_Recapitulated" = "gray87",
      "Recapitulated"     = "lightskyblue"
    ),
    labels = c(
      "Not_Recapitulated" = "Not Recapitulated",
      "Recapitulated"     = "Recapitulated"
    )
  )

dd_gene_plot
#ggplot2::ggsave(filename = paste0(results_path, "dd_genes.pdf"), plot = dd_gene_plot,  device = "pdf", width = 7, height = 5, units = "in")
```

## Examples of DD Genes

We load the cell annotated data.
```{r}
path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'

diploid_crossfilt_data <- readRDS(paste0(path,'parent_crossfilt_celltypes_of_interest.rds'))
```


### DD but not DE

Normalize cell counts.
```{r}
ct = "Radial glia"
# Subset to Radial glia cells only
ct_obj <- subset(diploid_crossfilt_data, subset = celltype == ct)

# Recompute CPM on this subset
counts <- GetAssayData(ct_obj, assay = "RNA", slot = "counts")
cpm_mat <- edgeR::cpm(counts, log = TRUE, prior.count = 0)
```

Plot distribution of expression.
```{r}
gene <- "DNAH14"
ct   <- "Radial.glia"

de_col <- paste0("de.adj.pvalue.diploid.", ct)
dd_col <- paste0("dd.lfsr.diploid.", ct)

vals <- dd_results %>%
  dplyr::filter(gene == !!gene) %>%
  dplyr::select(
    de = !!rlang::sym(de_col),
    dd = !!rlang::sym(dd_col)
  ) %>%
  dplyr::slice(1)

plot_df <- data.frame(
  expr    = as.numeric(cpm_mat[gene, ]),
  species = ct_obj$species
)

# rounding helpers
round_down <- function(x, digits = 1) {
  floor(x * 10^digits) / 10^digits
}

round_up <- function(x, digits = 2) {
  ceiling(x * 10^digits) / 10^digits
}

de_label <- format(round_down(vals$de, 1), nsmall = 1)
dd_label <- format(round_up(vals$dd, 2), nsmall = 2)

# plotmath label
label_expr <- bquote(
  atop(
    bold("DE:") ~ "FDR > " ~ .(de_label),
    bold("DD:") ~ "LFSR < " ~ .(dd_label)
  )
)

x_pos <- 9.49
y_pos <- Inf

dd_ex1_plot <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = expr, color = species, fill = species)
) +
  ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::scale_fill_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::labs(
    x = "Expression (log2 CPM)",
    y = "Density",
    title = bquote(italic(.(gene)))
  ) +
  ggplot2::annotate(
    "label",
    x = x_pos, y = y_pos,
    label = as.character(as.expression(label_expr)),
    parse = TRUE,
    hjust = 0,
    vjust = 1.25,
    size = 4,
    label.size = 0.3,
    label.padding = grid::unit(0.25, "lines"),
    fill = "white",
    color = "black"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title   = ggplot2::element_text(size = 14),
    axis.title   = ggplot2::element_text(size = 14),
    axis.text    = ggplot2::element_text(size = 12),
    axis.line    = ggplot2::element_blank(),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.text  = ggplot2::element_text(size = 12),
    legend.title = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(title = NULL),
    fill  = ggplot2::guide_legend(title = NULL)
  )

dd_ex1_plot
#ggsave(paste0(results_path, 'dd_gene_ex_dd.pdf'), plot = dd_ex1_plot, device = "pdf", width = 7, height = 5, units = "in")
```

### DD and DE Opposite Signs

Normalize cell counts.
```{r}
ct = "Glioblast"
# Subset to glioblast cells only
ct_obj <- subset(diploid_crossfilt_data, subset = celltype == ct)

# Recompute CPM on this subset
counts <- GetAssayData(ct_obj, assay = "RNA", slot = "counts")
cpm_mat <- edgeR::cpm(counts, log = TRUE, prior.count = 0)
```

Plot distribution of expression.
```{r}
gene <- "CCT3"
ct   <- "Glioblast"

de_col <- paste0("de.adj.pvalue.diploid.", ct)
dd_col <- paste0("dd.lfsr.diploid.", ct)

vals <- dd_results %>%
  dplyr::filter(gene == !!gene) %>%
  dplyr::select(
    de = !!rlang::sym(de_col),
    dd = !!rlang::sym(dd_col)
  ) %>%
  dplyr::slice(1)

plot_df <- data.frame(
  expr    = as.numeric(cpm_mat[gene, ]),
  species = ct_obj$species
)

# rounding helpers
round_down <- function(x, digits = 1) {
  floor(x * 10^digits) / 10^digits
}

round_up <- function(x, digits = 2) {
  ceiling(x * 10^digits) / 10^digits
}

de_label <- format(round_up(vals$de, 2), nsmall = 2)
dd_label <- format(round_up(vals$dd, 2), nsmall = 2)

# plotmath label
label_expr <- bquote(
  atop(
    bold("DE:") ~ "FDR < " ~ .(de_label),
    bold("DD:") ~ "LFSR < " ~ .(dd_label)
  )
)

x_pos <- 9.38
y_pos <- Inf

dd_ex2_plot <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = expr, color = species, fill = species)
) +
  ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::scale_fill_manual(
    values = c("human" = "lightcoral", "chimp" = "turquoise3"),
    breaks = c("human", "chimp"),
    labels = c("human" = "human", "chimp" = "chimpanzee")
  ) +
  ggplot2::labs(
    x = "Expression (log2 CPM)",
    y = "Density",
    title = bquote(italic(.(gene)))
  ) +
  ggplot2::annotate(
    "label",
    x = x_pos, y = y_pos,
    label = as.character(as.expression(label_expr)),
    parse = TRUE,
    hjust = 0,
    vjust = 1.25,
    size = 4,
    label.size = 0.3,
    label.padding = grid::unit(0.25, "lines"),
    fill = "white",
    color = "black"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title   = ggplot2::element_text(size = 14),
    axis.title   = ggplot2::element_text(size = 14),
    axis.text    = ggplot2::element_text(size = 12),
    axis.line    = ggplot2::element_blank(),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.text  = ggplot2::element_text(size = 12),
    legend.title = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(title = NULL),
    fill  = ggplot2::guide_legend(title = NULL)
  )

dd_ex2_plot
#ggsave(paste0(results_path, 'dd_gene_ex_de2.pdf'), plot = dd_ex2_plot, device = "pdf", width = 7, height = 5, units = "in")
```


## Effect Size Plot

We plot DD effect sizes of human vs chimpanzee comparisons. This includes all cell types together. Light grey represents DD genes in the diploid (LFSR < 0.05) and black shows the subset recapitulated in the allotetraploid (LFSR < 0.05 and same sign of effect).
```{r}
plot_df <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col   <- paste0("dd.lfsr.diploid.", ct)
  same_col   <- paste0("dd.recapitulated.same.sign.", ct)
  dip_col    <- paste0("dd.logFC.diploid.", ct)
  allo_col   <- paste0("dd.logFC.allotetraploid.", ct)

  dd_results %>%
    dplyr::select(
      gene,
      all_of(lfsr_col),
      all_of(same_col),
      all_of(dip_col),
      all_of(allo_col)
    ) %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      sig_dd = .data[[lfsr_col]] < 0.05,
      same_sign = as.integer(.data[[same_col]]),   # 1 = black, 0/NA = grey
      diploid_logFC = as.numeric(.data[[dip_col]]),
      allotetraploid_logFC = as.numeric(.data[[allo_col]])
    ) %>%
    dplyr::filter(
      sig_dd,
      !is.na(diploid_logFC),
      !is.na(allotetraploid_logFC)
    ) %>%
    dplyr::mutate(
      same_sign = dplyr::if_else(is.na(same_sign), 0L, same_sign)
    )
}) %>%
  dplyr::distinct(
    gene, celltype, same_sign, diploid_logFC, allotetraploid_logFC
  )

# -----------------------------
# Quadrant percentages
# if you want only black dots, add:
# %>% dplyr::filter(same_sign == 1)
# -----------------------------
quad_df <- plot_df %>%
  dplyr::mutate(
    quadrant = dplyr::case_when(
      allotetraploid_logFC >= 0 & diploid_logFC >= 0 ~ "TR",
      allotetraploid_logFC <  0 & diploid_logFC >= 0 ~ "TL",
      allotetraploid_logFC <  0 & diploid_logFC <  0 ~ "BL",
      TRUE                                           ~ "BR"
    )
  ) %>%
  dplyr::count(quadrant, name = "n") %>%
  dplyr::mutate(pct = 100 * n / sum(n))

# -----------------------------
# Label positions
# -----------------------------
xlim <- c(-2.1, 2.1)
ylim <- c(-2.0, 2.0)

label_pos <- tibble::tibble(
  quadrant = c("TL", "TR", "BL", "BR"),
  x = c(xlim[1] + 0.12, xlim[2] - 0.12, xlim[1] + 0.12, xlim[2] - 0.12),
  y = c(ylim[2] - 0.12, ylim[2] - 0.12, ylim[1] + 0.12, ylim[1] + 0.12),
  hjust = c(0, 1, 0, 1),
  vjust = c(1, 1, 0, 0)
)

quad_labels <- label_pos %>%
  dplyr::left_join(quad_df, by = "quadrant") %>%
  dplyr::mutate(
    label = ifelse(is.na(pct), "0%", paste0(sprintf("%.1f", pct), "%"))
  )

# -----------------------------
# Shortened y = x segment
# -----------------------------
lim <- range(
  c(plot_df$allotetraploid_logFC, plot_df$diploid_logFC),
  na.rm = TRUE
)
short_lim <- as.numeric(
  stats::quantile(lim, probs = c(0.1, 0.75), na.rm = TRUE)
)



  pad <- 0.15
  
  blue_start <- short_lim[1] + pad
  blue_end   <- short_lim[2] - pad
  
  red_bottom <- ylim[1] + pad
  red_top    <- ylim[2] - pad
  
# -----------------------------
# Plot
# -----------------------------
p_effect_size <- ggplot(plot_df,
                        aes(x = allotetraploid_logFC, y = diploid_logFC)) +

  # grey points
  geom_point(
    data = dplyr::filter(plot_df, same_sign != 1),
    color = "grey75",
    alpha = 0.8,
    size = 1
  ) +

  # black points
  geom_point(
    data = dplyr::filter(plot_df, same_sign == 1),
    color = "black",
    alpha = 0.7,
    size = 1
  ) +

  # quadrant labels
  geom_label(
    data = quad_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = quad_labels$hjust,
    vjust = quad_labels$vjust,
    size = 4,
    label.size = 0.25,
    fill = "white",
    color = "black"
  ) +

  scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2),
    expand = expansion(mult = 0.02)
  ) +
  scale_y_continuous(
    breaks = c(-2, -1, 0, 1, 2),
    expand = expansion(mult = 0.02)
  ) +

  coord_equal(xlim = xlim, ylim = ylim, clip = "on") +

  # RED x = 0 vertical line (shortened)
  geom_segment(
    aes(
      x = 0,
      xend = 0,
      y = red_bottom,
      yend = red_top
    ),
    inherit.aes = FALSE,
    color = "red",
    linewidth = 0.7,
    arrow = arrow(length = unit(0.18, "cm"), ends = "both", type = "closed")
  ) +

  # BLUE y = x line (shortened)
  geom_segment(
    aes(
      x = blue_start,
      y = blue_start,
      xend = blue_end,
      yend = blue_end
    ),
    inherit.aes = FALSE,
    color = "blue",
    linewidth = 0.6,
    arrow = arrow(length = unit(0.18, "cm"), ends = "both", type = "closed")
  ) +

  labs(
    x = "logFC human/chimpanzee in allotetraploid cells",
    y = "logFC human/chimpanzee"
  ) +
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor = element_line(color = "grey90", linewidth = 0.25),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.line = element_blank(),
    axis.title = element_text(size = 10),
    axis.text  = element_text(size = 8)
  )

p_effect_size
#ggsave(paste0(results_path, 'cis_trans_effect_sizes.pdf'), plot = p_effect_size, device = "pdf", width = 7, height = 5, units = "in")
```


## Cis vs Trans Analysis

### Number of cis/trans DD genes

Number of DD genes for each cell type.The overall category using mean cis proportion of cell types DD.
```{r}
# -----------------------------
# Per-cell-type counts
# -----------------------------
celltype_counts <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.category.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.cis.category."),
    names_to = "celltype",
    values_to = "category"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.cis\\.category\\."),
    celltype = stringr::str_replace_all(celltype, "\\.", " "),
    category = factor(category, levels = c("cis", "neither", "trans", "mixed", "not DD"))
  ) %>%
  dplyr::count(celltype, category, .drop = FALSE) %>%
  tidyr::pivot_wider(
    names_from = category,
    values_from = n,
    values_fill = 0
  )
celltype_counts

```



### Effect size comparisons

We compare the DD effect sizes in diploid cells for all cell types together. These meet our cutoff of > 0.7 as cis and < 0.3 is trans.
```{r}
# -------------------------------
# 1. Cis category → long
# -------------------------------
cat_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.category.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.cis.category."),
    names_to = "celltype",
    values_to = "category"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.cis\\.category\\.")
  )

# -------------------------------
# 2. Effect size → long
# -------------------------------
eff_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.logFC.diploid.")) %>%
  tidyr::pivot_longer(
    cols = starts_with("dd.logFC.diploid."),
    names_to = "celltype",
    values_to = "effect_size"
  ) %>%
  dplyr::mutate(
    celltype = stringr::str_remove(celltype, "^dd\\.logFC\\.diploid\\.")
  )

# -------------------------------
# 3. Combine (gene × cell type instances)
# -------------------------------
ecdf_df <- cat_long %>%
  dplyr::inner_join(eff_long, by = c("gene", "celltype")) %>%
  dplyr::filter(celltype %in% celltypes) %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::mutate(
    category = factor(category, levels = c("cis", "neither", "trans"))
  )

ecdf_df_plot <- ecdf_df %>%
  dplyr::filter(category %in% c("cis", "trans")) %>%
  dplyr::mutate(
    category = factor(category, levels = c("cis", "trans"))
  )

ecdf_plot <- ggplot2::ggplot(
  ecdf_df_plot,
  ggplot2::aes(x = abs(effect_size), color = category)
) +
  ggplot2::stat_ecdf(linewidth = 1) +
  ggplot2::labs(
    x = "Absolute logFC",
    y = "Cumulative frequency",
    color = ""
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "cis"   = "#009E73",
      "trans" = "#D55E00"
    ),
    labels = c(
      cis   = expression(italic("cis")),
      trans = expression(italic("trans"))
    )
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(color = "black", linewidth = 1),
    panel.grid.major = ggplot2::element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = c(0.9, 0.05),
    legend.justification = c("right", "bottom"),
    legend.background = ggplot2::element_rect(
      fill = scales::alpha("white", 0.7),
      color = NA
    ),
    legend.key = ggplot2::element_blank(),
    legend.key.height = grid::unit(0.9, "cm"),
    legend.key.width  = grid::unit(1.2, "cm"),
    legend.text  = ggplot2::element_text(size = 14),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12)
  )

ecdf_plot
#ggsave(paste0(results_path, 'ecdf_cis_trans.pdf'), plot = ecdf_plot, device = "pdf", width = 7, height = 5, units = "in")
```

We use a K-S test to test for signficant differences in the effect sizes.
```{r}
# subset effect sizes
cis_vals <- ecdf_df_plot %>%
  dplyr::filter(category == "cis") %>%
  dplyr::pull(abs(effect_size))

trans_vals <- ecdf_df_plot %>%
  dplyr::filter(category == "trans") %>%
  dplyr::pull(abs(effect_size))

# KS test
ks_res <- ks.test(cis_vals, trans_vals)

ks_res

```

### Mean cis proportion of DD genes

We compare the mean cis proportion of cell types that are DD to the number of times a gene is DD in a cell type. This includes all genes that are DD in the diploid.
```{r}
# ----------------------------------
# Build long dataframe: one row per gene x cell type
# ----------------------------------
plot_df_long <- purrr::map_dfr(celltypes, function(ct) {

  lfsr_col <- paste0("dd.lfsr.diploid.", ct)
  cis_col  <- paste0("dd.cis.proportion.", ct)

  dd_results %>%
    dplyr::transmute(
      gene = gene,
      celltype = ct,
      dd_lfsr = .data[[lfsr_col]],
      cis_prop = .data[[cis_col]]
    )
})

# ----------------------------------
# Keep only diploid-DD gene/celltype pairs
# ----------------------------------
plot_df_long_sig <- plot_df_long %>%
  dplyr::filter(!is.na(dd_lfsr), dd_lfsr < 0.05, !is.na(cis_prop))

# ----------------------------------
# Collapse to gene-level summary
# ----------------------------------
plot_df <- plot_df_long_sig %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    n_ct = dplyr::n(),
    mean_cis_prop = mean(cis_prop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    n_ct = factor(n_ct, levels = sort(unique(n_ct)))
  )

# ----------------------------------
# Compute p-value (1 vs 5)
# ----------------------------------
p_val <- wilcox.test(
  mean_cis_prop ~ n_ct,
  data = plot_df %>% dplyr::filter(n_ct %in% c("1", "5"))
)$p.value

p_label <- paste0(
  "p < ",
  round_up_signif(p_val)
)

# ----------------------------------
# Bracket positioning
# ----------------------------------
y_max <- max(plot_df$mean_cis_prop, na.rm = TRUE)
h     <- 0.03
y_pos <- y_max + 0.055

# ----------------------------------
# Plot
# ----------------------------------
p_mean_cis_diploid <- ggpubr::ggboxplot(
  plot_df,
  x = "n_ct",
  y = "mean_cis_prop",
  fill = "n_ct",
  color = "black",
  palette = "Set2",
  add = "jitter",
  add.params = list(width = 0.15, alpha = 0.5, size = 1)
) +
  annotate("segment", x = 1, xend = 5, y = y_pos, yend = y_pos) +
  annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = 5, xend = 5, y = y_pos, yend = y_pos - h) +
  annotate("text", x = 3, y = y_pos + h, label = p_label, size = 4) +
  labs(
    x = "Number of cell types DD in diploid",
    y = expression(paste("Mean ", italic("cis"), " proportion"))
  ) +
  ggpubr::theme_pubr() +
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.line = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  ) +
  coord_cartesian(
    ylim = c(0, y_pos + 0.05),
    clip = "off"
  )

p_mean_cis_diploid
#ggsave(paste0(results_path, 'mean_cis_by_n_celltype_dd.pdf'), plot = p_mean_cis_diploid, device = "pdf", width = 7, height = 5, units = "in")
```

### LOEUF

Load LOF data.
```{r}
# Load data to get LOF scores
path_lof <- '/project/gilad/brendan/dispersion/pilot/cHDC_data/cellranger_cluster-mode_trial/analysis/datasets/evo_constraint/'
lof_scores <- read.table(paste0(path_lof,'gnomad.v4.1.constraint_metrics.tsv'), header = TRUE)
lof_scores <- lof_scores %>% filter(mane_select == "true")
```

We compare the loss of function in cis and trans genes. We use our overall definition of cis/trans genes by looking at mean cis proportion of cell type DD and determine cis (>0.7) or trans (<0.3).
```{r}
###### Plot
plot_df2 <- dd_results %>%
  dplyr::select(gene, dd.cis.category.overall) %>%
  dplyr::inner_join(
    lof_scores %>%
      dplyr::select(gene, lof.oe),
    by = "gene"
  ) %>%
  dplyr::mutate(
    cat = factor(
      as.character(dd.cis.category.overall),
      levels = c("cis", "neither", "mixed", "trans", "not DD")
    )
  )

# Wilcoxon: cis vs trans
pval <- plot_df2 %>%
  dplyr::filter(cat %in% c("cis", "trans")) %>%
  dplyr::summarise(p = wilcox.test(lof.oe ~ cat)$p.value) %>%
  dplyr::pull(p)

if (pval < 2e-16) {
  p_label <- "p < 2e-16"
} else {
  p_up <- round_up_signif(pval, digits = 1)
  p_label <- paste0(
    "p < ",
    round_up_signif(pval)
  )
}

y_pos <- min(1.65, max(plot_df2$lof.oe, na.rm = TRUE) * 1.12)
h <- 0.05

p_lof <- ggplot(plot_df2, aes(x = cat, y = lof.oe, fill = cat)) +
  geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

  # bracket spanning cis (1) to trans (4)
  annotate("segment", x = 1, xend = 4, y = y_pos, yend = y_pos) +
  annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = 4, xend = 4, y = y_pos, yend = y_pos - h) +
  annotate("text", x = 2.5, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +

  scale_fill_manual(values = c(
    "cis"     = "#009E73",
    "trans"   = "#D55E00",
    "neither" = "#0072B2",
    "mixed"   = "#CC79A7",
    "not DD"  = "grey70"
  )) +
  labs(x = "", y = "LOEUF") +
  ggpubr::theme_pubr() +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  coord_cartesian(ylim = c(0, 1.7)) +
  scale_x_discrete(labels = c(
    "cis"     = expression(italic("cis")),
    "trans"   = expression(italic("trans")),
    "neither" = "neither",
    "mixed"   = "mixed",
    "not DD"  = "not DD"
  ))

p_lof
#ggsave(paste0(results_path, 'lof_cis_trans.pdf'), plot = p_lof, device = "pdf", width = 7, height = 5, units = "in")
```



#### Supplementary varying cis/trans cutoff

Supplementary using mean cis proportion of cell types DD.
```{r}
lof_cutoff <- function(dd_results, lof_scores, cis_cutoff, trans_cutoff) {

  plot_df2 <- dd_results %>%
    dplyr::select(gene, dd.in.diploid.mean.cis.proportion) %>%
    dplyr::mutate(
      combined_category = dplyr::case_when(
        is.na(dd.in.diploid.mean.cis.proportion) ~ "not DD",
        dd.in.diploid.mean.cis.proportion > cis_cutoff   ~ "cis",
        dd.in.diploid.mean.cis.proportion < trans_cutoff ~ "trans",
        TRUE                                                ~ "neither"
      )
    ) %>%
    dplyr::inner_join(
      lof_scores %>%
        dplyr::select(gene, lof.oe),
      by = "gene"
    ) %>%
    dplyr::mutate(
      cat = factor(
        combined_category,
        levels = c("cis", "neither", "trans", "not DD")
      )
    )

  # Wilcoxon: cis vs trans
  pval <- plot_df2 %>%
  dplyr::filter(cat %in% c("cis", "trans")) %>%
  dplyr::summarise(p = wilcox.test(lof.oe ~ cat)$p.value) %>%
  dplyr::pull(p)

  p_label <- paste0(
    "p < ",
    round_up_signif(pval)
  )

  y_pos <- min(1.75, max(plot_df2$lof.oe, na.rm = TRUE) * 1.12)
  h <- 0.05

p_lof <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = lof.oe, fill = cat)) +
  ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

  # bracket spanning cis (1) to trans (3)
  ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
  ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
  ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +

  ggplot2::scale_fill_manual(values = c(
    "cis"     = "#009E73",
    "trans"   = "#D55E00",
    "neither" = "#0072B2",
    "not DD"  = "grey70"
  )) +
  ggplot2::labs(x = "", y = "LOEUF") +
  ggpubr::theme_pubr() +
  ggplot2::theme(
    axis.line = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor.x = ggplot2::element_blank(),
    legend.position = "none",
    axis.title = ggplot2::element_text(size = 14),
    axis.text = ggplot2::element_text(size = 12)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 1.8)) +
  ggplot2::scale_x_discrete(labels = c(
    "cis"     = expression(italic("cis")),
    "trans"   = expression(italic("trans")),
    "neither" = "neither",
    "not DD"  = "not DD"
  ))

  # add p-value label conditionally
  if (pval >= 1e-4) {
    p_lof <- p_lof +
      ggplot2::annotate(
        "text",
        x = 2,
        y = y_pos + h,
        label = p_label,
        size = 4
      )
  } else {
    p_lof <- p_lof +
      ggplot2::annotate(
        "text",
        x = 2,
        y = y_pos + h,
        label = p_label,
        size = 4,
        parse = TRUE
      )
  }

return(p_lof)
}
```

```{r}
p_0.6_0.4lof <-lof_cutoff(dd_results,lof_scores,0.6,0.4)
p_0.65_0.35lof <-lof_cutoff(dd_results,lof_scores,0.65,0.35)
p_0.7_0.3lof <-lof_cutoff(dd_results,lof_scores,0.7,0.3)
p_0.75_0.25lof <-lof_cutoff(dd_results,lof_scores,0.75,0.25)
p_0.8_0.2lof <-lof_cutoff(dd_results,lof_scores,0.8,0.2)
p_0.85_0.15lof <-lof_cutoff(dd_results,lof_scores,0.85,0.15)

#p_0.6_0.4lof + p_0.65_0.35lof + p_0.7_0.3lof + p_0.75_0.25lof + p_0.8_0.2lof + p_0.85_0.15lof
```

Supplementary using mixed definition.
```{r}

lof_cutoff <- function(dd_results, lof_scores, celltypes3, cis_cutoff, trans_cutoff) {

  # step 1: create per-cell-type cis/trans/neither/not DD categories
  dd_results2 <- dd_results %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(paste0("dd.cis.proportion.", celltypes3)),
        ~ {
          ct <- sub("^dd\\.cis\\.proportion\\.", "", dplyr::cur_column())
          dd_col <- paste0("dd.lfsr.diploid.", ct)

          if (!dd_col %in% names(dd_results)) {
            return(rep(NA_character_, length(.x)))
          }

          dd_flag <- dd_results[[dd_col]]

          out <- rep("not DD", length(.x))
          is_dd <- !is.na(dd_flag) & dd_flag < 0.05 & !is.na(.x)

          out[is_dd & .x > cis_cutoff] <- "cis"
          out[is_dd & .x < trans_cutoff] <- "trans"
          out[is_dd & .x >= trans_cutoff & .x <= cis_cutoff] <- "neither"

          out
        },
        .names = "dd.cis.category.{sub('^dd\\\\.cis\\\\.proportion\\\\.', '', .col)}"
      )
    )

  # step 2: only use per-cell-type category columns
  cat_cols <- paste0("dd.cis.category.", celltypes3)

  # step 3: create overall category (same logic as master_df_filtered)
  dd_results2 <- dd_results2 %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      dd.cis.category.overall = {
        vals <- dplyr::c_across(dplyr::all_of(cat_cols))
        vals <- vals[!is.na(vals)]

        has_cis     <- "cis" %in% vals
        has_trans   <- "trans" %in% vals
        has_neither <- "neither" %in% vals

        if (!has_cis & !has_trans & !has_neither) {
          "not DD"
        } else if (has_cis & has_trans) {
          "mixed"
        } else if (has_cis) {
          "cis"
        } else if (has_trans) {
          "trans"
        } else if (has_neither) {
          "neither"
        } else {
          "not DD"
        }
      }
    ) %>%
    dplyr::ungroup()

  # step 4: prepare plotting df
  plot_df2 <- dd_results2 %>%
    dplyr::select(gene, dd.cis.category.overall) %>%
    dplyr::inner_join(
      lof_scores %>% dplyr::select(gene, lof.oe),
      by = "gene"
    ) %>%
    dplyr::mutate(
      cat = factor(
        as.character(dd.cis.category.overall),
        levels = c("cis", "mixed", "trans", "neither", "not DD")
      )
    )

  # step 5: Wilcoxon test (cis vs trans only)
  pval <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans")) %>%
    dplyr::summarise(p = wilcox.test(lof.oe ~ cat)$p.value) %>%
    dplyr::pull(p)

  p_label <- paste0("p < ", round_up_signif(pval))

  y_pos <- min(1.75, max(plot_df2$lof.oe, na.rm = TRUE) * 1.12)
  h <- 0.05

  # step 6: plot
  p_lof <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = lof.oe, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    # bracket (cis = 1, trans = 3 in new order)
    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "mixed"   = "#CC79A7",
      "trans"   = "#D55E00",
      "neither" = "#0072B2",
      "not DD"  = "grey70"
    )) +

    ggplot2::labs(x = "", y = "LOEUF") +
    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 1),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 14),
      axis.text = ggplot2::element_text(size = 12)
    ) +
    ggplot2::coord_cartesian(ylim = c(0, 1.8)) +
    ggplot2::scale_x_discrete(labels = c(
      "cis"     = expression(italic("cis")),
      "mixed"   = "mixed",
      "trans"   = expression(italic("trans")),
      "neither" = "neither",
      "not DD"  = "not DD"
    ))

  # add p-value label
  if (pval >= 1e-4) {
    p_lof <- p_lof +
      ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4)
  } else {
    p_lof <- p_lof +
      ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4, parse = TRUE)
  }

  return(p_lof)
}
```

```{r}
p_0.6_0.4lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.6,0.4)
p_0.65_0.35lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.65,0.35)
p_0.7_0.3lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.7,0.3)
p_0.75_0.25lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.75,0.25)
p_0.8_0.2lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.8,0.2)
p_0.85_0.15lof <-lof_cutoff(dd_results,lof_scores,celltypes3,0.85,0.15)

# p_0.6_0.4lof + p_0.65_0.35lof + p_0.7_0.3lof + p_0.75_0.25lof + p_0.8_0.2lof + p_0.85_0.15lof
```

### Percent coding identity

We load percent coding identity. We take more conservative estimate of coding divergence by taking the minimum difference between human to chimp and chimp to human.
```{r}
mart <- read.delim(
  gzfile("/project2/gilad/kenneth/Analysis/mart_export.txt.gz"),
  header = TRUE,
  stringsAsFactors = FALSE
)
mart <- mart %>%
  dplyr::mutate(
    percent_identity = pmin(
      as.numeric(X.id..query.gene.identical.to.target.Chimpanzee.gene),
      as.numeric(X.id..target.Chimpanzee.gene.identical.to.query.gene),
      na.rm = TRUE
    )
  )
```

Plot by cis/trans categories.
```{r}
plot_df <- dd_results %>%
  dplyr::select(gene, dd.cis.category.overall) %>%
  dplyr::left_join(
    mart %>%
      dplyr::select(Gene.name, percent_identity) %>%
      dplyr::rename(gene = Gene.name) %>%
      dplyr::mutate(percent_identity = as.numeric(percent_identity)),
    by = "gene"
  )

plot_df2 <- plot_df %>%
  dplyr::mutate(
    cat = factor(
      as.character(dd.cis.category.overall),
      levels = c("cis",  "neither","mixed", "trans", "not DD")
    )
  ) %>%
  dplyr::filter(!is.na(percent_identity), is.finite(percent_identity))

# cis vs trans only for significance
cis_trans <- plot_df2 %>%
  dplyr::filter(cat %in% c("cis", "trans")) %>%
  dplyr::mutate(
    cat = droplevels(cat)
  )

pval <- wilcox.test(percent_identity ~ cat, data = cis_trans)$p.value
p_label <- paste0("p < ", round_up_signif(pval))

ymin <- 80
ymax <- 103

y_pos <- min(ymax - 1, max(plot_df2$percent_identity, na.rm = TRUE) * 1.05)
h <- (ymax - ymin) * 0.03

p_pci <- ggplot(plot_df2, aes(x = cat, y = percent_identity, fill = cat)) +
  geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +
  annotate("segment", x = 1, xend = 4, y = y_pos, yend = y_pos) +
  annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = 4, xend = 4, y = y_pos, yend = y_pos - h) +
  annotate("text", x = 2.5, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +
  scale_fill_manual(values = c(
    "cis"     = "#009E73",
    "mixed"   = "#CC79A7",
    "trans"   = "#D55E00",
    "neither" = "#0072B2",
    "not DD"  = "grey70"
  )) +
  labs(x = "", y = "Percent coding identity") +
  ggpubr::theme_pubr() +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 1, fill = NA),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12)
  ) +
  coord_cartesian(ylim = c(ymin, ymax)) +
  scale_x_discrete(labels = c(
    "cis"     = expression(italic("cis")),
    "neither" = "neither",
    "mixed"   = "mixed",
    "trans"   = expression(italic("trans")),
    "not DD"  = "not DD"
  ))

p_pci
#ggsave(paste0(results_path, 'pci_cis_trans.pdf'), plot = p_pci, device = "pdf", width = 7, height = 5, units = "in")
```

#### Supplementary varying cis/trans

Using mean definition.

```{r}
pci_cutoff <- function(dd_results, mart, cis_cutoff, trans_cutoff) {

  plot_df <- dd_results %>%
    dplyr::select(gene, dd.in.diploid.mean.cis.proportion) %>%
    dplyr::mutate(
      combined_category = dplyr::case_when(
        is.na(dd.in.diploid.mean.cis.proportion) ~ "not DD",
        dd.in.diploid.mean.cis.proportion > cis_cutoff   ~ "cis",
        dd.in.diploid.mean.cis.proportion < trans_cutoff ~ "trans",
        TRUE                                             ~ "neither"
      )
    ) %>%
    dplyr::left_join(
      mart %>%
        dplyr::select(Gene.name, percent_identity) %>%
        dplyr::rename(gene = Gene.name) %>%
        dplyr::mutate(percent_identity = as.numeric(percent_identity)),
      by = "gene"
    )

  plot_df2 <- plot_df %>%
    dplyr::mutate(
      cat = factor(
        combined_category,
        levels = c("cis", "neither", "trans", "not DD")
      )
    ) %>%
    dplyr::filter(!is.na(percent_identity), is.finite(percent_identity))

  # cis vs trans only
  cis_trans <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans")) %>%
    dplyr::mutate(cat = droplevels(cat))

  pval <- wilcox.test(percent_identity ~ cat, data = cis_trans)$p.value
  p_label <- paste0("p < ", round_up_signif(pval))

  ymin <- 80
  ymax <- 103

  y_pos <- min(ymax - 1, max(plot_df2$percent_identity, na.rm = TRUE) * 1.05)
  h <- (ymax - ymin) * 0.03

  p <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = percent_identity, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "neither" = "#0072B2",
      "trans"   = "#D55E00",
      "not DD"  = "grey70"
    )) +
    ggplot2::labs(x = "", y = "Percent coding identity") +

    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", linewidth = 1, fill = NA),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 14),
      axis.text  = ggplot2::element_text(size = 12)
    ) +
    ggplot2::coord_cartesian(ylim = c(ymin, ymax)) +
    ggplot2::scale_x_discrete(labels = c(
      cis      = expression(italic("cis")),
      neither  = "neither",
      trans    = expression(italic("trans")),
      `not DD` = "not DD"
    ))

  return(p)
}
```

```{r}
p_0.6_0.4_pci  <- pci_cutoff(dd_results, mart, 0.6, 0.4)
p_0.65_0.35_pci <- pci_cutoff(dd_results, mart, 0.65, 0.35)
p_0.7_0.3_pci  <- pci_cutoff(dd_results, mart, 0.7, 0.3)
p_0.75_0.25_pci <- pci_cutoff(dd_results, mart, 0.75, 0.25)
p_0.8_0.2_pci  <- pci_cutoff(dd_results, mart, 0.8, 0.2)
p_0.85_0.15_pci <- pci_cutoff(dd_results, mart, 0.85, 0.15)

#p_0.6_0.4_pci + p_0.65_0.35_pci + p_0.7_0.3_pci + p_0.75_0.25_pci + p_0.8_0.2_pci + p_0.85_0.15_pci
```

Using mixed definition.

```{r}
pci_cutoff <- function(dd_results, mart, celltypes3, cis_cutoff, trans_cutoff) {

  # step 1: create per-cell-type cis/trans/neither/not DD categories
  dd_results2 <- dd_results %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(paste0("dd.cis.proportion.", celltypes3)),
        ~ {
          ct <- sub("^dd\\.cis\\.proportion\\.", "", dplyr::cur_column())
          dd_col <- paste0("dd.lfsr.diploid.", ct)

          if (!dd_col %in% names(dd_results)) {
            return(rep(NA_character_, length(.x)))
          }

          dd_flag <- dd_results[[dd_col]]

          out <- rep("not DD", length(.x))
          is_dd <- !is.na(dd_flag) & dd_flag < 0.05 & !is.na(.x)

          out[is_dd & .x > cis_cutoff] <- "cis"
          out[is_dd & .x < trans_cutoff] <- "trans"
          out[is_dd & .x >= trans_cutoff & .x <= cis_cutoff] <- "neither"

          out
        },
        .names = "dd.cis.category.{sub('^dd\\\\.cis\\\\.proportion\\\\.', '', .col)}"
      )
    )

  # step 2: only use the per-cell-type category columns
  cat_cols <- paste0("dd.cis.category.", celltypes3)

  # step 3: create overall category using same logic as master_df_filtered
  dd_results2 <- dd_results2 %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      dd.cis.category.overall = {
        vals <- dplyr::c_across(dplyr::all_of(cat_cols))
        vals <- vals[!is.na(vals)]

        has_cis     <- "cis" %in% vals
        has_trans   <- "trans" %in% vals
        has_neither <- "neither" %in% vals

        if (!has_cis & !has_trans & !has_neither) {
          "not DD"
        } else if (has_cis & has_trans) {
          "mixed"
        } else if (has_cis) {
          "cis"
        } else if (has_trans) {
          "trans"
        } else if (has_neither) {
          "neither"
        } else {
          "not DD"
        }
      }
    ) %>%
    dplyr::ungroup()

  # step 4: join percent coding identity
  plot_df <- dd_results2 %>%
    dplyr::select(gene, dd.cis.category.overall) %>%
    dplyr::left_join(
      mart %>%
        dplyr::select(Gene.name, percent_identity) %>%
        dplyr::rename(gene = Gene.name),
      by = "gene"
    )

  plot_df2 <- plot_df %>%
    dplyr::mutate(
      cat = factor(
        as.character(dd.cis.category.overall),
        levels = c("cis", "mixed", "trans", "neither", "not DD")
      ),
      percent_identity = as.numeric(percent_identity)
    ) %>%
    dplyr::filter(!is.na(percent_identity), is.finite(percent_identity))

  # step 5: Wilcoxon test on cis vs trans only
  cis_trans <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans")) %>%
    dplyr::mutate(cat = droplevels(cat))

  pval <- wilcox.test(percent_identity ~ cat, data = cis_trans)$p.value
  p_label <- paste0("p < ", round_up_signif(pval))

  ymin <- 80
  ymax <- 103
  y_pos <- min(ymax - 1, max(plot_df2$percent_identity, na.rm = TRUE) * 1.05)
  h <- (ymax - ymin) * 0.03

  # step 6: plot
  p <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = percent_identity, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    # bracket from cis (1) to trans (3)
    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4, parse = TRUE) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "mixed"   = "#CC79A7",
      "trans"   = "#D55E00",
      "neither" = "#0072B2",
      "not DD"  = "grey70"
    )) +
    ggplot2::labs(x = "", y = "Percent coding identity") +
    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", linewidth = 1, fill = NA),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 14),
      axis.text  = ggplot2::element_text(size = 12)
    ) +
    ggplot2::coord_cartesian(ylim = c(ymin, ymax)) +
    ggplot2::scale_x_discrete(labels = c(
      "cis"     = expression(italic("cis")),
      "mixed"   = "mixed",
      "trans"   = expression(italic("trans")),
      "neither" = "neither",
      "not DD"  = "not DD"
    ))

  return(p)
}
```

```{r}
p_0.6_0.4_pci   <- pci_cutoff(dd_results, mart, celltypes3, 0.6, 0.4)
p_0.65_0.35_pci <- pci_cutoff(dd_results, mart, celltypes3, 0.65, 0.35)
p_0.7_0.3_pci   <- pci_cutoff(dd_results, mart, celltypes3, 0.7, 0.3)
p_0.75_0.25_pci <- pci_cutoff(dd_results, mart, celltypes3, 0.75, 0.25)
p_0.8_0.2_pci   <- pci_cutoff(dd_results, mart, celltypes3, 0.8, 0.2)
p_0.85_0.15_pci <- pci_cutoff(dd_results, mart, celltypes3, 0.85, 0.15)

# p_0.6_0.4_pci + p_0.65_0.35_pci + p_0.7_0.3_pci +   p_0.75_0.25_pci + p_0.8_0.2_pci + p_0.85_0.15_pci
```

### dN/dS

Load table with dN/dS ratios.
```{r}
# load data table
dn.ds <- read.table( gzfile("/project2/gilad/kenneth/References/Overall_dN_dS.bed.gz"), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
```

```{r}
plot_df <- dd_results %>%
  dplyr::select(gene, dd.cis.category.overall) %>%
  dplyr::left_join(
    dn.ds %>%
      dplyr::select(V4, V11) %>%
      dplyr::rename(dn.ds = V11) %>%
      dplyr::mutate(dn.ds = as.numeric(dn.ds)),
    by = c("gene" = "V4")
  )

plot_df2 <- plot_df %>%
  dplyr::mutate(
    cat = factor(
      as.character(dd.cis.category.overall),
      levels = c("cis", "neither","mixed", "trans", "not DD")
    )
  ) %>%
  dplyr::filter(!is.na(dn.ds), is.finite(dn.ds))

# cis vs trans only for significance
cis_trans <- plot_df2 %>%
  dplyr::filter(cat %in% c("cis", "trans"))

pval <- wilcox.test(dn.ds ~ cat, data = cis_trans)$p.value
p_label <- paste0("p = ", format.pval(pval, digits = 2, eps = 1e-300))

# ---- set y-limits for dn/ds panel ----
ymin <- 0
ymax <- 0.6

# ---- bracket positions INSIDE the panel ----
y_pos <- min(ymax - 0.01, max(plot_df2$dn.ds, na.rm = TRUE) * 1.1)
h     <- 0.02

p <- ggplot(plot_df2, aes(x = cat, y = dn.ds, fill = cat)) +
  geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

  # significance bracket inside plot box
  annotate("segment", x = 1, xend = 4, y = y_pos, yend = y_pos) +
  annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
  annotate("segment", x = 4, xend = 4, y = y_pos, yend = y_pos - h) +
  annotate("text", x = 2.5, y = y_pos + h, label = p_label, size = 4) +

  scale_fill_manual(values = c(
    "cis"     = "#009E73",
    "trans"   = "#D55E00",
    "neither" = "#0072B2",
    "mixed"   = "#CC79A7",
    "not DD"  = "grey70"
  )) +
  labs(x = "", y = "dN/dS") +
 ggpubr::theme_pubr()+
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 1, fill = NA),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12)
  ) +
  coord_cartesian(ylim = c(ymin, ymax)) +
  scale_x_discrete(labels = c(
    cis      = expression(italic("cis")),
    trans    = expression(italic("trans")),
    neither  = "neither",
    mixed = "mixed",
    `not DD` = "not DD"
  ))

p
ggsave(paste0(results_path, 'dnds_cis_trans.pdf'), plot = p, device = "pdf", width = 7, height = 5, units = "in")
```

#### Supplementary varying cis/trans cutoff

Using mean definition for cis and trans genes.
```{r}
dn_ds_cutoff <- function(dd_results, dn.ds, cis_cutoff, trans_cutoff) {

  plot_df <- dd_results %>%
    dplyr::select(gene, dd.in.diploid.mean.cis.proportion) %>%
    dplyr::mutate(
      combined_category = dplyr::case_when(
        is.na(dd.in.diploid.mean.cis.proportion) ~ "not DD",
        dd.in.diploid.mean.cis.proportion > cis_cutoff   ~ "cis",
        dd.in.diploid.mean.cis.proportion < trans_cutoff ~ "trans",
        TRUE                                                ~ "neither"
      )
    ) %>%
    dplyr::left_join(
      dn.ds %>%
        dplyr::select(V4, V11) %>%
        dplyr::rename(dn.ds = V11),
      by = c("gene" = "V4")
    )

  plot_df2 <- plot_df %>%
    dplyr::mutate(
      cat = factor(
        combined_category,
        levels = c("cis", "neither", "trans", "not DD")
      ),
      dn.ds = as.numeric(dn.ds)
    ) %>%
    dplyr::filter(!is.na(dn.ds), is.finite(dn.ds))

  # cis vs trans only for significance
  cis_trans <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans"))

  pval <- wilcox.test(dn.ds ~ cat, data = cis_trans)$p.value
  p_label <- paste0("p = ", format.pval(pval, digits = 2, eps = 1e-300))

  ymin <- 0
  ymax <- 0.7

  y_pos <- min(ymax - 0.03, max(plot_df2$dn.ds, na.rm = TRUE) * 1.1)
  h <- 0.02

  p <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = dn.ds, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "neither" = "#0072B2",
      "trans"   = "#D55E00",
      "not DD"  = "grey70"
    )) +
    ggplot2::labs(x = "", y = "dN/dS") +

    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", linewidth = 1, fill = NA),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 14),
      axis.text  = ggplot2::element_text(size = 12)
    ) +
    ggplot2::coord_cartesian(ylim = c(ymin, ymax)) +
    ggplot2::scale_x_discrete(labels = c(
      cis      = expression(italic("cis")),
      neither  = "neither",
      trans    = expression(italic("trans")),
      `not DD` = "not DD"
    ))

  return(p)
}
```


```{r}
p_0.6_0.4dnds <-dn_ds_cutoff(dd_results,dn.ds,0.6,0.4)
p_0.65_0.35dnds <-dn_ds_cutoff(dd_results,dn.ds,0.65,0.35)
p_0.7_0.3dnds <-dn_ds_cutoff(dd_results,dn.ds,0.7,0.3)
p_0.75_0.25dnds <-dn_ds_cutoff(dd_results,dn.ds,0.75,0.25)
p_0.8_0.2dnds <-dn_ds_cutoff(dd_results,dn.ds,0.8,0.2)
p_0.85_0.15dnds <-dn_ds_cutoff(dd_results,dn.ds,0.85,0.15)

#p_0.6_0.4dnds + p_0.65_0.35dnds + p_0.7_0.3dnds + p_0.75_0.25dnds + p_0.8_0.2dnds + p_0.85_0.15dnds
```



Using the mixed definition for cis and trans genes.
```{r}
dn_ds_cutoff <- function(dd_results, dn.ds, celltypes3, cis_cutoff, trans_cutoff) {

  # step 1: create per-cell-type cis/trans/neither/not DD categories
  dd_results2 <- dd_results %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(paste0("dd.cis.proportion.", celltypes3)),
        ~ {
          ct <- sub("^dd\\.cis\\.proportion\\.", "", dplyr::cur_column())
          dd_col <- paste0("dd.lfsr.diploid.", ct)

          if (!dd_col %in% names(dd_results)) {
            return(rep(NA_character_, length(.x)))
          }

          dd_flag <- dd_results[[dd_col]]

          out <- rep("not DD", length(.x))
          is_dd <- !is.na(dd_flag) & dd_flag < 0.05 & !is.na(.x)

          out[is_dd & .x > cis_cutoff] <- "cis"
          out[is_dd & .x < trans_cutoff] <- "trans"
          out[is_dd & .x >= trans_cutoff & .x <= cis_cutoff] <- "neither"

          out
        },
        .names = "dd.cis.category.{sub('^dd\\\\.cis\\\\.proportion\\\\.', '', .col)}"
      )
    )

  # step 2: only use the per-cell-type category columns
  cat_cols <- paste0("dd.cis.category.", celltypes3)

  # step 3: create overall category using same logic as master_df_filtered
  dd_results2 <- dd_results2 %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      dd.cis.category.overall = {
        vals <- dplyr::c_across(dplyr::all_of(cat_cols))
        vals <- vals[!is.na(vals)]

        has_cis     <- "cis" %in% vals
        has_trans   <- "trans" %in% vals
        has_neither <- "neither" %in% vals

        if (!has_cis & !has_trans & !has_neither) {
          "not DD"
        } else if (has_cis & has_trans) {
          "mixed"
        } else if (has_cis) {
          "cis"
        } else if (has_trans) {
          "trans"
        } else if (has_neither) {
          "neither"
        } else {
          "not DD"
        }
      }
    ) %>%
    dplyr::ungroup()

  # step 4: join dN/dS
  plot_df <- dd_results2 %>%
    dplyr::select(gene, dd.cis.category.overall) %>%
    dplyr::left_join(
      dn.ds %>%
        dplyr::select(V4, V11) %>%
        dplyr::rename(dn.ds = V11),
      by = c("gene" = "V4")
    )

  plot_df2 <- plot_df %>%
    dplyr::mutate(
      cat = factor(
        as.character(dd.cis.category.overall),
        levels = c("cis", "mixed", "trans", "neither", "not DD")
      ),
      dn.ds = as.numeric(dn.ds)
    ) %>%
    dplyr::filter(!is.na(dn.ds), is.finite(dn.ds))

  # step 5: Wilcoxon test on cis vs trans only
  cis_trans <- plot_df2 %>%
    dplyr::filter(cat %in% c("cis", "trans"))

  pval <- wilcox.test(dn.ds ~ cat, data = cis_trans)$p.value
  p_label <- paste0("p = ", format.pval(pval, digits = 2, eps = 1e-300))

  ymin <- 0
  ymax <- 0.6
  y_pos <- min(ymax - 0.03, max(plot_df2$dn.ds, na.rm = TRUE) * 1.1)
  h <- 0.02

  # step 6: plot
  p <- ggplot2::ggplot(plot_df2, ggplot2::aes(x = cat, y = dn.ds, fill = cat)) +
    ggplot2::geom_boxplot(color = "black", alpha = 0.7, outlier.shape = NA) +

    # bracket from cis (1) to trans (3)
    ggplot2::annotate("segment", x = 1, xend = 3, y = y_pos, yend = y_pos) +
    ggplot2::annotate("segment", x = 1, xend = 1, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("segment", x = 3, xend = 3, y = y_pos, yend = y_pos - h) +
    ggplot2::annotate("text", x = 2, y = y_pos + h, label = p_label, size = 4) +

    ggplot2::scale_fill_manual(values = c(
      "cis"     = "#009E73",
      "mixed"   = "#CC79A7",
      "trans"   = "#D55E00",
      "neither" = "#0072B2",
      "not DD"  = "grey70"
    )) +
    ggplot2::labs(x = "", y = "dN/dS") +
    ggpubr::theme_pubr() +
    ggplot2::theme(
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(color = "black", linewidth = 1, fill = NA),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor.y = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 14),
      axis.text  = ggplot2::element_text(size = 12)
    ) +
    ggplot2::coord_cartesian(ylim = c(ymin, ymax)) +
    ggplot2::scale_x_discrete(labels = c(
      "cis"     = expression(italic("cis")),
      "mixed"   = "mixed",
      "trans"   = expression(italic("trans")),
      "neither" = "neither",
      "not DD"  = "not DD"
    ))

  return(p)
}
```

```{r}
p_0.6_0.4dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.6,0.4)
p_0.65_0.35dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.65,0.35)
p_0.7_0.3dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.7,0.3)
p_0.75_0.25dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.75,0.25)
p_0.8_0.2dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.8,0.2)
p_0.85_0.15dnds <-dn_ds_cutoff(dd_results,dn.ds, celltypes3,0.85,0.15)

# p_0.6_0.4dnds + p_0.65_0.35dnds + p_0.7_0.3dnds + p_0.75_0.25dnds + p_0.8_0.2dnds + p_0.85_0.15dnds
```



### Supplementary: Determining cis/trans cutoffs

#### Number of DD Genes

Using cis cutoffs.
```{r}
celltypes <- c("ENS.glia", "Glioblast", "Neuroblast", "Neuronal.IPC", "Radial.glia")
cutoffs   <- seq(0.5, 1, by = 0.1)

# ----------------------------------------------
# 1. Wide -> long for cis proportion
#    dd.cis.proportion.<celltype>
# ----------------------------------------------
cis_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.proportion.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.cis.proportion."),
    names_to  = "celltype",
    values_to = "cis_prop"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.cis\\.proportion\\.", "", celltype)
  )

# ----------------------------------------------
# 2. Wide -> long for diploid DD significance
#    dd.lfsr.diploid.<celltype>
# ----------------------------------------------
lfsr_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.lfsr.diploid.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.lfsr.diploid."),
    names_to  = "celltype",
    values_to = "dd_lfsr_diploid"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.lfsr\\.diploid\\.", "", celltype)
  )

# ----------------------------------------------
# 3. Merge and filter to diploid DD genes only
#    condition: dd.lfsr.diploid < 0.05
# ----------------------------------------------
df_dd <- cis_long %>%
  dplyr::inner_join(lfsr_long, by = c("gene", "celltype")) %>%
  dplyr::filter(
    celltype %in% celltypes,
    !is.na(dd_lfsr_diploid),
    dd_lfsr_diploid < 0.05
  )

# ----------------------------------------------
# 4. Count cis genes passing each cutoff by cell type
#    condition: cis_prop >= cutoff
# ----------------------------------------------
summary_plot_dd <- tidyr::expand_grid(
  celltype = celltypes,
  cutoff   = cutoffs
) %>%
  dplyr::left_join(df_dd, by = "celltype") %>%
  dplyr::group_by(celltype, cutoff) %>%
  dplyr::summarise(
    n_cis_genes = sum(cis_prop >= cutoff, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------
# 5. Plot
# ----------------------------------------------
p_cis_cutoff_genes <- ggplot2::ggplot(
  summary_plot_dd,
  ggplot2::aes(x = cutoff, y = n_cis_genes, color = celltype)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggplot2::coord_cartesian(ylim = c(0, 800)) +
  ggplot2::scale_y_continuous(breaks = c(0, 200,400,600,800)) +
  ggplot2::scale_x_continuous(breaks = cutoffs, limits = c(0.5, 1)) +
  ggplot2::labs(
    x = expression(italic("cis") ~ "proportion"),
    y = expression("#" ~ italic("cis") ~ "DD genes"),
    color = "Cell Type"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(
      color = "black",
      fill  = NA,
      linewidth = 1
    ),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12)
  )

p_cis_cutoff_genes
#ggsave(paste0(results_path, 'supplement/cis_cutoff_num_dd_genes.pdf'), plot = p_cis_cutoff_genes, device = "pdf", width = 7, height = 5, units = "in")
```

Using trans cutoffs.
```{r}
cutoffs_trans <- seq(0, 0.5, by = 0.1)

# ----------------------------------------------
# 1. Wide -> long for cis proportion
#    dd.cis.proportion.<celltype>
# ----------------------------------------------
cis_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.cis.proportion.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.cis.proportion."),
    names_to  = "celltype",
    values_to = "cis_prop"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.cis\\.proportion\\.", "", celltype)
  )

# ----------------------------------------------
# 2. Wide -> long for diploid DD significance
#    dd.lfsr.diploid.<celltype>
# ----------------------------------------------
lfsr_long <- dd_results %>%
  dplyr::select(gene, starts_with("dd.lfsr.diploid.")) %>%
  tidyr::pivot_longer(
    cols      = starts_with("dd.lfsr.diploid."),
    names_to  = "celltype",
    values_to = "dd_lfsr_diploid"
  ) %>%
  dplyr::mutate(
    celltype = sub("^dd\\.lfsr\\.diploid\\.", "", celltype)
  )

# ----------------------------------------------
# 3. Merge and filter to diploid DD genes only
#    condition: dd.lfsr.diploid < 0.05
# ----------------------------------------------
df_dd <- cis_long %>%
  dplyr::inner_join(lfsr_long, by = c("gene", "celltype")) %>%
  dplyr::filter(
    celltype %in% celltypes,
    !is.na(dd_lfsr_diploid),
    dd_lfsr_diploid < 0.05
  )

# ----------------------------------------------
# 4. Count trans genes at each cutoff by cell type
#    Define trans as: cis_prop <= cutoff
# ----------------------------------------------
summary_plot_trans <- tidyr::expand_grid(
  celltype = celltypes,
  cutoff   = cutoffs_trans
) %>%
  dplyr::left_join(df_dd, by = "celltype") %>%
  dplyr::group_by(celltype, cutoff) %>%
  dplyr::summarise(
    n_trans_genes = sum(cis_prop <= cutoff, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------
# 5. Plot
# ----------------------------------------------
p_trans_cutoff_genes <- ggplot2::ggplot(
  summary_plot_trans,
  ggplot2::aes(x = cutoff, y = n_trans_genes, color = celltype)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggplot2::coord_cartesian(ylim = c(0, 800)) +
  ggplot2::scale_y_continuous(breaks = c(0, 200,400,600,800)) +
  ggplot2::scale_x_continuous(breaks = cutoffs_trans, limits = c(0, 0.5)) +
  ggplot2::scale_color_discrete(
    labels = function(x) gsub("\\.", " ", x)
  ) +
  ggplot2::labs(
    x = expression(italic("cis") ~ "proportion"),
    y = expression("#" ~ italic("trans") ~ "DD genes"),
    title = "",
    color = "Cell Type"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    panel.border = ggplot2::element_rect(
      color = "black",
      fill  = NA,
      linewidth = 1
    ),
    axis.title = ggplot2::element_text(size = 14),
    axis.text  = ggplot2::element_text(size = 12)
  )

p_trans_cutoff_genes
#ggsave(paste0(results_path, 'supplement/trans_cutoff_num_dd_genes.pdf'), plot = p_trans_cutoff_genes, device = "pdf", width = 7, height = 5, units = "in")
```

#### DD genes that exist in multiple cell types


Using cis cutoffs.
```{r}
N_ct <- length(celltypes)

# For each cutoff, count how many cell types each gene is cis-like in
# (defined as dd.cis.proportion >= cutoff among diploid DD genes)
gene_ct_counts_cis_dd <- df_dd %>%
  dplyr::filter(celltype %in% celltypes) %>%
  tidyr::expand_grid(cutoff = cutoffs) %>%
  dplyr::filter(cis_prop >= cutoff) %>%
  dplyr::distinct(gene, celltype, cutoff) %>%
  dplyr::count(cutoff, gene, name = "n_celltypes_cis")

# Histogram: number of genes by exact number of cell types (exclude 0)
plot_df_cis_dd <- gene_ct_counts_cis_dd %>%
  dplyr::filter(n_celltypes_cis >= 1) %>%
  dplyr::count(cutoff, n_celltypes_cis, name = "n_genes") %>%
  tidyr::complete(
    cutoff,
    n_celltypes_cis = 1:N_ct,
    fill = list(n_genes = 0)
  ) %>%
  dplyr::mutate(
    n_celltypes_cis = factor(n_celltypes_cis, levels = 1:N_ct),
    cutoff = factor(cutoff, levels = sort(unique(cutoffs)))
  )

# Plot
p_cis_dd_celltypes <- ggplot2::ggplot(
  plot_df_cis_dd,
  ggplot2::aes(x = n_celltypes_cis, y = n_genes)
) +
  ggplot2::geom_col(fill = "#619CFF") +
  ggplot2::geom_text(
    ggplot2::aes(label = n_genes),
    vjust = -0.3,
    size = 3
  ) +
  ggplot2::facet_wrap(~ cutoff, ncol = 3) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.1))
  ) +
  ggplot2::labs(
    x = expression("# cell types where" ~ italic("cis") ~ "DD genes exist"),
    y = "# DD genes",
    title = ""
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    ),
    axis.line = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(
      fill = "grey88",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = ggplot2::element_text(face = "bold"),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.3)
  )

p_cis_dd_celltypes
#ggsave(paste0(results_path, 'supplement/cis_cutoff_num_dd_genes_per_celltype.pdf'), plot = p_cis_dd_celltypes, device = "pdf", width = 7, height = 5, units = "in")
```

Using trans cutoffs.
```{r}
# For each cutoff, count how many cell types each gene is trans-like in
# (defined as dd.cis.proportion <= cutoff among diploid DD genes)
gene_ct_counts_trans_dd <- df_dd %>%
  dplyr::filter(celltype %in% celltypes) %>%
  tidyr::expand_grid(cutoff = cutoffs_trans) %>%
  dplyr::filter(cis_prop <= cutoff) %>%
  dplyr::distinct(gene, celltype, cutoff) %>%
  dplyr::count(cutoff, gene, name = "n_celltypes_trans")

# Histogram: number of genes by exact number of cell types (exclude 0)
plot_df_trans_dd <- gene_ct_counts_trans_dd %>%
  dplyr::filter(n_celltypes_trans >= 1) %>%
  dplyr::count(cutoff, n_celltypes_trans, name = "n_genes") %>%
  tidyr::complete(
    cutoff,
    n_celltypes_trans = 1:N_ct,
    fill = list(n_genes = 0)
  ) %>%
  dplyr::mutate(
    n_celltypes_trans = factor(n_celltypes_trans, levels = 1:N_ct),
    cutoff = factor(cutoff, levels = sort(unique(cutoffs_trans)))
  )

# Plot
p_trans_cutoff_celltype <- ggplot2::ggplot(
  plot_df_trans_dd,
  ggplot2::aes(x = n_celltypes_trans, y = n_genes)
) +
  ggplot2::geom_col(fill = "#619CFF") +
  ggplot2::geom_text(
    ggplot2::aes(label = n_genes),
    vjust = -0.3,
    size = 3
  ) +
  ggplot2::facet_wrap(~ cutoff, ncol = 3) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.1))
  ) +
  ggplot2::labs(
    x = expression("# cell types where" ~ italic("trans") ~ "DD genes exist"),
    y = "# DD genes",
    title = ""
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    ),
    axis.line = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(
      fill = "grey88",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = ggplot2::element_text(face = "bold"),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.3)
  )

p_trans_cutoff_celltype
#ggsave(paste0(results_path, 'supplement/trans_cutoff_num_dd_genes_per_celltype.pdf'), plot = p_trans_cutoff_celltype, device = "pdf", width = 7, height = 5, units = "in")
```

## Supplementary: Cell Counts
```{r}
path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'
cell_annotations <- read.csv(paste0(path,'barr_crossfilt_cell_annotations.csv'))

# cell types you want to keep
celltypes_keep <- c(
  "Neuroblast",
  "ENS glia",
  "Radial glia",
  "Neuronal IPC",
  "Glioblast"
)

plot_df <- cell_annotations %>%
  mutate(
    species = case_when(
      str_starts(cell_barcode, "hhy") ~ "Allotetraploid",
      str_starts(cell_barcode, "chy") ~ "Allotetraploid",
      str_starts(cell_barcode, "h")   ~ "Human",
      str_starts(cell_barcode, "c")   ~ "Chimpanzee",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(
    celltype %in% celltypes_keep,
    !is.na(species)
  ) %>%
  dplyr::count(celltype, species, name = "cell_count") %>%
  mutate(
    celltype = factor(celltype, levels = celltypes_keep),
    species = factor(species, levels = c("Human", "Chimpanzee", "Allotetraploid"))
  )

y_max <- max(plot_df$cell_count, na.rm = TRUE)

dodge <- position_dodge(width = 0.9)

p_cell_counts <- ggplot(plot_df, aes(x = celltype, y = cell_count, fill = species)) +
  geom_col(
    position = dodge,
    width = 0.75,
    color = "black",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = scales::comma(cell_count)),
    position = dodge,
    vjust = -0.25,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Human" = "#D55E00",
      "Chimpanzee" = "#0072B2",
      "Allotetraploid" = "#009E73"
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    x = "",
    y = "Number of cells",
    fill = ""
  ) +
  coord_cartesian(ylim = c(0, 40000), clip = "off") +
  theme_classic(base_size = 12) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.line = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "top",
    plot.margin = margin(10,10,10,10)
  )

p_cell_counts
#ggsave(paste0(results_path, 'supplement/cellcounts.pdf'), plot = p_cell_counts, device = "pdf", width = 7, height = 5, units = "in")
```




