---
title: "Simulation scRNA-seq"
output: html_document
date: "2026-03-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The simulation framework is designed to simulate scRNA-seq data using a Gamma-Poisson distribution. It has the flexibility to alter dispersion levels.

# **Simulation Details**

We simulate a matrix $Y \in \mathbb{R}^{n\times p}$ that contains raw expression values (scRNA-seq counts) for $p$ genes and $n$ single cells. We set $\pi$ as the percent of genes that are overdispersed. We use a Gamma-Poisson model for each gene and can set the dispersion level for each simulated gene.

```{r, eval=F,include=F}
## **Poisson Distribution**

### ***Model***

We first simulate a Poisson distribution as our null model for each gene, where there is no dispersion parameter.

$$
y_{ij} \sim Pois(L_i\lambda_j) 
$$

Where the variables for cell $i$ and gene $j$ are:

-   $y_{ij}$ is the number of unique molecular identifier (UMI) for cell $i$ and gene $j$

-   $L_i$ is the library size of cell $i$

-   $\lambda_j$ is the relative expression value of gene $j$

### ***Density Function***

The density function for the Poisson distribution is:

$$
f(y;\lambda) = \frac{\lambda^y e^{-\lambda}}{y!}
$$

Where for $y =0,1,2,...$ the $E[Y] = Var[Y] = \lambda$.

### ***Simulation Parameters***

We use the **rpois(**$n$, $\lambda$) function in R to simulate the Poisson distribution. The library size and relative expression values are calculated from a real data set (human-chimp hybrid data). Each $Y$ is calculated for cells from the same species, cell type, and replicate. We simulate for $n=500$ cells.
```

## **Gamma-Poisson (NB) Distribution**

### ***Model***

We simulate the Negative Binomial distribution using the Gamma-Poisson distribution, since the Gamma distribution is a conjugate prior of the Poisson distribution. The mean is fixed for each bin of mean expression. The dispersion parameter differs from the background versus the dispersed gene. Under this model,

$$
y_{ij} \sim Pois(L_i \lambda_{ij})
$$

$$
\lambda_{ij} \sim \begin{equation}
  f_j(\cdot) =
    \begin{cases}
      Gamma(\mu_{j},\phi_{j_0}) & \text{for background gene} \\
      Gamma(\mu_j,\phi_j) & \text{for overdispersed gene} \\
    \end{cases}       
\end{equation}
$$

Where the variables for cell $i$ and gene $j$ are:

-   $y_{ij}$ is the number of unique molecular identifier (UMI) for cell $i$ and gene $j$

-   $L_i$ is the library size of cell $i$

-   $\lambda_{ij}$ is the latent expression value of gene $j$

-   $\mu_j$ is the mean relative expression value of $\lambda_{ij}$ for gene $j$

-   $\phi_j$ is the dispersion value of $\lambda_{ij}$ for gene $j$

-   $\phi_{j_0}$ is the null dispersion value for a background gene

### ***Density Function***

The density function for the Gamma distribution is:

$$
f(y; \mu,\phi) = \frac{(\phi^{-1}\mu^{-1})^{\phi^{-1}}}{\Gamma(\phi^{-1})} y^{\phi^{-1}-1} e^{\phi^{-1}\mu^{-1}y}
$$

Where for $y =0,1,2,...$ the $E[Y] = \mu$ and $Var[Y] = \mu^2 \phi$.

### ***Simulation Parameters***

We will first simulate $\lambda_j$ using the **rgamma(**$n$,$\alpha$,$\sigma$) function in R with mean $\alpha\sigma$ and variance $\alpha\sigma^2$. Thus, $\mu=\alpha\sigma$ and $\phi = 1/\alpha$. We calculate $\mu_j$ for each gene $j$ and $L_i$ for each cell $i$ from the hybrid data set. For genes with dispersion, the $E[f] = \mu_0$. We next simulate the Poisson distribution using the **rpois(**$n$, $\lambda$) function in R.

## Simulation
```{r}
# function to simulate a null gene and dispersed gene
sim_gene <- function(n,alpha,sigma,Li, sigma_multiplier=1){

  # Simulate a single gene with following parameters:
  # n:                number of cells
  # alpha:            alpha parameter
  # sigma:            sigma parameter
  # Li:               library size
  # sigma_multiplier: factor to increase variance
 
  # Returns expression values for single gene in n cells
  
  # initialize Y
  Y <- rep(NA,n) 
  
  # calculate latent expression using gamma distribution
  # ensure that simulation has mean within 10% of target
  # perc_diff = 1
  # while(perc_diff > 0.05){
  lambdaj <- rgamma(n,shape = alpha/sigma_multiplier,scale = sigma*sigma_multiplier)
  # perc_diff = (alpha*sigma - mean(lambdaj))/alpha*sigma
  # }

  
  #lambdaj <- rgamma(n,shape = alpha/sigma_multiplier,scale = sigma*sigma_multiplier)
  
  # for each cell use the latent expression and library size to simulate expression
  for(i in 1:n){
  
    Y[i] <- rpois(1,Li*lambdaj[i])
  }
  
  return(Y)
}


# calculate alpha given mean expression and fixed sigma
calc_alpha <- function(mu, sigma){
  alpha = mu/sigma
  return(alpha)
}



# Function to simulate genes within a bin of mean expression
sim_bin <- function(n,p,phi,sigma_multiplier_ovr,sigma_multiplier_undr,Li,mean_expr_lwr,mean_expr_upr){
  
  # Simulate all genes with following parameters:
  # n:                      number of cells
  # p:                      number of genes
  # phi:                    estimated dispersion parameter
  # sigma_multiplier_ovr:   factor applied to sigma parameter for 5% of genes
  # sigma_multiplier_undr:  factor applied to sigma parameter for 5% of genes
  # Li:                     library size
  # mean_expr_lwr:          lower bound of mean expression window
  # mean_expr_upr:          upper bound of mean expression window
 
  # Returns matrix of expression for n cells and p genes. 
  # Returns vector of index values for simulated dispersed genes.
  
  # initialize Y for bin of expression
  Y_bin <- matrix(NA,nrow=n,ncol=p)
  
  alpha_bin <- c()
  sigma_bin <- c()
  
  n_bins_exp_window = 100
  # breakdown expression range into p values
  bin_expression <- seq(mean_expr_lwr,mean_expr_upr,length.out=n_bins_exp_window)
  
  # randomly select p expression means
  bin_sim_idx <- sample.int(n_bins_exp_window, p, replace = TRUE)
  # randomly select 10% to be significantly dispersed
  bin_disp_idx <- sample.int(n_bins_exp_window, 0.1*p, replace = FALSE)
  # make half of indices overdispersed and half underdispersed
  half <- length(bin_disp_idx) / 2
  bin_disp_idx_ovr <- bin_disp_idx[1:half]
  bin_disp_idx_undr <- bin_disp_idx[(half + 1):length(bin_disp_idx)]
  # get expression means
  sim_expression_values <- bin_expression[bin_sim_idx]
  
  # simulate genes for each expression mean
  for(i in 1:p){
    
    # get expression mean
    mu <- sim_expression_values[i]
    
    # calculate sigma parameter
    sigma <- mu*phi/Li
    alpha <- 1/phi
    #sigma <- mu*phi + 1
    sigma_bin <- c(sigma_bin,sigma)
    #sigma_disp <- mu*phi*phi_multiplier + 1
    
    # calculate alpha for fixed sigma based on expression mean
    #alpha <- calc_alpha(mu,sigma)
    alpha_bin <- c(alpha_bin,alpha)
    #alpha_disp <- calc_alpha(mu,sigma_disp)
    
    # for dispersed genes multiply sigma by factor
    if(i %in% bin_disp_idx_ovr){
      
       Y_bin[,i] <- sim_gene(n,alpha,sigma,Li,sigma_multiplier_ovr)
    }
    else if(i %in% bin_disp_idx_undr){
      Y_bin[,i] <- sim_gene(n,alpha,sigma,Li,sigma_multiplier_undr)
    }
    else{
      Y_bin[,i] <- sim_gene(n,alpha,sigma,Li,1)
    }
  }
  
  return(list("Y_bin"=Y_bin, "disp_idx_ovr"=bin_disp_idx_ovr,"disp_idx_undr"=bin_disp_idx_undr, "alpha_bin"=alpha_bin, "sigma_bin"=sigma_bin))
}

sim_Y <- function(n,p,phi,sigma_multiplier_ovr,sigma_multiplier_undr,Li,min_exp, max_exp,n_bins){
  Y <- c()
  disp_idx_ovr <- c()
  disp_idx_undr <- c()
  alphas <- c()
  sigmas <- c()
  #n_bins = p/100
  n_genes_per_bin = p/n_bins
  mean_intervals <- seq(min_exp, max_exp,length.out=(n_bins+1))
  
  for(i in 1:n_bins){
    
    mean_expr_lwr <- mean_intervals[i]
    mean_expr_upr <- mean_intervals[i+1]
    
    gene_sim <- sim_bin(n,n_genes_per_bin,phi,sigma_multiplier_ovr,sigma_multiplier_undr,Li,mean_expr_lwr,mean_expr_upr)
    Y_bin <- gene_sim$Y_bin
    bin_disp_idx_ovr <- gene_sim$disp_idx_ovr + (i-1)*n_bins
    bin_disp_idx_undr <- gene_sim$disp_idx_undr + (i-1)*n_bins
    alpha_bin <- gene_sim$alpha_bin
    sigma_bin <- gene_sim$sigma_bin
    
    Y <- cbind(Y,Y_bin)
    disp_idx_ovr <- c(disp_idx_ovr,bin_disp_idx_ovr)
    disp_idx_undr <- c(disp_idx_undr,bin_disp_idx_undr)
    alphas <- c(alphas,alpha_bin)
    sigmas <- c(sigmas, sigma_bin)
  }
  return(list("Y"=Y,"disp_idx_ovr"=disp_idx_ovr,"disp_idx_undr"=disp_idx_undr,"alpha"=alphas, "sigma"=sigmas))
}



```

## Real Data Comparison

We first calculate the mean gene expression for in the hybrid dataset using Human Cardiomyocytes from Replicate 3. Replicate 3 was chosen since there are 546 cells.

```{r, message=F, warning=F, eval=F}
require(Seurat)
require(SeuratDisk)
require(SeuratData)
library(ggplot2)
library(cowplot)
library(scTenifoldNet)
HC <- readRDS("/project/gilad/awchen55/differentialDispersion/data/hybrid_lines_raw_data/human.ASE.Rep3.rds")
cardiomyocytes_data <- subset(HC, subset = (labels == "Cardiomyocytes"))
cardiomyocytes_expression_data <- t(cardiomyocytes_data[['RNA']]$counts)
rm(HC)
```

Filter out low expression genes.
```{r, eval=F}
# calculate mean expression and remove genes with low expression
cardiomyocytes_expression_data_filtered <- cardiomyocytes_expression_data[, colSums(cardiomyocytes_expression_data != 0) > 10]

# calculate library size
library_sizes <- rowSums(cardiomyocytes_expression_data_filtered)

# calculate mean expression
mean_expression <- colMeans(cardiomyocytes_expression_data_filtered)
```


We simulate from 546 cells of cardiomyocytes. We simulate each gene from the real dataset and compare to the simulation.

```{r}
set.seed(20250503)
n=546
phi = 0.654
avg_li = mean(library_sizes)
# calculate sigma parameter
sigmas <- mean_expression*phi/avg_li
alpha <- 1/phi
p=length(sigmas)
Y <- matrix(NA,nrow=n,ncol=p)
for(i in 1:p){
  Y[,i] <- sim_gene(n,alpha,sigmas[i],avg_li)
}


```

Convert to Seurat Object to run memento
```{r}
counts <- t(Y)  # now p (genes) x n (cells)

# add names (recommended)
rownames(counts) <- paste0("gene_", seq_len(nrow(counts)))
colnames(counts) <- paste0("cell_", seq_len(ncol(counts)))

# (optional) metadata for this simulated dataset
meta <- data.frame(
  celltype   = rep("CellType", ncol(counts)),
  individual = rep("Ind1",        ncol(counts)),
  replicate  = rep("Rep1",        ncol(counts)),
  row.names  = colnames(counts),
  stringsAsFactors = FALSE
)

seu <- CreateSeuratObject(
  counts   = counts, 
   project = "SimData", 
   min.cells = 3,  # Minimum cells expressing a gene
   min.features = 10)  # Minimum features per cell

seu@meta.data <- meta
# sanity checks
seu
dim(seu[["RNA"]]$counts)   # genes x cells
head(seu@meta.data)
file_name <- '/project/gilad/awchen55/differentialDispersion/data/simulation_mean_var_independence/sim_cardiomyocytes'
saveRDS(seu, file = paste0(file_name,".rds"))
# SaveH5Seurat(seu, filename = paste0(file_name, ".h5Seurat"), overwrite = TRUE)
# Convert(filename = paste0(file_name, ".h5Seurat", dest = "h5ad", overwrite = TRUE))
```

Convert into components to convert Seurat to anndata.
```{r}
library(Matrix)
outdir <- '/project/gilad/awchen55/differentialDispersion/data/simulation_mean_var_independence/'
# counts (genes x cells)
counts <- GetAssayData(seu, assay = "RNA", slot = "counts")

Matrix::writeMM(counts, file = file.path(outdir, "counts.mtx"))
writeLines(rownames(counts), con = file.path(outdir, "genes.txt"))
writeLines(colnames(counts), con = file.path(outdir, "barcodes.txt"))

meta <- seu@meta.data
meta$barcode <- rownames(meta)  # keep an explicit key
write.csv(meta, file = file.path(outdir, "cell_metadata.csv"), row.names = FALSE)
```




## QQ Plot to compare real data to simulated data

```{r}
file_name <- '/project/gilad/awchen55/differentialDispersion/data/simulation_mean_var_independence/sim_cardiomyocytes'
y_sim <- readRDS(paste0(file_name,".rds"))
results_path <- '/project/gilad/awchen55/differentialDispersion/variability_paper/supplement/'
```

```{r}
library(ggplot2)
library(dplyr)

# Flatten matrices to vectors
real_vals <- as.numeric(cardiomyocytes_expression_data_filtered)
sim_vals  <- as.numeric(t(y_sim[["RNA"]]$counts))

# Use all quantiles
n <- min(length(real_vals), length(sim_vals))

qq_df <- data.frame(
  real = quantile(real_vals, probs = seq(0, 1, length.out = n), na.rm = TRUE),
  sim  = quantile(sim_vals,  probs = seq(0, 1, length.out = n), na.rm = TRUE)
)

p <- ggplot(qq_df, aes(x = real, y = sim)) +
  geom_point(alpha =0.6, size = 1.5, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(
    title = "",
    x = "Real Data Quantiles",
    y = "Simulated Data Quantiles"
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

print(p)

# Save as PNG instead of PDF for dense plots
ggsave(
  filename = paste0(results_path, "qq_plot_real_vs_simulated.png"),
  plot = p,
  width = 6,
  height = 6,
  dpi = 300
)
```




## Compare mean and dispersion in simulation]

```{r}
outdir <- '/project/gilad/awchen55/differentialDispersion/data/simulation_mean_var_independence/'
var_est <- read.csv(paste0(outdir,"memento_var_estimate_simulation.csv"))
mean_est <- read.csv(paste0(outdir,"memento_mean_estimate_simulation.csv"))
```

```{r}
plot_df <- var_est %>%
  dplyr::select(gene, sg.Ind1.Rep1.CellType) %>%
  dplyr::rename(dispersion = sg.Ind1.Rep1.CellType) %>%
  dplyr::left_join(
    mean_est %>%
      dplyr::select(gene, sg.Ind1.Rep1.CellType) %>%
      dplyr::rename(mean = sg.Ind1.Rep1.CellType),
    by = "gene"
  ) %>%
  dplyr::filter(!is.na(mean), !is.na(dispersion))

rho <- cor(plot_df$mean, plot_df$dispersion, method = "spearman")

label_text <- paste0("Spearman \u03C1 = ", round(rho, 3))

x_pos <- max(plot_df$mean, na.rm = TRUE) * 0.98
y_pos <- max(plot_df$dispersion, na.rm = TRUE) * 0.98

p <- ggplot(plot_df, aes(x = mean, y = dispersion)) +
  geom_point(alpha = 0.6, size = 1.5, color = "blue") +
  annotate(
    "text",
    x = x_pos,
    y = y_pos,
    label = label_text,
    hjust = 1,
    vjust = 1,
    size = 5
  ) +
  labs(
    x = "Mean",
    y = "Dispersion"
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

p

ggsave(
  paste0(results_path, "mean_dispersion_independence_simulated.pdf"),
  plot = p,
  width = 6,
  height = 6
)
```
