---
title: "Perform DD and Create Final Data Table"
output: html_document
date: "2026-03-04"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and functions

```{r, warning=FALSE}
library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(SingleCellExperiment)
library(variancePartition)
library(dreamlet)
library(Matrix)
library(mashr)
library(edgeR)
library(matrixTests)
```


```{r}
# a function that goes through and generates pseudobulk for cell matching metadata columns
generate.psuedobulk <- function(object, labels, assay="RNA",slot="counts") {
  factorlist <- list()
  for (i in labels) factorlist[[i]] <- unique(object@meta.data[,i])
  meta <- expand.grid(factorlist, stringsAsFactors = FALSE)
  rownames(meta) <- apply(meta, 1, function(x) paste0(x, collapse = '.'))
  
  # build the output matrix
  n <- nrow(meta)
  out <- matrix(nrow=dim(object[[assay]])[1], ncol=n, data=0)
  
  rownames(out) <- rownames(object[[assay]])
  colnames(out) <- rownames(meta)
  
  ncells <- c()
  total.cells <- dim(object[[assay]])[2]
  for (i in 1:n)
  {
    #prog(i,n)
    cells <- 1:total.cells
    for (j in names(meta)) {
      keep  <- which(object@meta.data[[j]] == meta[i,j])
      cells <- cells[cells %in% keep]
    }
    ncells[i] <- length(cells)
    if (length(cells)==1) {
      out[,i] <- slot(object[[assay]], slot)[,cells]
    } else {
      out[,i] <- Matrix::rowSums(slot(object[[assay]], slot)[,cells])
    }
  }
  meta$ncells <- ncells
  return(list(counts=out, meta=meta))
}

# keep only levels with more than 'threshold' cells
filter.pseudobulk <- function(pseudobulk, threshold = 0) {
  w <- which(pseudobulk$meta$ncells > threshold)
  pseudobulk$counts <- pseudobulk$counts[,w]
  pseudobulk$meta   <- pseudobulk$meta[w,]
  pseudobulk
}

add_sex <- function(bulk) {
  bulk$meta$sex <- rep("F", nrow(bulk$meta))
  bulk$meta$sex[bulk$meta$individual=="NA19160"] <- "M"
  bulk$meta$sex[bulk$meta$individual=="C3649"]   <- "M"
  bulk$meta$sex[bulk$meta$individual=="H28126"]  <- "M"
  bulk
}
```


## Differential Expression

### DE for Allotetraploid

We run a Wilcoxon signed-rank test between species in the allotetraploid line.
```{r, eval=FALSE}
path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'
path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/'

ct <- c('Neuroblast', 'Glioblast', 'Radial glia', 'ENS glia', 'Neuronal IPC')
ct_names <- c('Neuroblast', 'Glioblast', 'Radial_glia', 'ENS_glia', 'Neuronal_IPC')

for(i in 1:length(ct)){
  
  hybrid_data <- readRDS(paste0(path,'hybrid_crossfilt_',ct_names[i],'.rds'))
  # subset cells for one species
  human_cells <-  rownames(hybrid_data@meta.data[hybrid_data@meta.data$species == "human", ])
  chimp_cells <-  rownames(hybrid_data@meta.data[hybrid_data@meta.data$species == "chimp", ])
  
  # raw counts for just those cells
  raw_counts_human <- GetAssayData(hybrid_data, assay = "RNA", slot = "counts")[, human_cells]
  raw_counts_chimp <- GetAssayData(hybrid_data, assay = "RNA", slot = "counts")[, chimp_cells]
  
  
  
  require(matrixTests)
  DE <- row_wilcoxon_paired(
    as.matrix(raw_counts_chimp),
    as.matrix(raw_counts_human),
    alternative = "two.sided",
    null = 0,
    exact = NA,
    correct = TRUE
  )
  
  hmean <- rowSums(raw_counts_human)
  cmean <- rowSums(raw_counts_chimp)
    
  DE$logFC <- log2( (hmean + 1) / (cmean + 1) )
  DE$umi <- hmean + cmean 
                            
  
  write.csv(DE, paste0(path_out,'hybrid_wilcoxon_results_human_minus_chimp_', ct_names[i], '.csv'))
}
```

### DE for Diploid
We run DE analysis using LIMMA. We first load the annotated data.
```{r, eval=FALSE}
path_data <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'
parent_combined <- readRDS(paste0(path_data, "parent_crossfilt_combined_metadata.rds"))
hybrid_combined <- readRDS(paste0(path_data, "hybrid_crossfilt_combined_metadata.rds"))
```

Next, we pseudobulk by species, cell type, individual, and replicate.
```{r, eval=FALSE}
path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'
path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/limma_results/'
labels <- c('individual','species','replicate','celltype')
parent_data_pb <- generate.psuedobulk(parent_combined,labels)
parent_data_pb_filt <- filter.pseudobulk(parent_data_pb)
parent_data_pb_filt <- add_sex(parent_data_pb_filt)
saveRDS(parent_data_pb_filt, file = paste0(path_out,"parent_data_pb_filt.rds"))
```

Now run LIMMA using dreamlet package. We model species as fixed effect and individual as a random effect.
```{r, eval=FALSE}
path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/limma_results/'
bulk <- readRDS(paste0(path_out,'parent_data_pb_filt.rds'))
celltypes <- c('ENS glia', 'Glioblast',  'Neuroblast', 'Neuronal IPC', 'Radial glia')

for(i in 1:length(celltypes)){
  # Get logical index of samples that match your condition
  keep <- bulk$meta$celltype == celltypes[i]
  
  # Subset both counts and metadata
  bulk_subset <- list(
    counts = bulk$counts[, keep],
    meta   = bulk$meta[keep, ]
  )
  
  # make sure enough counts to run dream
  keep <- rowSums(bulk_subset$counts) > 0
  bulk_subset$counts <- bulk_subset$counts[keep, , drop = FALSE]
  
  ### Run DREAM from Kenneth pipeline
  sex=FALSE
  voomspan=0.5
  ddf = "Satterthwaite"  
  
  d0 <- DGEList(bulk_subset$counts)
  
  keep <- filterByExpr(d0$counts, group=bulk_subset$meta$species, d0$samples$lib.size, min.count = 5)
  
  drop <- which(keep==0)
  if(length(drop)>0) d0 <- d0[-drop,] 
  
  d0 <- calcNormFactors(d0, method="TMMwsp")
  
  form <- ~ 0 + species + (1|individual) 
  if (sex) form <- ~ 0 + species + (1|individual) + (1|sex)
  
  # estimate weights using linear mixed model of dream
  vobjDream = voomWithDreamWeights(d0, form, bulk_subset$meta, normalize.method = "cyclicloess", plot = FALSE, save.plot = TRUE, span=voomspan)
  
  L = getContrast( vobjDream, form, bulk_subset$meta, c("specieshuman","specieschimp"))
  fit = dream( vobjDream, form, bulk_subset$meta, L, ddf = ddf)
  
  saveRDS(fit,paste0(path_out, "parent_dream_fit_human_minus_chimp_indv_random_",celltypes[i],".rds"))
}
```

Create matrices for beta, SE, and p-values.
```{r, eval=False}
path_de_parent <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/limma_results/'
celltypes <- c('ENS glia', 'Glioblast', 'Neuroblast', 'Neuronal IPC', 'Radial glia')

merged_df        <- NULL  # logFC
merged_df_se     <- NULL  # SE
merged_df_pval   <- NULL  # p-value
merged_df_adj_pval <- NULL  # adj p-value

for (ct in celltypes) {

  memento_limma_fit <- readRDS(paste0(path_de_parent, "parent_dream_fit_human_minus_chimp_indv_random_", ct, ".rds"))
  memento_limma_res_parent <- variancePartition::topTable(memento_limma_fit, coef = "L1", number = Inf)
  memento_limma_res_parent$SE <- memento_limma_res_parent$logFC / memento_limma_res_parent$t

  if (is.null(rownames(memento_limma_res_parent))) stop(paste("No rownames in", ct))
  genes <- rownames(memento_limma_res_parent)

  # Build per-ct data.frames with gene as an explicit column
  logFC_df <- data.frame(gene = genes, logFC = memento_limma_res_parent$logFC, row.names = NULL)
  colnames(logFC_df)[2] <- paste0("de.logFC.diploid.", ct)

  SE_df <- data.frame(gene = genes, SE = memento_limma_res_parent$SE, row.names = NULL)
  colnames(SE_df)[2] <- paste0("de.SE.diploid.", ct)

  pval_df <- data.frame(gene = genes, pval = memento_limma_res_parent$P.Value, row.names = NULL)
  colnames(pval_df)[2] <- paste0("de.pvalue.diploid.", ct)

  adj_pval_df <- data.frame(gene = genes, adj.pval = memento_limma_res_parent$adj.P.Val, row.names = NULL)
  colnames(adj_pval_df)[2] <- paste0("de.adj.pvalue.diploid.", ct)

  # Merge using UNION of genes (all = TRUE)
  if (is.null(merged_df)) {
    merged_df <- logFC_df
    merged_df_se <- SE_df
    merged_df_pval <- pval_df
    merged_df_adj_pval <- adj_pval_df
  } else {
    merged_df <- merge(merged_df, logFC_df, by = "gene", all = TRUE)
    merged_df_se <- merge(merged_df_se, SE_df, by = "gene", all = TRUE)
    merged_df_pval <- merge(merged_df_pval, pval_df, by = "gene", all = TRUE)
    merged_df_adj_pval <- merge(merged_df_adj_pval, adj_pval_df, by = "gene", all = TRUE)
  }
}

# Put gene back as rownames if you prefer
rownames(merged_df) <- merged_df$gene; merged_df$gene <- NULL
rownames(merged_df_se) <- merged_df_se$gene; merged_df_se$gene <- NULL
rownames(merged_df_pval) <- merged_df_pval$gene; merged_df_pval$gene <- NULL
rownames(merged_df_adj_pval) <- merged_df_adj_pval$gene; merged_df_adj_pval$gene <- NULL
write.csv(merged_df, paste0(path_de_parent,"parent_limma_all_celltypes_indv_random_logFC.csv"), row.names = TRUE)
write.csv(merged_df_se, paste0(path_de_parent,"parent_limma_all_celltypes_indv_random_SE.csv"), row.names = TRUE)
write.csv(merged_df_pval, paste0(path_de_parent,"parent_limma_all_celltypes_indv_random_pval.csv"), row.names = TRUE)
write.csv(merged_df_adj_pval, paste0(path_de_parent,"parent_limma_all_celltypes_indv_random_adj_pval.csv"), row.names = TRUE)
```


## Differential Dispersion

We run differential dispersion (DD) analysis. We estimate dispersion from  residual variance estimates using Memento for each species, individual, and replicate. We run these estimates in LIMMA, modeling species as a fixed effect and individual as a random effect.

### Load data

We first load dispersion estimates and acquire the metadata. We add sex to the metadata.

```{r}
# load dispersion estimates
path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/cell_annotations/'
memento_var <- read.csv(paste0(path,"parent_crossfilt_memento_var_estimate_cot.csv"))

# Get column names
rownames(memento_var) = memento_var$gene

memento_var <- memento_var[-c(1, 2)]
cols <- colnames(memento_var)

# Split by "."
split_cols <- strsplit(cols, "\\.")

# Extract metadata from each name
meta_df <- do.call(rbind, lapply(split_cols, function(x) {
  data.frame(
    species = x[2],
    individual = x[3],
    replicate = x[4],
    celltype = paste(x[5:length(x)], collapse = " ")  # handles multiword like "Radial glia"
  )
}))

# Make sure rownames align with the columns
rownames(meta_df) <- cols

# Combine into a list
var_list <- list(
  var = memento_var,
  meta = meta_df
)


var_list <- add_sex(var_list)
head(var_list$meta)
```

### Run LIMMA
Next, we run LIMMA using the dreamlet package. We quantile normalize dispersion estimates before running LIMMA.
```{r, eval=FALSE}
path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/'
celltypes <- c('ENS glia','Glioblast',  'Neuroblast', 'Neuronal IPC', 'Radial glia')

for(i in 1:length(celltypes)){
# Get logical index of samples that match your condition
  keep <- var_list$meta$celltype == celltypes[i]
  
  # Quantile normalization
  bulk_subset_qn <- list(
    var = normalizeQuantiles(var_list$var[, keep]),
    meta   = var_list$meta[keep, ]
  )
   ### Run DREAM 
  sex=FALSE
  voomspan=0.5
  ddf = "Satterthwaite"  
  
  # species as fixed effect, individual random effect
  form <- ~ 0 + species + (1|individual)
  if (sex) form <- ~ 0 + species + (1|individual) + (1|sex)
  
  L_qn <- makeContrastsDream(
  formula = form,
  data = bulk_subset_qn$meta,
  contrasts = c("specieshuman-specieschimp")
)
  
  # estimate weights using linear mixed model of dream qn
  dream_fit_qn <- dream(bulk_subset_qn$var, form, bulk_subset_qn$meta, L_qn)
  dream_fit_qn <- eBayes(dream_fit_qn)
  saveRDS(dream_fit_qn,paste0(path_out, "parent_memento_dream_dv_qn_norm_fit_human_minus_chimp_random_indv_",celltypes[i],".rds"))
}
```

### Run MASH

We run mash to increase power for detecting differential dispersion.

#### Mash for diploid

Get the betas, SEs, and p-values from the output of LIMMA. Create matrix with these values for each gene and cell type.
```{r, eval=FALSE}
path_ml_parent <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/'
path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/mash_results/'
# Example
celltypes <- c('ENS glia', 'Glioblast',  'Neuroblast','Neuronal IPC', 'Radial glia')

# Initialize overlap genes and merged dataframe
overlap_genes <- NULL
merged_df <- NULL
merged_df_se <- NULL
merged_df_pval <- NULL
for (ct in celltypes) {
  # Get the dataframe dynamically (if they're separate objects)
  memento_limma_fit <- readRDS(paste0(path_ml_parent, "parent_memento_dream_dv_qn_norm_fit_human_minus_chimp_random_indv_",ct,".rds"))
  memento_limma_res_parent <- topTable(memento_limma_fit, coef = "specieshuman - specieschimpokayy", number = Inf)
  memento_limma_res_parent$SE <- memento_limma_res_parent$logFC/memento_limma_res_parent$t
  
  # Make sure rownames exist
  if (is.null(rownames(memento_limma_res_parent))) stop(paste("No rownames in", ct))
  
  # Extract the logFC column and rename it
  logFC_df <- data.frame(logFC = memento_limma_res_parent$logFC)
  colnames(logFC_df) <- paste0("dd.logFC.diploid.", ct)
  rownames(logFC_df) <- rownames(memento_limma_res_parent)
  
  # Same for SE
  SE_df <- data.frame(logFC = memento_limma_res_parent$SE)
  colnames(SE_df) <- paste0("dd.SE.diploid.", ct)
  rownames(SE_df) <- rownames(memento_limma_res_parent)
  
  # Same for pval
  pval_df <- data.frame(logFC = memento_limma_res_parent$P.Value)
  colnames(pval_df) <- paste0("dd.pvalue.diploid.", ct)
  rownames(pval_df) <- rownames(memento_limma_res_parent)

  # Merge into growing combined dataframe
  if (is.null(merged_df)) {
    merged_df <- logFC_df
    merged_df_se <- SE_df
    merged_df_pval <- pval_df
    overlap_genes <- rownames(logFC_df)
  } else {
    overlap_genes <- intersect(overlap_genes, rownames(logFC_df))
    
    merged_df <- merge(merged_df, logFC_df, by = "row.names", all = FALSE)
    rownames(merged_df) <- merged_df$Row.names
    merged_df$Row.names <- NULL
    
    merged_df_se <- merge(merged_df_se, SE_df, by = "row.names", all = FALSE)
    rownames(merged_df_se) <- merged_df_se$Row.names
    merged_df_se$Row.names <- NULL
    
    merged_df_pval <- merge(merged_df_pval, pval_df, by = "row.names", all = FALSE)
    rownames(merged_df_pval) <- merged_df_pval$Row.names
    merged_df_pval$Row.names <- NULL
  }
  

 
}

# Keep only overlapping genes
merged_df <- merged_df[overlap_genes, ]
merged_df_se <- merged_df_se[overlap_genes, ]
merged_df_pval <- merged_df_pval[overlap_genes, ]
write.csv(merged_df, paste0(path_out,"parent_memento_dream_random_indv_all_celltypes_logFC.csv"), row.names = TRUE)
write.csv(merged_df_se, paste0(path_out,"parent_memento_dream_random_indv_all_celltypes_SE.csv"), row.names = TRUE)
write.csv(merged_df_pval, paste0(path_out,"parent_memento_dream_random_indv_all_celltypes_pval.csv"), row.names = TRUE)
```


We use the p-value distribution from LIMMA and use this to run mash.
```{r,eval=FALSE}
data_path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/mash_results/'

## Load the beta, SE, and pval matrices
betahat_matrix <-  read.csv(paste0(data_path, "parent_memento_dream_random_indv_all_celltypes_logFC.csv"))
sehat_matrix <- read.csv(paste0(data_path, "parent_memento_dream_random_indv_all_celltypes_SE.csv"))
pval_matrix <- read.csv(paste0(data_path, "parent_memento_dream_random_indv_all_celltypes_pval.csv"))
rownames(betahat_matrix) <- betahat_matrix$X
rownames(sehat_matrix) <- sehat_matrix$X
rownames(pval_matrix) <- pval_matrix$X
betahat_matrix <- betahat_matrix[, -1, drop = FALSE]
sehat_matrix <- sehat_matrix[, -1, drop = FALSE]
pval_matrix <- pval_matrix[, -1, drop = FALSE]
betahat_matrix <- as.matrix(betahat_matrix)
sehat_matrix   <- as.matrix(sehat_matrix)
pval_matrix   <- as.matrix(pval_matrix)

## Run mash
data_mash <- mash_set_data(betahat_matrix, pval=pval_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed, file = paste0(data_path ,"parent_memento_dream_random_indv_mash_pval_results.rds"))

```



#### Mash for allotetraploid
Get the betas, SEs, and p-values from the output of LIMMA. Create matrix with these values for each gene and cell type.
```{r,eval=FALSE}
path_memento_hybrid <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/memento_results/'
celltypes <- c( 'ENS glia', 'Glioblast',  'Neuroblast','Neuronal IPC', 'Radial glia')
celltypes2 <- c( 'ENS_glia', 'Glioblast',  'Neuroblast','Neuronal_IPC', 'Radial_glia')

path_out <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/mash_results/'

# Initialize overlap genes and merged dataframe
overlap_genes <- NULL
merged_df <- NULL
merged_df_se <- NULL
merged_df_pval <- NULL
for (i in 1:length(celltypes)) {
  # Get the dataframe dynamically (if they're separate objects)
  memento_res_hybrid <- read.csv(paste0(path_memento_hybrid, "hybrid_crossfilt_memento_results_",celltypes2[i],".csv"))
  
  # Extract the logFC column and rename it
  logFC_df <- data.frame(logFC = memento_res_hybrid$dv_coef)
  colnames(logFC_df) <- paste0("dd.logFC.allotetraploid.", celltypes[i])
  rownames(logFC_df) <- memento_res_hybrid$gene
  
  # Same for SE
  SE_df <- data.frame(logFC = memento_res_hybrid$dv_se)
  colnames(SE_df) <- paste0("dd.SE.allotetraploid.", celltypes[i])
  rownames(SE_df) <- memento_res_hybrid$gene
  
  # Same for pval
  pval_df <- data.frame(logFC = memento_res_hybrid$dv_pval)
  colnames(pval_df) <- paste0("dd.pvalue.allotetraploid.", celltypes[i])
  rownames(pval_df) <- memento_res_hybrid$gene

  # Merge into growing combined dataframe
  if (is.null(merged_df)) {
    merged_df <- logFC_df
    merged_df_se <- SE_df
    merged_df_pval <- pval_df
    overlap_genes <- rownames(logFC_df)
  } else {
    overlap_genes <- intersect(overlap_genes, rownames(logFC_df))
    
    merged_df <- merge(merged_df, logFC_df, by = "row.names", all = FALSE)
    rownames(merged_df) <- merged_df$Row.names
    merged_df$Row.names <- NULL
    
    merged_df_se <- merge(merged_df_se, SE_df, by = "row.names", all = FALSE)
    rownames(merged_df_se) <- merged_df_se$Row.names
    merged_df_se$Row.names <- NULL
    
    merged_df_pval <- merge(merged_df_pval, pval_df, by = "row.names", all = FALSE)
    rownames(merged_df_pval) <- merged_df_pval$Row.names
    merged_df_pval$Row.names <- NULL
  }
  

 
}


# Keep only overlapping genes
merged_df <- merged_df[overlap_genes, ]
merged_df_se <- merged_df_se[overlap_genes, ]
merged_df_pval <- merged_df_pval[overlap_genes, ]
write.csv(merged_df, paste0(path_out,"hybrid_memento_all_celltypes_logFC.csv"), row.names = TRUE)
write.csv(merged_df_se, paste0(path_out,"hybrid_memento_all_celltypes_SE.csv"), row.names = TRUE)
write.csv(merged_df_pval, paste0(path_out,"hybrid_memento_all_celltypes_pval.csv"), row.names = TRUE)
```

We use the p-value distribution from LIMMA and use this to run mash.
```{r,eval=FALSE}
data_path <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/mash_results/'

## Load the beta and SE matrices
betahat_matrix <-  read.csv(paste0(data_path, "hybrid_memento_all_celltypes_logFC.csv"))
sehat_matrix <- read.csv(paste0(data_path, "hybrid_memento_all_celltypes_SE.csv"))
pval_matrix <- read.csv(paste0(data_path, "hybrid_memento_all_celltypes_pval.csv"))
rownames(betahat_matrix) <- betahat_matrix$X
rownames(sehat_matrix) <- sehat_matrix$X
rownames(pval_matrix) <- pval_matrix$X
betahat_matrix <- betahat_matrix[, -1, drop = FALSE]
sehat_matrix <- sehat_matrix[, -1, drop = FALSE]
pval_matrix <- pval_matrix[, -1, drop = FALSE]
betahat_matrix <- as.matrix(betahat_matrix)
sehat_matrix   <- as.matrix(sehat_matrix)
pval_matrix   <- as.matrix(pval_matrix)

## Run mash
data_mash <- mash_set_data(betahat_matrix, pval=pval_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed, file = paste0(data_path ,"hybrid_memento_mash_pval_results.rds"))

```


## Create data table for DD analysis

```{r}
# Paths for each analysis
path_de_hybrid <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/'
path_de_parent <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DE_results/limma_results/'
path_dd_mash <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/mash_results/'
path_ml_parent <- '/project/gilad/awchen55/differentialDispersion/data/crossfilt_clustered_data/DV_results/'
```

### Add DE results
```{r}
celltypes <- c('ENS glia', 'Glioblast',  'Neuroblast','Neuronal IPC', 'Radial glia')
celltypes2 <- c('ENS_glia', 'Glioblast',  'Neuroblast','Neuronal_IPC', 'Radial_glia')
celltypes3 <- c('ENS.glia', 'Glioblast',  'Neuroblast','Neuronal.IPC', 'Radial.glia')

master_df <- NULL


###### DE Results from Allotetraploid
for (i in seq_along(celltypes2)) {
  # Read data
  de_hybrid_results <- read.csv(paste0(path_de_hybrid, "hybrid_wilcoxon_results_human_minus_chimp_", celltypes2[i], ".csv"))
  
  # adjusted p-value using Benjamini-Hochberg
  de_hybrid_results <- de_hybrid_results %>% 
    mutate(adj.p.val = ifelse(obs.paired < 5, NA, p.adjust(pvalue, method = "BH")))
  
  # Rename columns
  colnames(de_hybrid_results)[colnames(de_hybrid_results) == "X"] <- "gene"
  colnames(de_hybrid_results)[colnames(de_hybrid_results) == "pvalue"] <- paste0("de.pvalue.allotetraploid.", celltypes3[i])
  colnames(de_hybrid_results)[colnames(de_hybrid_results) == "logFC"]  <- paste0("de.logFC.allotetraploid.", celltypes3[i])
  colnames(de_hybrid_results)[colnames(de_hybrid_results) == "adj.p.val"]  <- paste0("de.adj.pvalue.allotetraploid.", celltypes3[i])
  
  # Keep only needed columns
  de_hybrid_results <- de_hybrid_results[, c("gene", 
                                             paste0("de.logFC.allotetraploid.", celltypes3[i]),
                                             paste0("de.pvalue.allotetraploid.", celltypes3[i]),
                                             paste0("de.adj.pvalue.allotetraploid.", celltypes3[i])
                                             )]
  
  # Merge into master table (full outer join)
  if (is.null(master_df)) {
    master_df <- de_hybrid_results
  } else {
    master_df <- full_join(master_df, de_hybrid_results, by = "gene")
  }
}

###### DE Results from Diploid

## Load the beta, SE, pval, adj pval matrices
betahat_matrix <-  read.csv(paste0(path_de_parent, "parent_limma_all_celltypes_indv_random_logFC.csv"))
sehat_matrix <- read.csv(paste0(path_de_parent, "parent_limma_all_celltypes_indv_random_SE.csv"))
pval_matrix <- read.csv(paste0(path_de_parent, "parent_limma_all_celltypes_indv_random_pval.csv"))
adj_pval_matrix <- read.csv(paste0(path_de_parent, "parent_limma_all_celltypes_indv_random_adj_pval.csv"))
colnames(betahat_matrix)[colnames(betahat_matrix) == "X"] <- "gene"
colnames(sehat_matrix)[colnames(sehat_matrix) == "X"] <- "gene"
colnames(pval_matrix)[colnames(pval_matrix) == "X"] <- "gene"
colnames(adj_pval_matrix)[colnames(adj_pval_matrix) == "X"] <- "gene"

## Full join DE results
master_df <- dplyr::full_join(master_df, betahat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, sehat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, pval_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, adj_pval_matrix, by = "gene")
head(master_df)
```

### Add DD results


```{r}

###### DD Results from Allotetraploid

## Load the beta and SE matrices
betahat_matrix <-  read.csv(paste0(path_dd_mash, "hybrid_memento_human_minus_chimp_all_celltypes_logFC.csv"))[,-1]
sehat_matrix <- read.csv(paste0(path_dd_mash, "hybrid_memento_all_celltypes_SE.csv"))
colnames(betahat_matrix)[colnames(betahat_matrix) == "X"] <- "gene"
colnames(sehat_matrix)[colnames(sehat_matrix) == "X"] <- "gene"

## Load mash results to get lfsr
mash_results <- readRDS(paste0(path_dd_mash, "hybrid_memento_mash_pval_results.rds"))

# get lfsr
lfsrs <- get_lfsr(mash_results) %>%
  as.data.frame() %>%
  rownames_to_column("gene")

# change column names
colnames(lfsrs) <- gsub("dd\\.logFC\\.allotetraploid\\.", 
                     "dd.lfsr.allotetraploid.", 
                     colnames(lfsrs))

## Full join DD results from allotetraploid
master_df <- dplyr::full_join(master_df, betahat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, sehat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, lfsrs, by = "gene")

###### DD Results from Diploid

## Load the beta and SE matrices
betahat_matrix <-  read.csv(paste0(path_dd_mash, "parent_memento_dream_random_indv_all_celltypes_logFC.csv"))
sehat_matrix <- read.csv(paste0(path_dd_mash, "parent_memento_dream_random_indv_all_celltypes_SE.csv"))
colnames(betahat_matrix)[colnames(betahat_matrix) == "X"] <- "gene"
colnames(sehat_matrix)[colnames(sehat_matrix) == "X"] <- "gene"

## Load mash results to get lfsr
mash_results <- readRDS(paste0(path_dd_mash, "parent_memento_dream_random_indv_mash_pval_results.rds"))

# get lfsr
lfsrs <- get_lfsr(mash_results) %>%
  as.data.frame() %>%
  rownames_to_column("gene")

# change column names
colnames(lfsrs) <- gsub("dd\\.logFC\\.diploid\\.", 
                     "dd.lfsr.diploid.", 
                     colnames(lfsrs))

## Full join DD results from diploid
master_df <- dplyr::full_join(master_df, betahat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, sehat_matrix, by = "gene")
master_df <- dplyr::full_join(master_df, lfsrs, by = "gene")
```

### Add additional metrics

We filter for only testable genes (genes that can be tested in both allotetraploid and diploid settings).
```{r}
# Identify dd.logFC columns
allo_cols <- grep("^dd\\.logFC\\.allotetraploid\\.", names(master_df), value = TRUE)
dip_cols  <- grep("^dd\\.logFC\\.diploid\\.",        names(master_df), value = TRUE)

# Extract the celltype suffix after the last prefix
allo_ct <- sub("^dd\\.logFC\\.allotetraploid\\.", "", allo_cols)
dip_ct  <- sub("^dd\\.logFC\\.diploid\\.",        "", dip_cols)

# Only keep celltypes present in BOTH diploid and allotetraploid
common_ct <- intersect(allo_ct, dip_ct)

# Build matched column vectors in the same order
allo_cols_m <- paste0("dd.logFC.allotetraploid.", common_ct)
dip_cols_m  <- paste0("dd.logFC.diploid.",        common_ct)

# A gene is "testable" in a celltype if BOTH entries are non-NA
# Keep gene if testable in AT LEAST ONE celltype
keep_gene <- rowSums(!is.na(master_df[, allo_cols_m, drop = FALSE]) &
                     !is.na(master_df[, dip_cols_m,  drop = FALSE])) > 0

# Filter master_df
master_df_filtered <- master_df[keep_gene, , drop = FALSE]
```

#### Idenitfy Recapitulated DD genes

We add a flag for recapitulated DD genes. These are genes that are significant (LFSR < 0.05) in both settings and have the same sign of effect size (logFC). 1 if genes is recapitulated, 0 if not.
```{r}
celltypes3 <- c('ENS.glia', 'Glioblast',  'Neuroblast','Neuronal.IPC', 'Radial.glia')
for (ct in celltypes3) {

  lfsr_dip <- paste0("dd.lfsr.diploid.", ct)
  lfsr_allo <- paste0("dd.lfsr.allotetraploid.", ct)
  logfc_dip <- paste0("dd.logFC.diploid.", ct)
  logfc_allo <- paste0("dd.logFC.allotetraploid.", ct)
  new_col <- paste0("dd.recapitulated.same.sign.", ct)

  master_df_filtered[[new_col]] <- as.integer(
    master_df_filtered[[lfsr_dip]] < 0.05 &
    master_df_filtered[[lfsr_allo]] < 0.05 &
    sign(master_df_filtered[[logfc_dip]]) ==
    sign(master_df_filtered[[logfc_allo]])
  )
}

# add column that shows the number of cell types gene is DD recapitulated
master_df_filtered <- master_df_filtered %>%
  dplyr::mutate(
    dd.recapitulated.same.sign.n.celltypes =
      rowSums(dplyr::select(., starts_with("dd.recapitulated.same.sign.")), na.rm = TRUE)
  )
```


#### Calculate cis proportion

We calculate the cis proportion by cell type.
```{r}

# Calculate cis proportion
for (ct in celltypes3) {
  
  allotetraploid_col <- paste0("dd.logFC.allotetraploid.", ct)
  diploid_col <- paste0("dd.logFC.diploid.", ct)
  prop_col   <- paste0("dd.cis.proportion.", ct)
  
  # Only compute if both columns exist
  if (all(c(allotetraploid_col, diploid_col) %in% colnames(master_df_filtered))) {
    master_df_filtered[[prop_col]] <- abs(master_df_filtered[[allotetraploid_col]]) /
      (abs(master_df_filtered[[allotetraploid_col]]) + abs(master_df_filtered[[diploid_col]] - master_df_filtered[[allotetraploid_col]]))
  } else {
    warning(paste("Missing columns for", ct))
  }
}
```


We calculate the mean cis proportion for ONLY cell types that are DD and recapitulated.
```{r}
master_df_filtered <- master_df_filtered %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    dd.recapitulated.mean.cis.proportion = mean(
      ifelse(
        c_across(paste0("dd.recapitulated.same.sign.", celltypes3)) == 1,
        c_across(paste0("dd.cis.proportion.", celltypes3)),
        NA
      ),
      na.rm = TRUE
    )
  ) %>%
  dplyr::ungroup()
```



Categorize DD gene as cis or trans per cell type. If cis proportion > 0.70 then cis, < 0.30 then trans, else neither.
```{r}
cis_cutoff   <- 0.70
trans_cutoff <- 0.30

for (ct in celltypes3) {

  recap_col <- paste0("dd.recapitulated.same.sign.", ct)
  cisprop_col <- paste0("dd.cis.proportion.", ct)
  out_col <- paste0("dd.cis.category.", ct)

  recap <- master_df_filtered[[recap_col]]
  cisprop <- master_df_filtered[[cisprop_col]]

  # default: not DD
  cat_vec <- rep("not DD", nrow(master_df_filtered))

  # only classify among recapitulated DD (same sign)
  is_dd <- !is.na(recap) & recap == 1 & !is.na(cisprop)

  cat_vec[is_dd & cisprop > cis_cutoff] <- "cis"
  cat_vec[is_dd & cisprop < trans_cutoff] <- "trans"
  cat_vec[is_dd & cisprop <= cis_cutoff & cisprop >= trans_cutoff] <- "neither"

  # store as factor with consistent ordering (optional)
  master_df_filtered[[out_col]] <- factor(
    cat_vec,
    levels = c("cis", "neither", "trans", "not DD")
  )
}
```

Create overall category for cis/trans DD genes. 
- If cis in at least one cell type and neither or not DD in all other cell types, then cis
- If trans in at least one cell type and neither or not DD in all other cell types, then trans
- If neither in at least one cell type and not DD in all other cell types, then neither
- If cis or trans in different cell types, then mixed
- Else, not DD
```{r}
cat_cols <- names(master_df_filtered)[startsWith(names(master_df_filtered), "dd.cis.category.")]

master_df_filtered <- master_df_filtered %>%
  rowwise() %>%
  mutate(
    dd.cis.category.overall = {
      vals <- c_across(all_of(cat_cols))
      vals <- vals[!is.na(vals)]

      has_cis     <- "cis" %in% vals
      has_trans   <- "trans" %in% vals
      has_neither <- "neither" %in% vals

      if (!has_cis & !has_trans & !has_neither) {
        "not DD"
      } else if (has_cis & has_trans) {
        "mixed"
      } else if (has_cis) {
        "cis"
      } else if (has_trans) {
        "trans"
      } else if (has_neither) {
        "neither"
      } else {
        "not DD"
      }
    }
  ) %>%
  ungroup()
```

Create an overall category for cis or trans DD genes based on mean cis proportion of cell types that are DD.
```{r}
master_df_filtered <- master_df_filtered %>%
  dplyr::mutate(
    dd.cis.category.overall.mean.definition = dplyr::case_when(
      is.nan(dd.recapitulated.mean.cis.proportion) ~ "not DD",
      dd.recapitulated.mean.cis.proportion > 0.7  ~ "cis",
      dd.recapitulated.mean.cis.proportion < 0.3  ~ "trans",
      TRUE                                        ~ "neither"
    )
  )
```


```{r}
master_df_filtered
```


## Save data table
```{r}
path_final <- '/project/gilad/awchen55/differentialDispersion/results/variability_paper/data/'

write.csv(master_df_filtered,paste0(path_final,"DD_interspecies_results.csv"))
writexl::write_xlsx(
  master_df_filtered,
  path = paste0(path_final, "DD_interspecies_results.xlsx")
)
```

